---
title: "01 Data Audit and QC"
subtitle: "Zebrafish Sequential Behavioral Battery — Data Usability Audit"
author: "Caique Ortiz Colombo Arnoldi Fabris"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

# Question

Are the wrangled battery-level CSVs usable for downstream endpoint registry, model-readiness, battery-comparison, and PCA notebooks?

# Why this notebook exists

This notebook audits the battery-level files created by `00_data_wrangling.ipynb`.

It checks file availability, battery identity, fish identifiers, treatment-by-exposure cell sizes, datetime metadata, missingness, range violations, duration flags, derived-variable consistency, condition factor consistency, QC flags, and exclusion logs.

This notebook does **not** run treatment-effect models, battery-effect tests, PCA, equivalence testing, or manuscript interpretation. Its output is an audited set of BT1, BT2, and BT3 datasets plus QC logs for Notebook 02.

# What assumptions matter

The current repository stores wrangled battery-level files in `data_processed/`:

- `battery_suite_1.csv`
- `battery_suite_2.csv`
- `battery_suite_3.csv`

Notebook 01 audits those files and writes:

- `bt1_audited.csv`
- `bt2_audited.csv`
- `bt3_audited.csv`

The endpoint registry from `00_project_setup.Rmd` is used when available. If the registry file is missing, this notebook uses a local fallback registry only for QC targeting.

# What decision this supports

This notebook decides whether the data can proceed to `02_endpoint_hierarchy_and_assumptions.Rmd`, which now functions as the endpoint registry and model-readiness notebook.

---

# 01.1 Project paths and helper functions

```{r project-root}
find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = FALSE)
  candidates <- current

  while (TRUE) {
    parent <- dirname(tail(candidates, 1))
    if (identical(parent, tail(candidates, 1))) break
    candidates <- c(candidates, parent)
  }

  for (candidate in unique(candidates)) {
    has_rproj <- length(list.files(candidate, pattern = "\\.Rproj$", full.names = TRUE)) > 0
    has_git <- dir.exists(file.path(candidate, ".git"))
    has_data_dirs <- dir.exists(file.path(candidate, "data_raw")) &&
      dir.exists(file.path(candidate, "data_processed"))

    if (has_rproj || has_git || has_data_dirs) {
      return(candidate)
    }
  }

  current
}

PROJECT_ROOT <- find_project_root()

paths <- list(
  project_root = PROJECT_ROOT,
  data_processed = file.path(PROJECT_ROOT, "data_processed"),
  outputs = file.path(PROJECT_ROOT, "outputs"),
  outputs_logs = file.path(PROJECT_ROOT, "outputs", "logs"),
  outputs_tables = file.path(PROJECT_ROOT, "outputs", "tables")
)

for (path in paths) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE, showWarnings = FALSE)
}

paths
```

```{r helper-functions}
read_csv_required <- function(path) {
  if (!file.exists(path)) {
    stop("Required file not found: ", path)
  }
  read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
}

read_csv_optional <- function(path) {
  if (!file.exists(path)) return(NULL)
  read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
}

write_table <- function(df, filename, folder = paths$outputs_logs) {
  path <- file.path(folder, filename)
  write.csv(df, path, row.names = FALSE)
  path
}

has_cols <- function(df, cols) {
  cols[cols %in% names(df)]
}

first_existing_col <- function(df, candidates) {
  hits <- candidates[candidates %in% names(df)]
  if (length(hits) == 0) return(NA_character_)
  hits[1]
}

as_bool <- function(x) {
  if (is.logical(x)) return(x)
  if (is.numeric(x)) return(x == 1)
  tolower(as.character(x)) %in% c("true", "1", "yes", "y")
}

count_true <- function(x) {
  sum(as_bool(x), na.rm = TRUE)
}

is_missing_value <- function(x) {
  is.na(x) | trimws(as.character(x)) == ""
}

as_num <- function(x) {
  suppressWarnings(as.numeric(x))
}

safe_min <- function(x) {
  x <- as_num(x)
  if (all(is.na(x))) return(NA_real_)
  min(x, na.rm = TRUE)
}

safe_max <- function(x) {
  x <- as_num(x)
  if (all(is.na(x))) return(NA_real_)
  max(x, na.rm = TRUE)
}

safe_mean <- function(x) {
  x <- as_num(x)
  if (all(is.na(x))) return(NA_real_)
  mean(x, na.rm = TRUE)
}

safe_median <- function(x) {
  x <- as_num(x)
  if (all(is.na(x))) return(NA_real_)
  median(x, na.rm = TRUE)
}

safe_sd <- function(x) {
  x <- as_num(x)
  if (sum(!is.na(x)) < 2) return(NA_real_)
  sd(x, na.rm = TRUE)
}

safe_iqr <- function(x) {
  x <- as_num(x)
  if (sum(!is.na(x)) < 2) return(NA_real_)
  IQR(x, na.rm = TRUE)
}

empty_df <- function(...) {
  data.frame(..., stringsAsFactors = FALSE)
}
```

---

# 01.2 Load wrangled battery-level files

```{r load-files}
bt1_path <- file.path(paths$data_processed, "battery_suite_1.csv")
bt2_path <- file.path(paths$data_processed, "battery_suite_2.csv")
bt3_path <- file.path(paths$data_processed, "battery_suite_3.csv")

bt1 <- read_csv_required(bt1_path)
bt2 <- read_csv_required(bt2_path)
bt3 <- read_csv_required(bt3_path)

battery_data <- list(BT1 = bt1, BT2 = bt2, BT3 = bt3)

wrangling_exclusion_log <- read_csv_optional(file.path(paths$outputs_logs, "wrangling_exclusion_log.csv"))
if (is.null(wrangling_exclusion_log)) {
  wrangling_exclusion_log <- read_csv_optional(file.path(paths$outputs_logs, "exclusion_log.csv"))
}

wrangling_qc_log <- read_csv_optional(file.path(paths$outputs_logs, "wrangling_qc_log.csv"))
endpoint_registry <- read_csv_optional(file.path(paths$outputs_logs, "00_endpoint_registry.csv"))

loaded_files <- data.frame(
  dataset = c("BT1", "BT2", "BT3"),
  path = c(bt1_path, bt2_path, bt3_path),
  exists = c(file.exists(bt1_path), file.exists(bt2_path), file.exists(bt3_path)),
  rows = c(nrow(bt1), nrow(bt2), nrow(bt3)),
  columns = c(ncol(bt1), ncol(bt2), ncol(bt3)),
  stringsAsFactors = FALSE
)

loaded_files
```

```{r export-loaded-files}
write_table(loaded_files, "01_qc_loaded_files.csv")
```

---

# 01.3 Battery identity

Expected battery meaning:

- BT1: full sequence.
- BT2: LDT/NTT alternative or isolated context, depending on endpoint.
- BT3: isolated endurance context.

This section checks whether the files and available presence flags are compatible with those meanings.

```{r battery-identity}
presence_cols <- c("ldt_present_flag", "ntt_present_flag", "endurance_present_flag")

battery_identity <- do.call(
  rbind,
  lapply(names(battery_data), function(b) {
    df <- battery_data[[b]]

    data.frame(
      battery_file = b,
      n_rows = nrow(df),
      battery_values = if ("battery" %in% names(df)) paste(sort(unique(df$battery)), collapse = "; ") else NA_character_,
      has_ldt_presence_col = "ldt_present_flag" %in% names(df),
      has_ntt_presence_col = "ntt_present_flag" %in% names(df),
      has_endurance_presence_col = "endurance_present_flag" %in% names(df),
      ldt_present_count = if ("ldt_present_flag" %in% names(df)) count_true(df$ldt_present_flag) else NA_integer_,
      ntt_present_count = if ("ntt_present_flag" %in% names(df)) count_true(df$ntt_present_flag) else NA_integer_,
      endurance_present_count = if ("endurance_present_flag" %in% names(df)) count_true(df$endurance_present_flag) else NA_integer_,
      expected_context = switch(
        b,
        BT1 = "full sequential battery",
        BT2 = "alternative or isolated LDT/NTT context",
        BT3 = "isolated endurance context",
        "not specified"
      ),
      stringsAsFactors = FALSE
    )
  })
)

battery_identity
```

```{r export-battery-identity}
write_table(battery_identity, "01_qc_battery_identity.csv")
```

---

# 01.4 Dimensions, schema, and variable availability

```{r dimensions-schema}
dataset_structure <- do.call(
  rbind,
  lapply(names(battery_data), function(b) {
    df <- battery_data[[b]]

    data.frame(
      battery = b,
      n_rows = nrow(df),
      n_columns = ncol(df),
      n_unique_fish_id = if ("fish_id" %in% names(df)) length(unique(df$fish_id)) else NA_integer_,
      has_battery_column = "battery" %in% names(df),
      has_treatment_column = "treatment" %in% names(df),
      has_exposure_column = "exposure" %in% names(df),
      has_datetime_column = "datetime" %in% names(df),
      stringsAsFactors = FALSE
    )
  })
)

all_columns <- sort(unique(unlist(lapply(battery_data, names))))

schema_consistency <- data.frame(
  column = all_columns,
  in_BT1 = all_columns %in% names(bt1),
  in_BT2 = all_columns %in% names(bt2),
  in_BT3 = all_columns %in% names(bt3),
  stringsAsFactors = FALSE
)

schema_consistency$present_in_all <- schema_consistency$in_BT1 & schema_consistency$in_BT2 & schema_consistency$in_BT3

dataset_structure
schema_consistency
```

```{r endpoint-registry-availability}
if (is.null(endpoint_registry)) {
  registry_variables <- c(
    "dist_total", "mov_total", "vel_mean", "stratum_pref",
    "time_bright", "first_choice_binary", "latency", "mtpc", "num_changes",
    "resistance_index", "attempts", "last_flux", "time_in_last_flux",
    "blood_sugar", "condition_factor_K", "Kc", "Kf", "wt", "ls", "sex"
  )

  endpoint_registry_local <- data.frame(
    variable = registry_variables,
    pca_included = ifelse(
      registry_variables %in% c(
        "dist_total", "mov_total", "vel_mean", "stratum_pref",
        "time_bright", "first_choice_binary", "latency", "mtpc", "num_changes",
        "resistance_index", "attempts", "blood_sugar", "condition_factor_K", "Kc", "Kf"
      ),
      "registry_or_pca_candidate",
      "qc_or_support"
    ),
    stringsAsFactors = FALSE
  )
} else {
  endpoint_registry_local <- endpoint_registry
}

registry_variables <- unique(endpoint_registry_local$variable)

condition_aliases <- c("condition_factor_K", "Kc", "Kf", "kf", "condition_factor", "fulton_k")
registry_variables_expanded <- unique(c(registry_variables, condition_aliases))

registry_availability <- do.call(
  rbind,
  lapply(registry_variables_expanded, function(v) {
    data.frame(
      variable = v,
      in_BT1 = v %in% names(bt1),
      in_BT2 = v %in% names(bt2),
      in_BT3 = v %in% names(bt3),
      stringsAsFactors = FALSE
    )
  })
)

registry_availability
```

```{r export-dimensions-schema}
write_table(dataset_structure, "01_qc_dataset_structure.csv")
write_table(schema_consistency, "01_qc_schema_consistency.csv")
write_table(registry_availability, "01_qc_registry_variable_availability.csv")
```

---

# 01.5 Fish ID integrity

This section checks duplicated IDs within each battery and ID overlap between batteries.

```{r fish-id-integrity}
fish_id_integrity <- do.call(
  rbind,
  lapply(names(battery_data), function(b) {
    df <- battery_data[[b]]

    if (!"fish_id" %in% names(df)) {
      return(data.frame(
        battery = b,
        n_rows = nrow(df),
        n_unique_fish_id = NA_integer_,
        duplicated_fish_id_count = NA_integer_,
        missing_fish_id_count = NA_integer_,
        stringsAsFactors = FALSE
      ))
    }

    fish_id <- as.character(df$fish_id)

    data.frame(
      battery = b,
      n_rows = nrow(df),
      n_unique_fish_id = length(unique(fish_id[!is_missing_value(fish_id)])),
      duplicated_fish_id_count = sum(duplicated(fish_id) & !is_missing_value(fish_id)),
      missing_fish_id_count = sum(is_missing_value(fish_id)),
      stringsAsFactors = FALSE
    )
  })
)

fish_ids <- lapply(battery_data, function(df) {
  if (!"fish_id" %in% names(df)) return(character(0))
  x <- as.character(df$fish_id)
  unique(x[!is_missing_value(x)])
})

cross_battery_overlap <- data.frame(
  comparison = c("BT1_vs_BT2", "BT1_vs_BT3", "BT2_vs_BT3"),
  overlap_n = c(
    length(intersect(fish_ids$BT1, fish_ids$BT2)),
    length(intersect(fish_ids$BT1, fish_ids$BT3)),
    length(intersect(fish_ids$BT2, fish_ids$BT3))
  ),
  stringsAsFactors = FALSE
)

fish_id_independence_flag <- all(cross_battery_overlap$overlap_n == 0)

fish_id_integrity
cross_battery_overlap
fish_id_independence_flag
```

```{r export-fish-id-integrity}
write_table(fish_id_integrity, "01_qc_fish_id_integrity.csv")
write_table(cross_battery_overlap, "01_qc_cross_battery_fish_id_overlap.csv")
```

---

# 01.6 Treatment by exposure cells

```{r cell-counts}
cell_counts_one <- function(df, battery_label) {
  required <- c("treatment", "exposure")
  if (!all(required %in% names(df))) {
    return(data.frame(
      battery = battery_label,
      treatment = NA_character_,
      exposure = NA_character_,
      n = NA_integer_,
      stringsAsFactors = FALSE
    ))
  }

  counts <- as.data.frame(table(df$treatment, df$exposure), stringsAsFactors = FALSE)
  names(counts) <- c("treatment", "exposure", "n")
  counts$battery <- battery_label
  counts[, c("battery", "treatment", "exposure", "n")]
}

cell_counts <- do.call(
  rbind,
  lapply(names(battery_data), function(b) cell_counts_one(battery_data[[b]], b))
)

cell_counts
```

```{r export-cell-counts}
write_table(cell_counts, "01_qc_cell_counts_treatment_by_exposure.csv")
```

---

# 01.7 Datetime audit

Datetime is audit metadata. It is not used here as an inferential covariate.

```{r datetime-audit}
datetime_audit <- do.call(
  rbind,
  lapply(names(battery_data), function(b) {
    df <- battery_data[[b]]

    if (!"datetime" %in% names(df)) {
      return(data.frame(
        battery = b,
        n_rows = nrow(df),
        datetime_present = FALSE,
        datetime_missing = NA_integer_,
        parse_failed_count = NA_integer_,
        min_datetime = NA_character_,
        max_datetime = NA_character_,
        stringsAsFactors = FALSE
      ))
    }

    dt_raw <- df$datetime
    dt_missing <- is_missing_value(dt_raw)
    parsed <- suppressWarnings(as.POSIXct(dt_raw, tz = "UTC"))

    parse_failed <- if ("datetime_parse_failed_flag" %in% names(df)) {
      as_bool(df$datetime_parse_failed_flag)
    } else {
      is.na(parsed) & !dt_missing
    }

    data.frame(
      battery = b,
      n_rows = nrow(df),
      datetime_present = TRUE,
      datetime_missing = sum(dt_missing),
      parse_failed_count = sum(parse_failed, na.rm = TRUE),
      min_datetime = if (all(is.na(parsed))) NA_character_ else as.character(min(parsed, na.rm = TRUE)),
      max_datetime = if (all(is.na(parsed))) NA_character_ else as.character(max(parsed, na.rm = TRUE)),
      stringsAsFactors = FALSE
    )
  })
)

datetime_audit
```

```{r datetime-by-cell}
summarize_dates_by_cell <- function(df, battery_label) {
  required <- c("treatment", "exposure", "datetime")

  if (!all(required %in% names(df))) {
    return(data.frame(
      battery = battery_label,
      treatment = NA_character_,
      exposure = NA_character_,
      n_rows = nrow(df),
      n_datetime_missing = NA_integer_,
      n_datetime_parsed = NA_integer_,
      unique_dates = NA_character_,
      datetime_note = "Required datetime fields not present.",
      stringsAsFactors = FALSE
    ))
  }

  parsed <- suppressWarnings(as.POSIXct(df$datetime, tz = "UTC"))
  tmp <- data.frame(
    battery = battery_label,
    treatment = as.character(df$treatment),
    exposure = as.character(df$exposure),
    date = as.Date(parsed),
    datetime_missing = is_missing_value(df$datetime),
    datetime_parsed = !is.na(parsed),
    stringsAsFactors = FALSE
  )

  groups <- split(tmp, list(tmp$battery, tmp$treatment, tmp$exposure), drop = TRUE)

  if (length(groups) == 0) {
    return(data.frame(
      battery = battery_label,
      treatment = NA_character_,
      exposure = NA_character_,
      n_rows = nrow(df),
      n_datetime_missing = sum(tmp$datetime_missing, na.rm = TRUE),
      n_datetime_parsed = sum(tmp$datetime_parsed, na.rm = TRUE),
      unique_dates = NA_character_,
      datetime_note = "No treatment/exposure groups available.",
      stringsAsFactors = FALSE
    ))
  }

  out <- do.call(
    rbind,
    lapply(groups, function(g) {
      unique_dates <- sort(unique(g$date[!is.na(g$date)]))

      data.frame(
        battery = unique(g$battery)[1],
        treatment = unique(g$treatment)[1],
        exposure = unique(g$exposure)[1],
        n_rows = nrow(g),
        n_datetime_missing = sum(g$datetime_missing, na.rm = TRUE),
        n_datetime_parsed = sum(g$datetime_parsed, na.rm = TRUE),
        unique_dates = if (length(unique_dates) == 0) NA_character_ else paste(unique_dates, collapse = "; "),
        datetime_note = if (length(unique_dates) == 0) "No parsed datetime/date available for this cell." else "Parsed date(s) available.",
        stringsAsFactors = FALSE
      )
    })
  )

  rownames(out) <- NULL
  out
}

datetime_by_cell <- do.call(
  rbind,
  lapply(names(battery_data), function(b) summarize_dates_by_cell(battery_data[[b]], b))
)

datetime_by_cell
```

```{r export-datetime-audit}
write_table(datetime_audit, "01_qc_datetime_audit.csv")
write_table(datetime_by_cell, "01_qc_datetime_by_treatment_exposure.csv")
```

---

# 01.8 Missingness audit

Missingness is documented. It is not automatically treated as exclusion.

```{r missingness-by-variable}
missingness_by_variable <- do.call(
  rbind,
  lapply(names(battery_data), function(b) {
    df <- battery_data[[b]]

    data.frame(
      battery = b,
      variable = names(df),
      n_rows = nrow(df),
      missing_n = vapply(df, function(x) sum(is_missing_value(x)), integer(1)),
      stringsAsFactors = FALSE
    )
  })
)

missingness_by_variable$missing_pct <- round(
  100 * missingness_by_variable$missing_n / missingness_by_variable$n_rows,
  2
)

missingness_by_variable
```

```{r missingness-by-cell}
key_missing_fields <- unique(c(
  registry_variables_expanded,
  "fish_id", "battery", "treatment", "exposure", "datetime",
  "source_file", "source_dataset", "source_row_index",
  "wt", "lt", "ls", "weight", "length", "standard_length",
  "conductivity", "pH", "ph", "temperature", "temp", "oxygen", "dissolved_oxygen",
  "mov_bottom", "mov_upper", "dist_bottom", "dist_upper", "vel_bottom", "vel_upper"
))

missingness_by_cell_parts <- lapply(names(battery_data), function(b) {
  df <- battery_data[[b]]
  available <- has_cols(df, key_missing_fields)

  if (!all(c("treatment", "exposure") %in% names(df)) || length(available) == 0) {
    return(data.frame())
  }

  do.call(
    rbind,
    lapply(available, function(v) {
      tmp <- data.frame(
        battery = b,
        treatment = as.character(df$treatment),
        exposure = as.character(df$exposure),
        variable = v,
        missing = is_missing_value(df[[v]]),
        stringsAsFactors = FALSE
      )

      groups <- split(tmp, list(tmp$battery, tmp$treatment, tmp$exposure, tmp$variable), drop = TRUE)

      do.call(
        rbind,
        lapply(groups, function(g) {
          data.frame(
            battery = unique(g$battery)[1],
            treatment = unique(g$treatment)[1],
            exposure = unique(g$exposure)[1],
            variable = unique(g$variable)[1],
            n_rows = nrow(g),
            missing_n = sum(g$missing, na.rm = TRUE),
            missing_pct = round(100 * sum(g$missing, na.rm = TRUE) / nrow(g), 2),
            stringsAsFactors = FALSE
          )
        })
      )
    })
  )
})

missingness_by_cell <- do.call(rbind, missingness_by_cell_parts)

missingness_by_cell
```

```{r export-missingness}
write_table(missingness_by_variable, "01_qc_missingness_by_variable.csv")
write_table(missingness_by_cell, "01_qc_missingness_by_cell.csv")
```

---

# 01.9 Range and impossible-value checks

These checks flag values outside plausible bounds. They do not delete data.

```{r range-summary}
numeric_candidates <- unique(c(
  registry_variables_expanded,
  "time_bright_prop", "irn_computed",
  "mov_bottom", "mov_upper", "dist_bottom", "dist_upper", "vel_bottom", "vel_upper",
  "wt", "lt", "ls", "weight", "length", "standard_length",
  "conductivity", "pH", "ph", "temperature", "temp", "oxygen", "dissolved_oxygen"
))

numeric_candidates <- setdiff(numeric_candidates, c("fish_id", "battery", "treatment", "exposure", "datetime", "sex"))

range_summary <- do.call(
  rbind,
  lapply(names(battery_data), function(b) {
    df <- battery_data[[b]]
    available <- has_cols(df, numeric_candidates)

    if (length(available) == 0) return(data.frame())

    do.call(
      rbind,
      lapply(available, function(v) {
        x <- as_num(df[[v]])
        data.frame(
          battery = b,
          variable = v,
          n_nonmissing = sum(!is.na(x)),
          min = safe_min(x),
          median = safe_median(x),
          mean = safe_mean(x),
          sd = safe_sd(x),
          iqr = safe_iqr(x),
          max = safe_max(x),
          negative_n = sum(x < 0, na.rm = TRUE),
          stringsAsFactors = FALSE
        )
      })
    )
  })
)

range_summary
```

```{r bounded-range-checks}
bounded_checks <- data.frame(
  battery = character(),
  variable = character(),
  rule = character(),
  violation_n = integer(),
  stringsAsFactors = FALSE
)

add_bounded_check <- function(out, df, battery_label, variable, rule_label, predicate) {
  if (!variable %in% names(df)) return(out)

  x <- as_num(df[[variable]])
  violation <- predicate(x)

  rbind(
    out,
    data.frame(
      battery = battery_label,
      variable = variable,
      rule = rule_label,
      violation_n = sum(violation, na.rm = TRUE),
      stringsAsFactors = FALSE
    )
  )
}

for (b in names(battery_data)) {
  df <- battery_data[[b]]

  bounded_checks <- add_bounded_check(
    bounded_checks, df, b, "time_bright",
    "time_bright should not be negative",
    function(x) x < 0
  )

  bounded_checks <- add_bounded_check(
    bounded_checks, df, b, "time_bright",
    "time_bright > 480 is flagged for LDT scoring-window audit",
    function(x) x > 480
  )

  bounded_checks <- add_bounded_check(
    bounded_checks, df, b, "time_bright",
    "time_bright > 600 is impossible under a 10-min LDT scoring window",
    function(x) x > 600
  )

  bounded_checks <- add_bounded_check(
    bounded_checks, df, b, "mov_total",
    "mov_total > 480 is flagged for NTT scoring-window audit",
    function(x) x > 480
  )

  bounded_checks <- add_bounded_check(
    bounded_checks, df, b, "stratum_pref",
    "stratum_pref must be between 0 and 100 when non-missing",
    function(x) x < 0 | x > 100
  )

  bounded_checks <- add_bounded_check(
    bounded_checks, df, b, "time_bright_prop",
    "time_bright_prop must be between 0 and 1 when non-missing",
    function(x) x < 0 | x > 1
  )

  ph_col <- first_existing_col(df, c("pH", "ph"))
  if (!is.na(ph_col)) {
    bounded_checks <- add_bounded_check(
      bounded_checks, df, b, ph_col,
      "pH values outside 0-14 are impossible",
      function(x) x < 0 | x > 14
    )
  }
}

bounded_checks
```

```{r export-range-checks}
write_table(range_summary, "01_qc_range_summary.csv")
write_table(bounded_checks, "01_qc_bounded_range_checks.csv")
```

---

# 01.10 Duration flag audit

This audit summarizes duration-related flags created during wrangling and range checks.

```{r duration-audit}
duration_flags <- c(
  "ldt_time_bright_gt_480_flag",
  "ldt_time_bright_gt_600_flag",
  "ntt_mov_total_gt_480_flag"
)

duration_audit <- do.call(
  rbind,
  lapply(names(battery_data), function(b) {
    df <- battery_data[[b]]
    available <- has_cols(df, duration_flags)

    if (length(available) == 0) {
      return(data.frame(
        battery = b,
        flag = NA_character_,
        flagged_n = NA_integer_,
        stringsAsFactors = FALSE
      ))
    }

    data.frame(
      battery = b,
      flag = available,
      flagged_n = vapply(available, function(v) count_true(df[[v]]), integer(1)),
      stringsAsFactors = FALSE
    )
  })
)

duration_audit
```

```{r export-duration-audit}
write_table(duration_audit, "01_qc_protocol_duration_flags.csv")
```

---

# 01.11 Derived-variable consistency audit

Derived-variable checks are diagnostic. A failed check means the definition must be reviewed; it is not an automatic exclusion.

```{r derived-variable-audit}
derived_audit <- data.frame(
  battery = character(),
  check = character(),
  n_checked = integer(),
  n_flagged = integer(),
  max_abs_difference = numeric(),
  note = character(),
  stringsAsFactors = FALSE
)

add_derived_check <- function(out, battery_label, check_name, observed, expected, tolerance, note) {
  observed <- as_num(observed)
  expected <- as_num(expected)

  ok_index <- !is.na(observed) & !is.na(expected)
  diff <- abs(observed[ok_index] - expected[ok_index])

  rbind(
    out,
    data.frame(
      battery = battery_label,
      check = check_name,
      n_checked = sum(ok_index),
      n_flagged = sum(diff > tolerance, na.rm = TRUE),
      max_abs_difference = if (length(diff) == 0) NA_real_ else max(diff, na.rm = TRUE),
      note = note,
      stringsAsFactors = FALSE
    )
  )
}

for (b in names(battery_data)) {
  df <- battery_data[[b]]

  if (all(c("dist_total", "mov_total", "vel_mean") %in% names(df))) {
    dist <- as_num(df$dist_total)
    mov <- as_num(df$mov_total)
    expected_vel <- ifelse(!is.na(mov) & mov > 0, dist / mov, NA_real_)

    derived_audit <- add_derived_check(
      derived_audit,
      b,
      "vel_mean_vs_dist_total_div_mov_total",
      df$vel_mean,
      expected_vel,
      tolerance = 0.05,
      note = "Checks whether vel_mean approximates dist_total / mov_total when mov_total > 0."
    )
  }

  if (all(c("time_bright", "num_changes", "mtpc") %in% names(df))) {
    time_bright <- as_num(df$time_bright)
    num_changes <- as_num(df$num_changes)
    expected_mtpc <- ifelse(!is.na(num_changes) & num_changes > 0, time_bright / num_changes, time_bright)

    derived_audit <- add_derived_check(
      derived_audit,
      b,
      "mtpc_vs_time_bright_div_num_changes",
      df$mtpc,
      expected_mtpc,
      tolerance = 0.05,
      note = "Checks documented mtpc edge-case rule: if num_changes == 0, mtpc equals time_bright."
    )
  }

  if (all(c("last_flux", "time_in_last_flux", "resistance_index") %in% names(df))) {
    last_flux <- as_num(df$last_flux)
    time_in_last_flux <- as_num(df$time_in_last_flux)

    expected_resistance <- rep(NA_real_, nrow(df))
    valid <- !is.na(last_flux) & !is.na(time_in_last_flux) & last_flux >= 0 & time_in_last_flux >= 0
    expected_resistance[valid] <- vapply(which(valid), function(i) {
      flux_sum <- sum(seq_len(max(0, as.integer(last_flux[i]) - 1)))
      flux_sum + last_flux[i] * time_in_last_flux[i] / 60
    }, numeric(1))

    derived_audit <- add_derived_check(
      derived_audit,
      b,
      "resistance_index_vs_last_flux_time_in_last_flux",
      df$resistance_index,
      expected_resistance,
      tolerance = 0.05,
      note = "Checks the documented resistance index formula."
    )
  }
}

derived_audit
```

```{r export-derived-audit}
write_table(derived_audit, "01_qc_derived_variable_audit.csv")
```

---

# 01.12 Condition factor audit

Kc, Kf, and condition-factor columns are treated as the same operational construct only if the formula and range are sound. The expected formula is a Fulton-type condition factor:

`K = 100 * weight / length^3`

The exact weight and length source columns are detected from available aliases.

```{r condition-factor-audit}
condition_aliases <- c("condition_factor_K", "Kc", "Kf", "kf", "condition_factor", "fulton_k")
weight_aliases <- c("wt", "weight", "W", "Weight")
length_aliases <- c("ls", "standard_length", "length", "lt", "Lt", "Ls")

condition_factor_audit <- do.call(
  rbind,
  lapply(names(battery_data), function(b) {
    df <- battery_data[[b]]

    k_col <- first_existing_col(df, condition_aliases)
    w_col <- first_existing_col(df, weight_aliases)
    l_col <- first_existing_col(df, length_aliases)

    if (is.na(k_col) || is.na(w_col) || is.na(l_col)) {
      return(data.frame(
        battery = b,
        condition_factor_column = k_col,
        weight_column = w_col,
        length_column = l_col,
        n_checked = 0,
        n_flagged = NA_integer_,
        max_abs_difference = NA_real_,
        median_condition_factor = if (!is.na(k_col)) safe_median(df[[k_col]]) else NA_real_,
        conclusion = "Insufficient columns to audit condition factor formula.",
        stringsAsFactors = FALSE
      ))
    }

    observed <- as_num(df[[k_col]])
    weight <- as_num(df[[w_col]])
    length <- as_num(df[[l_col]])
    expected <- ifelse(!is.na(weight) & !is.na(length) & length > 0, 100 * weight / (length^3), NA_real_)
    ok <- !is.na(observed) & !is.na(expected)
    diff <- abs(observed[ok] - expected[ok])

    data.frame(
      battery = b,
      condition_factor_column = k_col,
      weight_column = w_col,
      length_column = l_col,
      n_checked = sum(ok),
      n_flagged = sum(diff > 0.05, na.rm = TRUE),
      max_abs_difference = if (length(diff) == 0) NA_real_ else max(diff, na.rm = TRUE),
      median_condition_factor = safe_median(observed),
      conclusion = if (sum(ok) == 0) {
        "No non-missing paired values to audit."
      } else if (sum(diff > 0.05, na.rm = TRUE) == 0) {
        "Formula-consistent condition factor detected."
      } else {
        "Condition factor requires review before PCA inclusion."
      },
      stringsAsFactors = FALSE
    )
  })
)

condition_factor_audit
```

```{r export-condition-factor-audit}
write_table(condition_factor_audit, "01_qc_condition_factor_audit.csv")
```

---

# 01.13 QC flag summary

```{r qc-flag-summary}
all_flag_cols <- sort(unique(unlist(lapply(battery_data, function(df) names(df)[grepl("_flag$", names(df))]))))

if (length(all_flag_cols) == 0) {
  qc_flag_summary <- data.frame(
    battery = names(battery_data),
    flag = NA_character_,
    flagged_n = NA_integer_,
    stringsAsFactors = FALSE
  )
} else {
  qc_flag_summary <- do.call(
    rbind,
    lapply(names(battery_data), function(b) {
      df <- battery_data[[b]]
      available <- has_cols(df, all_flag_cols)

      if (length(available) == 0) {
        return(data.frame(
          battery = b,
          flag = NA_character_,
          flagged_n = NA_integer_,
          stringsAsFactors = FALSE
        ))
      }

      data.frame(
        battery = b,
        flag = available,
        flagged_n = vapply(available, function(v) count_true(df[[v]]), integer(1)),
        stringsAsFactors = FALSE
      )
    })
  )

  qc_flag_summary <- qc_flag_summary[order(qc_flag_summary$battery, -qc_flag_summary$flagged_n), ]
}

manual_review_summary <- do.call(
  rbind,
  lapply(names(battery_data), function(b) {
    df <- battery_data[[b]]

    data.frame(
      battery = b,
      n_rows = nrow(df),
      manual_review_flag_n = if ("manual_review_flag" %in% names(df)) count_true(df$manual_review_flag) else NA_integer_,
      stringsAsFactors = FALSE
    )
  })
)

qc_flag_summary
manual_review_summary
```

```{r export-qc-flags}
write_table(qc_flag_summary, "01_qc_flag_summary.csv")
write_table(manual_review_summary, "01_qc_manual_review_summary.csv")
```

---

# 01.14 Wrangling logs and exclusion log audit

```{r exclusion-log-audit}
if (is.null(wrangling_exclusion_log)) {
  exclusion_log_audit <- data.frame(
    item = "wrangling_exclusion_log",
    status = "missing",
    n_rows = NA_integer_,
    n_columns = NA_integer_,
    note = "No wrangling exclusion log found in outputs/logs.",
    stringsAsFactors = FALSE
  )
  exclusion_summary <- data.frame()
} else {
  exclusion_log_audit <- data.frame(
    item = "wrangling_exclusion_log",
    status = "found",
    n_rows = nrow(wrangling_exclusion_log),
    n_columns = ncol(wrangling_exclusion_log),
    note = "Wrangling exclusion log available for downstream traceability.",
    stringsAsFactors = FALSE
  )

  summary_cols <- has_cols(
    wrangling_exclusion_log,
    c("battery", "exclusion_scope", "exclusion_type", "retained_in_processed_dataset", "is_global_exclusion")
  )

  if (length(summary_cols) >= 2) {
    exclusion_summary <- as.data.frame(table(wrangling_exclusion_log[summary_cols]), stringsAsFactors = FALSE)
    names(exclusion_summary)[ncol(exclusion_summary)] <- "n"
  } else {
    exclusion_summary <- data.frame()
  }
}

if (is.null(wrangling_qc_log)) {
  wrangling_qc_log_audit <- data.frame(
    item = "wrangling_qc_log",
    status = "missing",
    n_rows = NA_integer_,
    n_columns = NA_integer_,
    stringsAsFactors = FALSE
  )
} else {
  wrangling_qc_log_audit <- data.frame(
    item = "wrangling_qc_log",
    status = "found",
    n_rows = nrow(wrangling_qc_log),
    n_columns = ncol(wrangling_qc_log),
    stringsAsFactors = FALSE
  )
}

exclusion_log_audit
wrangling_qc_log_audit
exclusion_summary
```

```{r export-exclusion-audit}
write_table(exclusion_log_audit, "01_qc_exclusion_log_audit.csv")
write_table(wrangling_qc_log_audit, "01_qc_wrangling_qc_log_audit.csv")
if (nrow(exclusion_summary) > 0) {
  write_table(exclusion_summary, "01_qc_exclusion_summary.csv")
}
```

---

# 01.15 Save audited datasets

The audited datasets preserve the wrangled battery-level data after this QC audit. They are the only datasets that downstream R notebooks should load.

```{r save-audited-datasets}
bt1_audited <- bt1
bt2_audited <- bt2
bt3_audited <- bt3

bt1_audited_path <- file.path(paths$data_processed, "bt1_audited.csv")
bt2_audited_path <- file.path(paths$data_processed, "bt2_audited.csv")
bt3_audited_path <- file.path(paths$data_processed, "bt3_audited.csv")

write.csv(bt1_audited, bt1_audited_path, row.names = FALSE)
write.csv(bt2_audited, bt2_audited_path, row.names = FALSE)
write.csv(bt3_audited, bt3_audited_path, row.names = FALSE)

audited_dataset_outputs <- data.frame(
  dataset = c("bt1_audited", "bt2_audited", "bt3_audited"),
  path = c(bt1_audited_path, bt2_audited_path, bt3_audited_path),
  rows = c(nrow(bt1_audited), nrow(bt2_audited), nrow(bt3_audited)),
  columns = c(ncol(bt1_audited), ncol(bt2_audited), ncol(bt3_audited)),
  stringsAsFactors = FALSE
)

audited_dataset_outputs
```

```{r export-audited-output-summary}
write_table(audited_dataset_outputs, "01_qc_audited_dataset_outputs.csv")
```

---

# 01.16 QC conclusion

This conclusion is non-inferential. It summarizes what was checked and what must be carried into Notebook 02.

```{r qc-conclusion}
qc_conclusion <- data.frame(
  item = c(
    "Wrangled battery files loaded",
    "Battery identity checked",
    "Schema and endpoint availability checked",
    "Fish-ID integrity checked",
    "Treatment by exposure cells counted",
    "Datetime audit exported",
    "Missingness audit exported",
    "Range and duration audits exported",
    "Derived-variable audit exported",
    "Condition factor audited",
    "QC flag summary exported",
    "Wrangling logs inspected",
    "Audited datasets saved",
    "Inferential statistics performed"
  ),
  status = c(
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    TRUE,
    !is.null(wrangling_exclusion_log) || !is.null(wrangling_qc_log),
    TRUE,
    FALSE
  ),
  note = c(
    "BT1, BT2, and BT3 were loaded from data_processed.",
    "Battery identity was checked using file label, battery column, and presence flags where available.",
    "Column presence and endpoint-registry availability were checked across batteries.",
    "Within-battery duplication and cross-battery fish-ID overlap were audited.",
    "Cell sizes are available for downstream model-feasibility checks.",
    "Datetime is treated as audit metadata, not as an inferential covariate here.",
    "Missingness is documented, not automatically treated as exclusion.",
    "Duration and impossible-value issues are flagged for later sensitivity decisions.",
    "Derived variables are checked diagnostically only.",
    "Kc/Kf/condition-factor consistency is checked before PCA use.",
    "Manual-review and other flags are summarized for downstream decisions.",
    "Wrangling logs are inspected if present.",
    "Audited datasets are written to data_processed for Notebook 02.",
    "No inferential statistics are run in Notebook 01."
  ),
  stringsAsFactors = FALSE
)

qc_conclusion
```

```{r export-qc-conclusion}
write_table(qc_conclusion, "01_qc_conclusion.csv")

qc_manifest <- list(
  loaded_files = loaded_files,
  battery_identity = battery_identity,
  dataset_structure = dataset_structure,
  fish_id_integrity = fish_id_integrity,
  cross_battery_overlap = cross_battery_overlap,
  cell_counts = cell_counts,
  condition_factor_audit = condition_factor_audit,
  audited_dataset_outputs = audited_dataset_outputs,
  qc_conclusion = qc_conclusion
)

saveRDS(qc_manifest, file.path(paths$outputs_logs, "01_qc_manifest.rds"))
```

The next notebook is `02_endpoint_hierarchy_and_assumptions.Rmd`. Despite the filename, its role is endpoint registry and model readiness. It should load only:

- `data_processed/bt1_audited.csv`
- `data_processed/bt2_audited.csv`
- `data_processed/bt3_audited.csv`

and use the endpoint registry plus this QC audit to decide which endpoints can be modeled, compared across battery contexts, and included in the single PCA.
