---
title: "03 Ethanol Factorial Models"
subtitle: "Endpoint-level treatment x exposure models"
author: "Caique Ortiz Colombo Arnoldi Fabris"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

# Purpose

This notebook tests ethanol-related effects at the endpoint level.

The target model is:

`endpoint ~ treatment * exposure`

The notebook uses the audited BT1 full-battery dataset and the endpoint registry/model-readiness outputs generated in Notebook 02. It does not test battery-context effects and it does not run PCA. Battery comparisons belong to Notebook 04. PCA belongs to Notebook 05. Robustness and sensitivity checks belong to Notebook 06.

The endpoint registry controls which variables are modeled. There is no primary/secondary hierarchy in this notebook. Endpoints are handled according to their practical role, data type, availability, and model-readiness status.

# 03.1 Load project paths, data, and registry outputs

```{r project-root}
get_project_root <- function() {
  if (requireNamespace("here", quietly = TRUE)) {
    return(here::here())
  }

  if (requireNamespace("rprojroot", quietly = TRUE)) {
    root <- tryCatch(
      as.character(rprojroot::find_root(rprojroot::is_rstudio_project)),
      error = function(e) NA_character_
    )
    if (!is.na(root)) return(root)
  }

  normalizePath(".", winslash = "/", mustWork = TRUE)
}

project_root <- get_project_root()

paths <- list(
  root = project_root,
  data_processed = file.path(project_root, "data_processed"),
  outputs = file.path(project_root, "outputs"),
  outputs_logs = file.path(project_root, "outputs", "logs"),
  outputs_tables = file.path(project_root, "outputs", "tables"),
  outputs_models = file.path(project_root, "outputs", "models"),
  outputs_figures = file.path(project_root, "outputs", "figures")
)

for (p in c(paths$outputs, paths$outputs_logs, paths$outputs_tables, paths$outputs_models, paths$outputs_figures)) {
  dir.create(p, recursive = TRUE, showWarnings = FALSE)
}

paths
```

```{r helper-functions}
read_csv_required <- function(path) {
  if (!file.exists(path)) stop("Required file not found: ", path)
  read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
}

read_csv_optional <- function(path) {
  if (!file.exists(path)) return(NULL)
  read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
}

write_table <- function(df, filename, folder = paths$outputs_tables) {
  path <- file.path(folder, filename)
  write.csv(df, path, row.names = FALSE)
  path
}

write_log <- function(df, filename) {
  path <- file.path(paths$outputs_logs, filename)
  write.csv(df, path, row.names = FALSE)
  path
}

save_model_object <- function(object, filename) {
  path <- file.path(paths$outputs_models, filename)
  saveRDS(object, path)
  path
}

safe_numeric <- function(x) {
  suppressWarnings(as.numeric(x))
}

as_bool <- function(x) {
  if (is.logical(x)) return(x)
  if (is.numeric(x)) return(x == 1)
  tolower(as.character(x)) %in% c("true", "1", "yes", "y")
}

standardize_exposure <- function(x) {
  y <- trimws(as.character(x))
  y <- ifelse(y %in% c("1", "1h", "1H"), "1h",
    ifelse(y %in% c("24", "24h", "24H"), "24h",
      ifelse(y %in% c("96", "96h", "96H"), "96h", y)
    )
  )
  factor(y, levels = c("1h", "24h", "96h"))
}

standardize_factors <- function(df) {
  if ("treatment" %in% names(df)) {
    df$treatment <- factor(as.character(df$treatment), levels = c("0", "0.5", "1"))
  }
  if ("exposure" %in% names(df)) {
    df$exposure <- standardize_exposure(df$exposure)
  }
  if ("battery" %in% names(df)) {
    df$battery <- factor(as.character(df$battery), levels = c("BT1", "BT2", "BT3"))
  }
  df
}

clean_for_model <- function(df, resolved_variable) {
  if (is.na(resolved_variable) || !resolved_variable %in% names(df)) {
    return(data.frame())
  }

  required <- c("treatment", "exposure")
  if (!all(required %in% names(df))) {
    return(data.frame())
  }

  out <- df

  if ("global_exclusion_flag" %in% names(out)) {
    out <- out[!as_bool(out$global_exclusion_flag), ]
  }

  out$value <- safe_numeric(out[[resolved_variable]])
  out <- out[!is.na(out$value) & !is.na(out$treatment) & !is.na(out$exposure), ]

  if (nrow(out) == 0) return(out)

  out$treatment <- droplevels(out$treatment)
  out$exposure <- droplevels(out$exposure)
  out
}

has_factorial_structure <- function(df) {
  if (nrow(df) < 6) return(FALSE)
  if (!all(c("treatment", "exposure", "value") %in% names(df))) return(FALSE)
  if (nlevels(df$treatment) < 2) return(FALSE)
  if (nlevels(df$exposure) < 2) return(FALSE)
  TRUE
}

extract_p_column <- function(df) {
  candidates <- grep("Pr|p.value|p_value|p-value", names(df), ignore.case = TRUE, value = TRUE)
  if (length(candidates) == 0) return(rep(NA_real_, nrow(df)))
  suppressWarnings(as.numeric(df[[candidates[1]]]))
}

extract_stat_column <- function(df) {
  candidates <- grep("F value|F$|statistic|Chisq|Chi|z value|t value", names(df), ignore.case = TRUE, value = TRUE)
  if (length(candidates) == 0) return(rep(NA_real_, nrow(df)))
  suppressWarnings(as.numeric(df[[candidates[1]]]))
}

extract_df_column <- function(df) {
  candidates <- grep("^Df$|Df|df|num Df", names(df), ignore.case = TRUE, value = TRUE)
  if (length(candidates) == 0) return(rep(NA_real_, nrow(df)))
  suppressWarnings(as.numeric(df[[candidates[1]]]))
}

empty_model_result <- function(endpoint, resolved_variable, domain, source_test, data_type, model_family, model_method, model_note) {
  data.frame(
    endpoint = endpoint,
    resolved_variable = resolved_variable,
    domain = domain,
    source_test = source_test,
    data_type = data_type,
    model_family = model_family,
    dataset = "BT1",
    model_method = model_method,
    term = NA_character_,
    df = NA_real_,
    statistic = NA_real_,
    p_value = NA_real_,
    model_note = model_note,
    stringsAsFactors = FALSE
  )
}

tidy_model_table <- function(anova_object, endpoint, resolved_variable, domain, source_test, data_type, model_family, model_method, model_note) {
  tab <- as.data.frame(anova_object)
  tab$term <- rownames(tab)
  rownames(tab) <- NULL

  data.frame(
    endpoint = endpoint,
    resolved_variable = resolved_variable,
    domain = domain,
    source_test = source_test,
    data_type = data_type,
    model_family = model_family,
    dataset = "BT1",
    model_method = model_method,
    term = tab$term,
    df = extract_df_column(tab),
    statistic = extract_stat_column(tab),
    p_value = extract_p_column(tab),
    model_note = model_note,
    stringsAsFactors = FALSE
  )
}

coefficient_table <- function(model, endpoint, resolved_variable, domain, source_test, data_type, model_family, transform = "none") {
  coefs <- tryCatch(summary(model)$coefficients, error = function(e) NULL)
  if (is.null(coefs)) return(data.frame())

  out <- as.data.frame(coefs)
  out$term <- rownames(out)
  rownames(out) <- NULL

  estimate_col <- grep("Estimate", names(out), value = TRUE)[1]
  se_col <- grep("Std. Error", names(out), value = TRUE)[1]
  p_col <- grep("Pr\\(", names(out), value = TRUE)[1]

  estimates <- suppressWarnings(as.numeric(out[[estimate_col]]))
  ses <- if (!is.na(se_col)) suppressWarnings(as.numeric(out[[se_col]])) else rep(NA_real_, nrow(out))
  pvals <- if (!is.na(p_col)) suppressWarnings(as.numeric(out[[p_col]])) else rep(NA_real_, nrow(out))

  if (transform == "exp") {
    estimate_out <- exp(estimates)
    ci_low <- exp(estimates - 1.96 * ses)
    ci_high <- exp(estimates + 1.96 * ses)
    scale <- "exponentiated_coefficient"
  } else {
    estimate_out <- estimates
    ci_low <- estimates - 1.96 * ses
    ci_high <- estimates + 1.96 * ses
    scale <- "model_coefficient"
  }

  data.frame(
    endpoint = endpoint,
    resolved_variable = resolved_variable,
    domain = domain,
    source_test = source_test,
    data_type = data_type,
    model_family = model_family,
    dataset = "BT1",
    term = out$term,
    effect_scale = scale,
    estimate = estimate_out,
    ci_low_approx = ci_low,
    ci_high_approx = ci_high,
    p_value = pvals,
    stringsAsFactors = FALSE
  )
}

make_binary01 <- function(x) {
  if (is.numeric(x) || is.logical(x)) {
    y <- as.numeric(x)
    y[!y %in% c(0, 1)] <- NA_real_
    return(y)
  }

  y <- tolower(trimws(as.character(x)))
  out <- rep(NA_real_, length(y))
  out[y %in% c("1", "true", "yes", "bright", "light", "claro", "white")] <- 1
  out[y %in% c("0", "false", "no", "dark", "escuro", "black")] <- 0
  out
}
```

```{r load-inputs}
bt1 <- read_csv_required(file.path(paths$data_processed, "bt1_audited.csv"))
bt2 <- read_csv_required(file.path(paths$data_processed, "bt2_audited.csv"))
bt3 <- read_csv_required(file.path(paths$data_processed, "bt3_audited.csv"))

bt1 <- standardize_factors(bt1)
bt2 <- standardize_factors(bt2)
bt3 <- standardize_factors(bt3)

endpoint_registry <- read_csv_optional(file.path(paths$outputs_logs, "02_endpoint_registry.csv"))
model_decision_matrix <- read_csv_optional(file.path(paths$outputs_logs, "02_model_decision_matrix.csv"))

if (is.null(model_decision_matrix)) {
  if (is.null(endpoint_registry)) {
    stop("Notebook 03 requires outputs/logs/02_model_decision_matrix.csv or outputs/logs/02_endpoint_registry.csv from Notebook 02.")
  }
  model_decision_matrix <- endpoint_registry
}

load_summary <- data.frame(
  object = c("bt1_audited", "bt2_audited", "bt3_audited", "endpoint_registry", "model_decision_matrix"),
  loaded = c(TRUE, TRUE, TRUE, !is.null(endpoint_registry), !is.null(model_decision_matrix)),
  rows = c(nrow(bt1), nrow(bt2), nrow(bt3), if (is.null(endpoint_registry)) NA_integer_ else nrow(endpoint_registry), nrow(model_decision_matrix)),
  columns = c(ncol(bt1), ncol(bt2), ncol(bt3), if (is.null(endpoint_registry)) NA_integer_ else ncol(endpoint_registry), ncol(model_decision_matrix)),
  stringsAsFactors = FALSE
)

write_log(load_summary, "03_ethanol_model_input_load_summary.csv")
load_summary
```

```{r package-availability}
package_availability <- data.frame(
  package = c("ARTool", "car", "emmeans", "MASS", "logistf", "broom"),
  available = c(
    requireNamespace("ARTool", quietly = TRUE),
    requireNamespace("car", quietly = TRUE),
    requireNamespace("emmeans", quietly = TRUE),
    requireNamespace("MASS", quietly = TRUE),
    requireNamespace("logistf", quietly = TRUE),
    requireNamespace("broom", quietly = TRUE)
  ),
  role = c(
    "Aligned-rank factorial models for non-normal continuous or bounded endpoints.",
    "Type-III ANOVA fallback for linear models.",
    "Estimated marginal means and post-hoc contrasts when justified.",
    "Negative-binomial models for overdispersed counts.",
    "Firth logistic regression for sparse binary endpoints.",
    "Optional model tidying."
  ),
  stringsAsFactors = FALSE
)

write_log(package_availability, "03_ethanol_model_package_availability.csv")
package_availability
```

# 03.2 Define the endpoint-level model set

The endpoint set is taken from Notebook 02. An endpoint is eligible here when it is available in BT1 and marked as a model candidate in the endpoint registry/model-readiness table.

This section does not rank endpoints. It records which endpoint-level models are run and which are skipped.

```{r model-set}
required_cols <- c("variable", "domain", "source_test", "data_type", "resolved_variable_BT1", "available_BT1")
missing_cols <- setdiff(required_cols, names(model_decision_matrix))
if (length(missing_cols) > 0) {
  stop("Model decision matrix is missing required columns: ", paste(missing_cols, collapse = ", "))
}

if (!"model_candidate" %in% names(model_decision_matrix)) {
  if ("endpoint_model" %in% names(model_decision_matrix)) {
    model_decision_matrix$model_candidate <- model_decision_matrix$endpoint_model %in% c("yes", "yes_if_justified", "yes_if_feasible", "optional_supportive")
  } else {
    model_decision_matrix$model_candidate <- FALSE
  }
}

if (!"recommended_model_family" %in% names(model_decision_matrix)) {
  model_decision_matrix$recommended_model_family <- ifelse(
    grepl("binary", model_decision_matrix$data_type, ignore.case = TRUE),
    "binary_logistic_or_firth",
    ifelse(
      model_decision_matrix$data_type %in% c("count", "ordinal/step"),
      "count_glm_or_negative_binomial",
      "continuous_factorial_art_or_lm"
    )
  )
}

model_plan <- model_decision_matrix[
  model_decision_matrix$model_candidate &
    model_decision_matrix$available_BT1 &
    !is.na(model_decision_matrix$resolved_variable_BT1),
]

model_plan <- model_plan[!model_plan$data_type %in% c("identifier", "design", "datetime", "categorical"), ]
model_plan <- model_plan[order(model_plan$source_test, model_plan$domain, model_plan$variable), ]

skip_plan <- model_decision_matrix[!(model_decision_matrix$variable %in% model_plan$variable), ]

write_table(model_plan, "03_ethanol_model_plan.csv")
write_log(model_plan, "03_ethanol_model_plan.csv")
write_log(skip_plan, "03_ethanol_model_skipped_registry_rows.csv")

model_plan
```

# 03.3 State the factorial question

The common target is `value ~ treatment * exposure`.

`treatment` tests ethanol concentration. `exposure` tests exposure duration. The interaction tests whether the ethanol concentration effect changes across exposure durations.

```{r factorial-question}
factorial_question <- data.frame(
  item = c(
    "model_target",
    "analysis_dataset",
    "treatment_main_effect",
    "exposure_main_effect",
    "interaction_effect",
    "battery_rule",
    "pca_rule"
  ),
  definition = c(
    "endpoint value ~ treatment * exposure",
    "BT1 audited full-battery dataset",
    "Association between ethanol concentration and the endpoint within the factorial model.",
    "Association between exposure duration and the endpoint within the factorial model.",
    "Whether ethanol concentration effects differ by exposure duration.",
    "Battery-context comparisons are not run in Notebook 03.",
    "PCA is not run in Notebook 03 and does not replace endpoint-level models."
  ),
  stringsAsFactors = FALSE
)

write_log(factorial_question, "03_ethanol_factorial_question.csv")
factorial_question
```

# 03.4 Model functions

Model choice is based on the endpoint registry and data type.

Continuous, bounded, derived, physiological, and biometric endpoints are modeled as factorial endpoints using ART-ANOVA when available, with a linear-model ANOVA fallback. Binary endpoints use Firth logistic regression when available, with logistic regression fallback. Count-like endpoints use negative-binomial models when available and variance suggests overdispersion, with Poisson fallback.

```{r model-functions-continuous}
fit_continuous_endpoint <- function(df, row) {
  endpoint <- row$variable
  resolved <- row$resolved_variable_BT1
  domain <- row$domain
  source_test <- row$source_test
  data_type <- row$data_type
  model_family <- row$recommended_model_family

  d <- clean_for_model(df, resolved)

  if (!has_factorial_structure(d)) {
    return(list(
      result = empty_model_result(endpoint, resolved, domain, source_test, data_type, model_family, "not_run", "Endpoint unavailable or insufficient factorial structure."),
      effect = data.frame(),
      posthoc = data.frame(),
      model_path = NA_character_,
      n_model = nrow(d)
    ))
  }

  formula_object <- value ~ treatment * exposure

  if (requireNamespace("ARTool", quietly = TRUE)) {
    fit_art <- tryCatch(ARTool::art(formula_object, data = d), error = function(e) e)
    if (!inherits(fit_art, "error")) {
      a <- tryCatch(anova(fit_art), error = function(e) e)
      if (!inherits(a, "error")) {
        result <- tidy_model_table(
          a, endpoint, resolved, domain, source_test, data_type, model_family,
          "ART-ANOVA",
          "Aligned Rank Transform factorial model. Interpret endpoint effects from treatment, exposure, and treatment:exposure terms."
        )

        model_path <- save_model_object(
          list(model = fit_art, analysis_data = d, endpoint = endpoint, resolved_variable = resolved, method = "ART-ANOVA"),
          paste0("03_model_ethanol_", endpoint, "_art.rds")
        )

        return(list(result = result, effect = data.frame(), posthoc = data.frame(), model_path = model_path, n_model = nrow(d)))
      }
    }
  }

  fit_lm <- tryCatch(lm(formula_object, data = d), error = function(e) e)
  if (inherits(fit_lm, "error")) {
    return(list(
      result = empty_model_result(endpoint, resolved, domain, source_test, data_type, model_family, "not_run", paste("Linear fallback failed:", fit_lm$message)),
      effect = data.frame(),
      posthoc = data.frame(),
      model_path = NA_character_,
      n_model = nrow(d)
    ))
  }

  if (requireNamespace("car", quietly = TRUE)) {
    a <- tryCatch(car::Anova(fit_lm, type = 3), error = function(e) anova(fit_lm))
    method <- "linear model ANOVA fallback; Type-III if car::Anova succeeded"
  } else {
    a <- anova(fit_lm)
    method <- "linear model ANOVA fallback; sequential ANOVA because car is unavailable"
  }

  result <- tidy_model_table(
    a, endpoint, resolved, domain, source_test, data_type, model_family,
    method,
    "Fallback model. Interpret with Notebook 02 diagnostics and Notebook 06 robustness checks."
  )

  effect <- coefficient_table(fit_lm, endpoint, resolved, domain, source_test, data_type, model_family, transform = "none")

  posthoc <- data.frame()
  pvals <- result$p_value[!is.na(result$p_value)]
  if (length(pvals) > 0 && any(pvals < 0.05) && requireNamespace("emmeans", quietly = TRUE)) {
    treatment_contrasts <- tryCatch(
      as.data.frame(emmeans::contrast(emmeans::emmeans(fit_lm, ~ treatment | exposure), method = "pairwise")),
      error = function(e) data.frame()
    )
    exposure_contrasts <- tryCatch(
      as.data.frame(emmeans::contrast(emmeans::emmeans(fit_lm, ~ exposure | treatment), method = "pairwise")),
      error = function(e) data.frame()
    )

    if (nrow(treatment_contrasts) > 0) {
      treatment_contrasts$contrast_family <- "treatment_within_exposure"
      treatment_contrasts$endpoint <- endpoint
      treatment_contrasts$resolved_variable <- resolved
      treatment_contrasts$dataset <- "BT1"
    }
    if (nrow(exposure_contrasts) > 0) {
      exposure_contrasts$contrast_family <- "exposure_within_treatment"
      exposure_contrasts$endpoint <- endpoint
      exposure_contrasts$resolved_variable <- resolved
      exposure_contrasts$dataset <- "BT1"
    }
    posthoc <- rbind(treatment_contrasts, exposure_contrasts)
  }

  model_path <- save_model_object(
    list(model = fit_lm, analysis_data = d, endpoint = endpoint, resolved_variable = resolved, method = method),
    paste0("03_model_ethanol_", endpoint, "_lm_fallback.rds")
  )

  list(result = result, effect = effect, posthoc = posthoc, model_path = model_path, n_model = nrow(d))
}
```

```{r model-functions-binary}
fit_binary_endpoint <- function(df, row) {
  endpoint <- row$variable
  resolved <- row$resolved_variable_BT1
  domain <- row$domain
  source_test <- row$source_test
  data_type <- row$data_type
  model_family <- row$recommended_model_family

  d <- df
  if ("global_exclusion_flag" %in% names(d)) d <- d[!as_bool(d$global_exclusion_flag), ]

  if (is.na(resolved) || !resolved %in% names(d)) {
    return(list(
      result = empty_model_result(endpoint, resolved, domain, source_test, data_type, model_family, "not_run", "Binary endpoint not available."),
      effect = data.frame(),
      posthoc = data.frame(),
      model_path = NA_character_,
      n_model = 0
    ))
  }

  d$value <- make_binary01(d[[resolved]])
  d <- d[!is.na(d$value) & !is.na(d$treatment) & !is.na(d$exposure), ]
  d$treatment <- droplevels(d$treatment)
  d$exposure <- droplevels(d$exposure)

  if (!has_factorial_structure(d) || length(unique(d$value)) < 2) {
    return(list(
      result = empty_model_result(endpoint, resolved, domain, source_test, data_type, model_family, "not_run", "Insufficient valid binary observations or insufficient factorial structure."),
      effect = data.frame(),
      posthoc = data.frame(),
      model_path = NA_character_,
      n_model = nrow(d)
    ))
  }

  formula_object <- value ~ treatment * exposure

  if (requireNamespace("logistf", quietly = TRUE)) {
    fit <- tryCatch(logistf::logistf(formula_object, data = d), error = function(e) e)
    if (!inherits(fit, "error")) {
      coefs <- data.frame(
        endpoint = endpoint,
        resolved_variable = resolved,
        domain = domain,
        source_test = source_test,
        data_type = data_type,
        model_family = model_family,
        dataset = "BT1",
        term = names(fit$coefficients),
        effect_scale = "odds_ratio",
        estimate = exp(as.numeric(fit$coefficients)),
        ci_low_approx = exp(as.numeric(fit$ci.lower)),
        ci_high_approx = exp(as.numeric(fit$ci.upper)),
        p_value = as.numeric(fit$prob),
        stringsAsFactors = FALSE
      )

      result <- data.frame(
        endpoint = endpoint,
        resolved_variable = resolved,
        domain = domain,
        source_test = source_test,
        data_type = data_type,
        model_family = model_family,
        dataset = "BT1",
        model_method = "Firth logistic regression",
        term = coefs$term,
        df = NA_real_,
        statistic = NA_real_,
        p_value = coefs$p_value,
        model_note = "Coefficient-level Firth logistic results. No-choice observations are not valid binary outcomes and are excluded from this model.",
        stringsAsFactors = FALSE
      )

      model_path <- save_model_object(
        list(model = fit, analysis_data = d, endpoint = endpoint, resolved_variable = resolved, method = "Firth logistic regression"),
        paste0("03_model_ethanol_", endpoint, "_firth.rds")
      )

      return(list(result = result, effect = coefs, posthoc = data.frame(), model_path = model_path, n_model = nrow(d)))
    }
  }

  fit <- tryCatch(glm(formula_object, data = d, family = binomial()), error = function(e) e)
  if (inherits(fit, "error")) {
    return(list(
      result = empty_model_result(endpoint, resolved, domain, source_test, data_type, model_family, "not_run", paste("Binary model failed:", fit$message)),
      effect = data.frame(),
      posthoc = data.frame(),
      model_path = NA_character_,
      n_model = nrow(d)
    ))
  }

  a <- tryCatch(anova(fit, test = "Chisq"), error = function(e) e)
  result <- if (inherits(a, "error")) {
    empty_model_result(endpoint, resolved, domain, source_test, data_type, model_family, "logistic regression fallback", paste("Could not extract likelihood-ratio table:", a$message))
  } else {
    tidy_model_table(a, endpoint, resolved, domain, source_test, data_type, model_family, "logistic regression fallback", "Ordinary logistic regression fallback. Use Notebook 02 binary-feasibility diagnostics and Notebook 06 robustness checks.")
  }

  effect <- coefficient_table(fit, endpoint, resolved, domain, source_test, data_type, model_family, transform = "exp")

  model_path <- save_model_object(
    list(model = fit, analysis_data = d, endpoint = endpoint, resolved_variable = resolved, method = "logistic regression fallback"),
    paste0("03_model_ethanol_", endpoint, "_logistic_fallback.rds")
  )

  list(result = result, effect = effect, posthoc = data.frame(), model_path = model_path, n_model = nrow(d))
}
```

```{r model-functions-count}
fit_count_endpoint <- function(df, row) {
  endpoint <- row$variable
  resolved <- row$resolved_variable_BT1
  domain <- row$domain
  source_test <- row$source_test
  data_type <- row$data_type
  model_family <- row$recommended_model_family

  d <- clean_for_model(df, resolved)

  if (!has_factorial_structure(d)) {
    return(list(
      result = empty_model_result(endpoint, resolved, domain, source_test, data_type, model_family, "not_run", "Count endpoint unavailable or insufficient factorial structure."),
      effect = data.frame(),
      posthoc = data.frame(),
      model_path = NA_character_,
      n_model = nrow(d)
    ))
  }

  d$value <- round(d$value)
  d <- d[d$value >= 0, ]
  if (nrow(d) == 0 || length(unique(d$value)) < 2) {
    return(list(
      result = empty_model_result(endpoint, resolved, domain, source_test, data_type, model_family, "not_run", "Count endpoint has insufficient non-negative variation."),
      effect = data.frame(),
      posthoc = data.frame(),
      model_path = NA_character_,
      n_model = nrow(d)
    ))
  }

  variance_to_mean <- ifelse(mean(d$value, na.rm = TRUE) > 0, var(d$value, na.rm = TRUE) / mean(d$value, na.rm = TRUE), NA_real_)
  formula_object <- value ~ treatment * exposure

  if (!is.na(variance_to_mean) && variance_to_mean > 1.5 && requireNamespace("MASS", quietly = TRUE)) {
    fit <- tryCatch(MASS::glm.nb(formula_object, data = d), error = function(e) e)
    model_method <- "negative-binomial GLM"
  } else {
    fit <- tryCatch(glm(formula_object, data = d, family = poisson()), error = function(e) e)
    model_method <- "Poisson GLM"
  }

  if (inherits(fit, "error")) {
    return(list(
      result = empty_model_result(endpoint, resolved, domain, source_test, data_type, model_family, "not_run", paste("Count model failed:", fit$message)),
      effect = data.frame(),
      posthoc = data.frame(),
      model_path = NA_character_,
      n_model = nrow(d)
    ))
  }

  a <- tryCatch(anova(fit, test = "Chisq"), error = function(e) e)
  result <- if (inherits(a, "error")) {
    empty_model_result(endpoint, resolved, domain, source_test, data_type, model_family, model_method, paste("Could not extract count-model ANOVA:", a$message))
  } else {
    tidy_model_table(a, endpoint, resolved, domain, source_test, data_type, model_family, model_method, paste0("Count endpoint model. Variance-to-mean ratio = ", round(variance_to_mean, 3), "."))
  }

  effect <- coefficient_table(fit, endpoint, resolved, domain, source_test, data_type, model_family, transform = "exp")

  model_path <- save_model_object(
    list(model = fit, analysis_data = d, endpoint = endpoint, resolved_variable = resolved, method = model_method, variance_to_mean = variance_to_mean),
    paste0("03_model_ethanol_", endpoint, "_count.rds")
  )

  list(result = result, effect = effect, posthoc = data.frame(), model_path = model_path, n_model = nrow(d))
}
```

```{r dispatch-model-function}
fit_registry_endpoint <- function(df, row) {
  data_type <- tolower(as.character(row$data_type))

  if (grepl("binary", data_type)) {
    return(fit_binary_endpoint(df, row))
  }

  if (data_type %in% c("count", "ordinal/step") || grepl("count", data_type)) {
    return(fit_count_endpoint(df, row))
  }

  fit_continuous_endpoint(df, row)
}
```

# 03.5 Run endpoint-level ethanol models

This section runs the planned endpoint-level models for BT1.

```{r run-endpoint-models}
model_objects <- lapply(seq_len(nrow(model_plan)), function(i) {
  fit_registry_endpoint(bt1, model_plan[i, ])
})

names(model_objects) <- model_plan$variable

factorial_results <- if (length(model_objects) == 0) {
  data.frame()
} else {
  do.call(rbind, lapply(model_objects, function(x) x$result))
}

effect_results <- if (length(model_objects) == 0) {
  data.frame()
} else {
  do.call(rbind, lapply(model_objects, function(x) x$effect))
}

posthoc_results <- if (length(model_objects) == 0) {
  data.frame()
} else {
  do.call(rbind, lapply(model_objects, function(x) x$posthoc))
}

model_paths <- if (length(model_objects) == 0) {
  data.frame()
} else {
  data.frame(
    endpoint = names(model_objects),
    model_path = vapply(model_objects, function(x) x$model_path, character(1)),
    n_model = vapply(model_objects, function(x) x$n_model, numeric(1)),
    stringsAsFactors = FALSE
  )
}

factorial_results
```

# 03.6 Multiple-testing correction

BH-FDR is applied to the endpoint-level factorial test table. Correction is applied within model-method families to avoid mixing unrelated coefficient-level and omnibus-style outputs indiscriminately.

```{r multiple-testing-correction}
if (nrow(factorial_results) > 0) {
  factorial_results$p_adj_bh <- NA_real_

  valid <- !is.na(factorial_results$p_value)
  families <- unique(factorial_results$model_method[valid])

  for (fam in families) {
    idx <- valid & factorial_results$model_method == fam
    factorial_results$p_adj_bh[idx] <- p.adjust(factorial_results$p_value[idx], method = "BH")
  }

  factorial_results$effect_flag_bh_0_05 <- !is.na(factorial_results$p_adj_bh) & factorial_results$p_adj_bh < 0.05
}

if (nrow(effect_results) > 0 && "p_value" %in% names(effect_results)) {
  effect_results$p_adj_bh <- NA_real_
  valid_effects <- !is.na(effect_results$p_value)
  effect_results$p_adj_bh[valid_effects] <- p.adjust(effect_results$p_value[valid_effects], method = "BH")
}

factorial_results
```

# 03.7 Domain summaries

These tables organize the model outputs by test/domain without changing the inferential hierarchy.

```{r domain-summaries}
domain_model_tables <- if (nrow(factorial_results) == 0) {
  list()
} else {
  split(factorial_results, factorial_results$source_test)
}

domain_summary <- if (nrow(factorial_results) == 0) {
  data.frame()
} else {
  aggregate(
    effect_flag_bh_0_05 ~ source_test + domain + endpoint,
    data = factorial_results,
    FUN = function(x) any(x, na.rm = TRUE)
  )
}

names(domain_summary)[names(domain_summary) == "effect_flag_bh_0_05"] <- "any_bh_significant_term"

domain_summary
```

```{r model-term-summary}
term_summary <- if (nrow(factorial_results) == 0) {
  data.frame()
} else {
  factorial_results[!is.na(factorial_results$term), c(
    "endpoint", "resolved_variable", "source_test", "domain", "model_method", "term", "p_value", "p_adj_bh", "effect_flag_bh_0_05", "model_note"
  )]
}

term_summary
```

# 03.8 Export outputs

```{r export-outputs}
write_table(factorial_results, "03_ethanol_factorial_results.csv")
write_table(effect_results, "03_ethanol_effect_estimates.csv")
write_table(posthoc_results, "03_ethanol_posthoc_contrasts.csv")
write_log(model_paths, "03_ethanol_model_paths.csv")
write_log(term_summary, "03_ethanol_model_term_summary.csv")
write_table(domain_summary, "03_ethanol_domain_summary.csv")

for (source in unique(factorial_results$source_test)) {
  safe_source <- gsub("[^A-Za-z0-9_]+", "_", tolower(source))
  write_table(factorial_results[factorial_results$source_test == source, ], paste0("03_ethanol_models_", safe_source, ".csv"))
}

export_summary <- data.frame(
  output = c(
    "03_ethanol_model_plan.csv",
    "03_ethanol_factorial_results.csv",
    "03_ethanol_effect_estimates.csv",
    "03_ethanol_posthoc_contrasts.csv",
    "03_ethanol_model_paths.csv",
    "03_ethanol_domain_summary.csv"
  ),
  location = c(
    paths$outputs_tables,
    paths$outputs_tables,
    paths$outputs_tables,
    paths$outputs_tables,
    paths$outputs_logs,
    paths$outputs_tables
  ),
  stringsAsFactors = FALSE
)

write_log(export_summary, "03_ethanol_export_summary.csv")
export_summary
```

# 03.9 Notebook conclusion

```{r conclusion}
notebook03_conclusion <- data.frame(
  item = c(
    "analysis_scope",
    "dataset",
    "model_target",
    "endpoint_selection",
    "battery_context",
    "pca_context",
    "next_notebook"
  ),
  conclusion = c(
    "Endpoint-level ethanol models were run for model-candidate endpoints available in BT1.",
    "BT1 audited full-battery dataset.",
    "endpoint value ~ treatment * exposure.",
    "Endpoint selection followed the Notebook 02 endpoint registry and model-readiness table.",
    "Battery-context effects were not tested here and remain assigned to Notebook 04.",
    "PCA was not run here and remains assigned to Notebook 05.",
    "Notebook 04 should test battery-context effects for comparable endpoint/cell combinations."
  ),
  stringsAsFactors = FALSE
)

write_log(notebook03_conclusion, "03_ethanol_notebook_conclusion.csv")
notebook03_conclusion
```
