---
title: "04 Battery Effect Analysis"
subtitle: "Zebrafish Sequential Behavioral Battery — Cell-wise Battery-Context Comparisons"
author: "Caique Ortiz Colombo Arnoldi Fabris"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

# Question

Do comparable endpoints differ between the full sequential battery and the alternative battery contexts within the same ethanol treatment × exposure cells?

# Why this notebook exists

Notebook 03 tests ethanol effects on endpoint-level outcomes in the BT1 full-battery context. This notebook asks a different methodological question: whether the testing context itself changes measured endpoints.

The analysis is cell-wise. Each comparison is made within the same treatment and exposure combination, so battery-context effects are not estimated by pooling across ethanol concentrations or exposure durations.

This notebook does not run PCA and does not reinterpret endpoint-level ethanol effects. It produces battery-context outputs for Notebook 06, reporting notebooks, and methods text.

# What assumptions matter

This notebook assumes that Notebook 01 produced audited datasets:

- `data_processed/bt1_audited.csv`
- `data_processed/bt2_audited.csv`
- `data_processed/bt3_audited.csv`

It also uses the endpoint registry from Notebook 02 when available:

- `outputs/logs/02_endpoint_registry.csv`
- `outputs/logs/02_model_decision_matrix.csv`

The project does not use a primary/secondary endpoint hierarchy. Endpoints are selected for battery comparison through the registry role `battery_comparison`, plus availability and protocol comparability.

Non-significant difference tests are not interpreted as evidence of sameness. A result is classified as `Inconclusive` unless a difference is detected or equivalence is formally supported by a predefined SESOI.

---

# 04.1 Load audited datasets and registry

```{r project-root}
find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = FALSE)
  candidates <- c(current, dirname(current))
  previous <- ""

  while (!identical(tail(candidates, 1), previous)) {
    previous <- tail(candidates, 1)
    parent <- dirname(previous)
    if (!identical(parent, previous)) {
      candidates <- c(candidates, parent)
    } else {
      break
    }
  }

  for (candidate in unique(candidates)) {
    has_rproj <- length(list.files(candidate, pattern = "\\.Rproj$", full.names = TRUE)) > 0
    has_git <- dir.exists(file.path(candidate, ".git"))
    has_data_dirs <- dir.exists(file.path(candidate, "data_raw")) && dir.exists(file.path(candidate, "data_processed"))
    if (has_rproj || has_git || has_data_dirs) return(candidate)
  }

  current
}

PROJECT_ROOT <- find_project_root()

paths <- list(
  project_root = PROJECT_ROOT,
  data_processed = file.path(PROJECT_ROOT, "data_processed"),
  outputs = file.path(PROJECT_ROOT, "outputs"),
  outputs_logs = file.path(PROJECT_ROOT, "outputs", "logs"),
  outputs_tables = file.path(PROJECT_ROOT, "outputs", "tables"),
  outputs_figures = file.path(PROJECT_ROOT, "outputs", "figures"),
  outputs_models = file.path(PROJECT_ROOT, "outputs", "models"),
  docs = file.path(PROJECT_ROOT, "docs")
)

for (path in paths) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE, showWarnings = FALSE)
}

paths
```

```{r packages-and-helpers}
required_packages <- c("dplyr", "tidyr", "purrr", "knitr")

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Required package is not installed: ", pkg)
  }
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
if (has_ggplot2) suppressPackageStartupMessages(library(ggplot2))

ALPHA <- 0.05
BOOTSTRAP_R <- 2000
BOOTSTRAP_SEED <- 20260504
MIN_CELL_N <- 2

treatment_levels <- c("0", "0.5", "1")
exposure_levels <- c("1h", "24h", "96h")
battery_levels <- c("BT1", "BT2", "BT3")

read_required_csv <- function(path) {
  if (!file.exists(path)) stop("Required file not found: ", path)
  read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
}

read_optional_csv <- function(path) {
  if (!file.exists(path)) return(NULL)
  read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
}

read_first_existing_csv <- function(paths) {
  existing <- paths[file.exists(paths)]
  if (length(existing) == 0) return(NULL)
  read.csv(existing[1], stringsAsFactors = FALSE, check.names = FALSE)
}

write_table <- function(df, filename, folder = paths$outputs_tables) {
  path <- file.path(folder, filename)
  write.csv(df, path, row.names = FALSE)
  path
}

write_log <- function(df, filename) {
  path <- file.path(paths$outputs_logs, filename)
  write.csv(df, path, row.names = FALSE)
  path
}

save_object <- function(object, filename) {
  path <- file.path(paths$outputs_models, filename)
  saveRDS(object, path)
  path
}

first_existing_name <- function(names_vector, candidates) {
  hit <- candidates[candidates %in% names_vector]
  if (length(hit) == 0) return(NA_character_)
  hit[1]
}

as_bool <- function(x) {
  if (is.logical(x)) return(x)
  if (is.numeric(x)) return(!is.na(x) & x != 0)
  tolower(trimws(as.character(x))) %in% c("true", "t", "1", "yes", "y")
}

as_numeric_clean <- function(x) {
  if (is.numeric(x)) return(x)
  suppressWarnings(as.numeric(as.character(x)))
}

as_binary_numeric <- function(x) {
  if (is.numeric(x)) return(ifelse(is.na(x), NA_real_, ifelse(x > 0, 1, 0)))
  y <- tolower(trimws(as.character(x)))
  dplyr::case_when(
    y %in% c("1", "true", "t", "yes", "y", "bright", "claro", "c") ~ 1,
    y %in% c("0", "false", "f", "no", "n", "dark", "escuro", "e") ~ 0,
    TRUE ~ NA_real_
  )
}

normalise_treatment <- function(x) {
  y <- trimws(as.character(x))
  y <- gsub(",", ".", y)
  y_lower <- tolower(y)
  y_lower <- gsub("%", "", y_lower)
  y_lower <- gsub("ethanol|etoh|dose|treatment|mg/l|mm|mol/l|molar| ", "", y_lower)
  out <- dplyr::case_when(
    grepl("control|ctrl", y_lower) ~ "0",
    grepl("^0(\\.0+)?$", y_lower) ~ "0",
    grepl("^0\\.5", y_lower) ~ "0.5",
    grepl("^1(\\.0+)?$", y_lower) ~ "1",
    TRUE ~ y
  )
  factor(out, levels = treatment_levels)
}

normalise_exposure <- function(x) {
  y <- tolower(trimws(as.character(x)))
  y <- gsub(" ", "", y)
  y <- gsub("hours|hour|hrs|hr", "h", y)
  out <- dplyr::case_when(
    grepl("96", y) ~ "96h",
    grepl("24", y) ~ "24h",
    grepl("^1h$|^1$|^01h$", y) ~ "1h",
    TRUE ~ y
  )
  factor(out, levels = exposure_levels)
}

standardise_design_columns <- function(df, battery_label) {
  nms <- names(df)
  id_col <- first_existing_name(nms, c("fish_id", "fishid", "fish", "animal_id", "specimen_id", "id"))
  treatment_col <- first_existing_name(nms, c("treatment", "ethanol", "ethanol_treatment", "dose", "concentration", "treat"))
  exposure_col <- first_existing_name(nms, c("exposure", "exposure_duration", "duration", "timepoint", "time_point"))

  if (is.na(treatment_col)) stop("No treatment column detected in ", battery_label)
  if (is.na(exposure_col)) stop("No exposure column detected in ", battery_label)

  df$fish_id <- if (!is.na(id_col)) as.character(df[[id_col]]) else paste0(battery_label, "_row_", seq_len(nrow(df)))
  df$treatment <- normalise_treatment(df[[treatment_col]])
  df$exposure <- normalise_exposure(df[[exposure_col]])
  df$battery <- factor(battery_label, levels = battery_levels)
  df
}

apply_global_exclusions <- function(df) {
  if ("global_exclusion_flag" %in% names(df)) {
    df <- df[!as_bool(df$global_exclusion_flag), , drop = FALSE]
  }
  df
}

clean_nonmissing_count <- function(x) {
  sum(!is.na(x) & trimws(as.character(x)) != "")
}
```

```{r load-data}
bt1 <- read_required_csv(file.path(paths$data_processed, "bt1_audited.csv"))
bt2 <- read_required_csv(file.path(paths$data_processed, "bt2_audited.csv"))
bt3 <- read_required_csv(file.path(paths$data_processed, "bt3_audited.csv"))

bt1 <- standardise_design_columns(bt1, "BT1")
bt2 <- standardise_design_columns(bt2, "BT2")
bt3 <- standardise_design_columns(bt3, "BT3")

bt1 <- apply_global_exclusions(bt1)
bt2 <- apply_global_exclusions(bt2)
bt3 <- apply_global_exclusions(bt3)

battery_data <- list(BT1 = bt1, BT2 = bt2, BT3 = bt3)

endpoint_registry <- read_first_existing_csv(c(
  file.path(paths$outputs_logs, "02_endpoint_registry.csv"),
  file.path(paths$outputs_logs, "assumptions_endpoint_registry.csv"),
  file.path(paths$outputs_logs, "00_endpoint_registry.csv")
))

model_decision_matrix <- read_first_existing_csv(c(
  file.path(paths$outputs_logs, "02_model_decision_matrix.csv"),
  file.path(paths$outputs_logs, "assumptions_model_decision_matrix.csv")
))

if (is.null(endpoint_registry)) {
  stop("Endpoint registry not found. Run 00_project_setup.Rmd and 02_endpoint_hierarchy_and_assumptions.Rmd before Notebook 04.")
}

if (!"variable" %in% names(endpoint_registry)) {
  stop("Endpoint registry must contain a 'variable' column.")
}

battery_structure <- purrr::imap_dfr(
  battery_data,
  function(df, battery_label) {
    data.frame(
      battery = battery_label,
      n_rows = nrow(df),
      n_fish_ids = length(unique(df$fish_id)),
      n_columns = ncol(df),
      stringsAsFactors = FALSE
    )
  }
)

write_table(battery_structure, "04_battery_dataset_structure.csv")
knitr::kable(battery_structure)
```

---

# 04.2 Confirm independence across battery contexts

Fish IDs are checked before battery-context comparisons. If IDs overlap between batteries, the comparison is no longer a clean independent-sample comparison and must be interpreted differently.

```{r independence-check}
within_battery_duplicates <- purrr::imap_dfr(
  battery_data,
  function(df, battery_label) {
    df %>%
      dplyr::count(fish_id, name = "n_records") %>%
      dplyr::filter(!is.na(fish_id), n_records > 1) %>%
      dplyr::mutate(battery = battery_label) %>%
      dplyr::select(battery, fish_id, n_records)
  }
)

pairwise_id_overlap <- data.frame(
  comparison = c("BT1_vs_BT2", "BT1_vs_BT3", "BT2_vs_BT3"),
  battery_a = c("BT1", "BT1", "BT2"),
  battery_b = c("BT2", "BT3", "BT3"),
  stringsAsFactors = FALSE
) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    n_overlap_ids = length(intersect(battery_data[[battery_a]]$fish_id, battery_data[[battery_b]]$fish_id)),
    independent_sample_assumption = n_overlap_ids == 0
  ) %>%
  dplyr::ungroup()

independence_confirmation <- data.frame(
  check = c(
    "duplicated IDs within batteries",
    "BT1-vs-BT2 overlap",
    "BT1-vs-BT3 overlap",
    "BT2-vs-BT3 overlap"
  ),
  result = c(
    ifelse(nrow(within_battery_duplicates) == 0, "no duplicated IDs detected", "duplicated IDs detected"),
    ifelse(pairwise_id_overlap$n_overlap_ids[pairwise_id_overlap$comparison == "BT1_vs_BT2"] == 0, "independent", "overlap detected"),
    ifelse(pairwise_id_overlap$n_overlap_ids[pairwise_id_overlap$comparison == "BT1_vs_BT3"] == 0, "independent", "overlap detected"),
    ifelse(pairwise_id_overlap$n_overlap_ids[pairwise_id_overlap$comparison == "BT2_vs_BT3"] == 0, "independent", "overlap detected")
  ),
  stringsAsFactors = FALSE
)

write_log(within_battery_duplicates, "04_within_battery_duplicate_ids.csv")
write_log(pairwise_id_overlap, "04_pairwise_battery_id_overlap.csv")
write_log(independence_confirmation, "04_independence_confirmation.csv")

knitr::kable(pairwise_id_overlap)
knitr::kable(independence_confirmation)
```

---

# 04.3 Define valid battery pairs

The comparison map is determined by protocol meaning, not by endpoint hierarchy.

- LDT endpoints: BT1 versus BT2.
- NTT endpoints: BT1 versus BT2.
- Endurance endpoints: BT1 versus BT3.

```{r valid-battery-pairs}
valid_battery_pairs <- data.frame(
  source_test = c("LDT", "NTT", "RN", "endurance", "Endurance"),
  dataset_a = c("BT1", "BT1", "BT1", "BT1", "BT1"),
  dataset_b = c("BT2", "BT2", "BT3", "BT3", "BT3"),
  comparison = c("BT1_vs_BT2", "BT1_vs_BT2", "BT1_vs_BT3", "BT1_vs_BT3", "BT1_vs_BT3"),
  comparison_rationale = c(
    "LDT endpoints are available in BT1 and the alternative LDT+NTT context.",
    "NTT endpoints are available in BT1 and the alternative LDT+NTT context.",
    "Endurance endpoints are available in BT1 and the isolated endurance context.",
    "Endurance endpoints are available in BT1 and the isolated endurance context.",
    "Endurance endpoints are available in BT1 and the isolated endurance context."
  ),
  stringsAsFactors = FALSE
)

write_table(valid_battery_pairs, "04_valid_battery_pairs.csv")
knitr::kable(valid_battery_pairs)
```

---

# 04.4 Resolve comparable endpoints

An endpoint is compared only when all of the following are true:

1. The endpoint registry marks it as a battery-comparison candidate.
2. The source test has a valid battery pair.
3. A usable column is available in both compared batteries.
4. Both batteries have non-missing data for the endpoint.

```{r endpoint-comparability}
normalise_registry <- function(registry) {
  if (!"source_test" %in% names(registry)) registry$source_test <- NA_character_
  if (!"domain" %in% names(registry)) registry$domain <- registry$source_test
  if (!"data_type" %in% names(registry)) registry$data_type <- "review"
  if (!"battery_comparison" %in% names(registry)) registry$battery_comparison <- "review"
  if (!"notes" %in% names(registry)) registry$notes <- ""
  registry
}

endpoint_registry <- normalise_registry(endpoint_registry)

is_battery_candidate <- function(value) {
  y <- tolower(trimws(as.character(value)))
  !(is.na(y) | y %in% c("no", "usually_no", "covariate_or_qc", "design/id variables", "design", "metadata"))
}

endpoint_type_from_registry <- function(data_type, variable) {
  dt <- tolower(as.character(data_type))
  dplyr::case_when(
    grepl("binary", dt) | variable == "first_choice_binary" ~ "binary",
    grepl("count|ordinal", dt) | variable %in% c("num_changes", "attempts", "last_flux") ~ "count_or_discrete",
    grepl("bounded|proportion", dt) | variable %in% c("time_bright_prop", "stratum_pref") ~ "bounded_continuous_or_proportion",
    TRUE ~ "continuous_or_derived"
  )
}

resolve_endpoint_column <- function(df, variable, battery_label = NA_character_) {
  aliases <- list(
    resistance_index = c("resistance_index", "irn_computed", "IRN", "irn"),
    condition_factor_K = c("condition_factor_K", "Kc", "Kf", "kf", "condition_factor", "fulton_k"),
    time_bright = c("time_bright", "bright_time", "time_light", "ldt_time_bright"),
    first_choice_binary = c("first_choice_binary", "first_choice_valid_binary", "first_choice"),
    stratum_pref = c("stratum_pref", "upper_stratum_pref", "top_pref", "time_top_prop"),
    vel_mean = c("vel_mean", "velocity_mean", "mean_velocity"),
    mov_total = c("mov_total", "movement_total", "total_movement_time"),
    dist_total = c("dist_total", "distance_total", "total_distance"),
    mtpc = c("mtpc", "mean_time_per_choice", "mean_time_per_entry"),
    blood_sugar = c("blood_sugar", "glycemia", "glucose")
  )

  candidates <- if (variable %in% names(aliases)) aliases[[variable]] else variable
  hit <- candidates[candidates %in% names(df)]
  if (length(hit) == 0) return(NA_character_)
  hit[1]
}

endpoint_available_in <- function(variable, battery_label) {
  df <- battery_data[[battery_label]]
  if (is.null(df)) return(list(column = NA_character_, available = FALSE, n_nonmissing = 0))
  column <- resolve_endpoint_column(df, variable, battery_label)
  if (is.na(column)) return(list(column = NA_character_, available = FALSE, n_nonmissing = 0))
  n_nonmissing <- clean_nonmissing_count(df[[column]])
  list(column = column, available = n_nonmissing > 0, n_nonmissing = n_nonmissing)
}

registry_for_battery <- endpoint_registry %>%
  dplyr::filter(is_battery_candidate(battery_comparison)) %>%
  dplyr::mutate(
    source_test_clean = dplyr::case_when(
      source_test %in% c("RN", "endurance", "Endurance") ~ source_test,
      TRUE ~ as.character(source_test)
    )
  ) %>%
  dplyr::left_join(valid_battery_pairs, by = c("source_test_clean" = "source_test"))

comparison_map <- registry_for_battery %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    endpoint_type = endpoint_type_from_registry(data_type, variable),
    resolved_column_a = endpoint_available_in(variable, dataset_a)$column,
    resolved_column_b = endpoint_available_in(variable, dataset_b)$column,
    n_nonmissing_a = endpoint_available_in(variable, dataset_a)$n_nonmissing,
    n_nonmissing_b = endpoint_available_in(variable, dataset_b)$n_nonmissing,
    present_in_dataset_a = endpoint_available_in(variable, dataset_a)$available,
    present_in_dataset_b = endpoint_available_in(variable, dataset_b)$available,
    comparable = !is.na(dataset_a) & !is.na(dataset_b) & present_in_dataset_a & present_in_dataset_b,
    comparability_decision = dplyr::case_when(
      is.na(dataset_b) ~ "excluded: no valid battery pair for source_test",
      !present_in_dataset_a ~ paste0("excluded: missing or empty in ", dataset_a),
      !present_in_dataset_b ~ paste0("excluded: missing or empty in ", dataset_b),
      TRUE ~ "retained: endpoint present in both compared batteries"
    )
  ) %>%
  dplyr::ungroup()

comparable_endpoints <- comparison_map %>%
  dplyr::filter(comparable) %>%
  dplyr::select(
    variable,
    domain,
    source_test,
    data_type,
    endpoint_type,
    comparison,
    dataset_a,
    dataset_b,
    resolved_column_a,
    resolved_column_b,
    n_nonmissing_a,
    n_nonmissing_b,
    notes
  )

endpoint_scope_audit <- comparison_map %>%
  dplyr::count(comparison, source_test, comparability_decision, name = "n_endpoints") %>%
  dplyr::arrange(comparison, source_test, comparability_decision)

write_table(comparison_map, "04_comparable_endpoint_map.csv")
write_table(comparable_endpoints, "04_comparable_endpoints_retained.csv")
write_log(endpoint_scope_audit, "04_endpoint_scope_audit.csv")

knitr::kable(comparison_map)
knitr::kable(comparable_endpoints)
```

---

# 04.5 Run cell-wise difference tests

Non-binary endpoints are compared with independent-sample Mann–Whitney rank-sum tests. Binary endpoints are compared with Fisher exact tests.

```{r effect-size-and-test-functions}
hedges_g <- function(x, y) {
  x <- x[is.finite(x)]
  y <- y[is.finite(y)]
  nx <- length(x)
  ny <- length(y)
  if (nx < 2 || ny < 2) return(NA_real_)
  sx <- stats::sd(x)
  sy <- stats::sd(y)
  if (!is.finite(sx) || !is.finite(sy)) return(NA_real_)
  sp2 <- ((nx - 1) * sx^2 + (ny - 1) * sy^2) / (nx + ny - 2)
  if (!is.finite(sp2) || sp2 <= 0) return(NA_real_)
  d <- (mean(y) - mean(x)) / sqrt(sp2)
  correction <- 1 - (3 / (4 * (nx + ny) - 9))
  d * correction
}

cliffs_delta <- function(x, y) {
  x <- x[is.finite(x)]
  y <- y[is.finite(y)]
  nx <- length(x)
  ny <- length(y)
  if (nx == 0 || ny == 0) return(NA_real_)
  comparisons <- outer(y, x, "-")
  (sum(comparisons > 0) - sum(comparisons < 0)) / (nx * ny)
}

safe_wilcox_p <- function(x, y) {
  x <- x[is.finite(x)]
  y <- y[is.finite(y)]
  if (length(x) < 1 || length(y) < 1) return(NA_real_)
  if (length(unique(c(x, y))) < 2) return(NA_real_)
  result <- tryCatch(
    stats::wilcox.test(x, y, exact = FALSE, correct = FALSE),
    error = function(e) NULL
  )
  if (is.null(result)) return(NA_real_)
  result$p.value
}

safe_fisher_binary <- function(x, y) {
  x <- as_binary_numeric(x)
  y <- as_binary_numeric(y)
  x <- x[is.finite(x)]
  y <- y[is.finite(y)]

  if (length(x) == 0 || length(y) == 0) {
    return(list(p_value = NA_real_, risk_difference = NA_real_, odds_ratio_b_vs_a = NA_real_))
  }

  tab <- matrix(
    c(sum(x == 1), sum(x == 0), sum(y == 1), sum(y == 0)),
    nrow = 2,
    byrow = TRUE
  )

  fisher_p <- tryCatch(
    stats::fisher.test(tab)$p.value,
    error = function(e) NA_real_
  )

  risk_difference <- mean(y == 1) - mean(x == 1)
  odds_a <- (sum(x == 1) + 0.5) / (sum(x == 0) + 0.5)
  odds_b <- (sum(y == 1) + 0.5) / (sum(y == 0) + 0.5)

  list(
    p_value = fisher_p,
    risk_difference = risk_difference,
    odds_ratio_b_vs_a = odds_b / odds_a
  )
}

get_cell_values <- function(variable, endpoint_type, dataset_a, dataset_b, column_a, column_b, treatment_i, exposure_i) {
  df_a <- battery_data[[dataset_a]]
  df_b <- battery_data[[dataset_b]]

  idx_a <- df_a$treatment == treatment_i & df_a$exposure == exposure_i
  idx_b <- df_b$treatment == treatment_i & df_b$exposure == exposure_i

  raw_x <- df_a[[column_a]][idx_a]
  raw_y <- df_b[[column_b]][idx_b]

  if (endpoint_type == "binary") {
    x <- as_binary_numeric(raw_x)
    y <- as_binary_numeric(raw_y)
  } else {
    x <- as_numeric_clean(raw_x)
    y <- as_numeric_clean(raw_y)
  }

  list(
    x = x[is.finite(x)],
    y = y[is.finite(y)]
  )
}

summarise_cell <- function(variable, domain, source_test, data_type, endpoint_type, comparison, dataset_a, dataset_b, resolved_column_a, resolved_column_b, treatment_i, exposure_i) {
  values <- get_cell_values(variable, endpoint_type, dataset_a, dataset_b, resolved_column_a, resolved_column_b, treatment_i, exposure_i)
  x <- values$x
  y <- values$y

  n_a <- length(x)
  n_b <- length(y)

  base <- data.frame(
    endpoint = variable,
    domain = domain,
    source_test = source_test,
    data_type = data_type,
    endpoint_type = endpoint_type,
    comparison = comparison,
    dataset_a = dataset_a,
    dataset_b = dataset_b,
    resolved_column_a = resolved_column_a,
    resolved_column_b = resolved_column_b,
    treatment = as.character(treatment_i),
    exposure = as.character(exposure_i),
    n_a = n_a,
    n_b = n_b,
    mean_a = ifelse(n_a > 0, mean(x), NA_real_),
    mean_b = ifelse(n_b > 0, mean(y), NA_real_),
    sd_a = ifelse(n_a > 1, stats::sd(x), NA_real_),
    sd_b = ifelse(n_b > 1, stats::sd(y), NA_real_),
    median_a = ifelse(n_a > 0, stats::median(x), NA_real_),
    median_b = ifelse(n_b > 0, stats::median(y), NA_real_),
    mean_difference_b_minus_a = ifelse(n_a > 0 & n_b > 0, mean(y) - mean(x), NA_real_),
    median_difference_b_minus_a = ifelse(n_a > 0 & n_b > 0, stats::median(y) - stats::median(x), NA_real_),
    stringsAsFactors = FALSE
  )

  if (endpoint_type == "binary") {
    binary_result <- safe_fisher_binary(x, y)
    base$test_method <- "Fisher exact test"
    base$p_value_raw <- binary_result$p_value
    base$hedges_g <- NA_real_
    base$rank_biserial <- NA_real_
    base$cliffs_delta <- NA_real_
    base$risk_difference_b_minus_a <- binary_result$risk_difference
    base$odds_ratio_b_vs_a <- binary_result$odds_ratio_b_vs_a
    base$feasibility_note <- ifelse(n_a == 0 | n_b == 0, "not tested: empty cell in at least one battery", "tested")
  } else {
    base$test_method <- "Mann-Whitney rank-sum"
    base$p_value_raw <- safe_wilcox_p(x, y)
    base$hedges_g <- hedges_g(x, y)
    base$rank_biserial <- cliffs_delta(x, y)
    base$cliffs_delta <- cliffs_delta(x, y)
    base$risk_difference_b_minus_a <- NA_real_
    base$odds_ratio_b_vs_a <- NA_real_
    base$feasibility_note <- dplyr::case_when(
      n_a == 0 | n_b == 0 ~ "not tested: empty cell in at least one battery",
      length(unique(c(x, y))) < 2 ~ "not tested: no variation across compared values",
      TRUE ~ "tested"
    )
  }

  base
}
```

```{r run-cellwise-comparisons}
cell_grid <- tidyr::expand_grid(
  treatment = treatment_levels,
  exposure = exposure_levels
)

if (nrow(comparable_endpoints) == 0) {
  cellwise_tests <- data.frame()
} else {
  cellwise_tests <- purrr::pmap_dfr(
    comparable_endpoints,
    function(variable, domain, source_test, data_type, endpoint_type, comparison, dataset_a, dataset_b, resolved_column_a, resolved_column_b, n_nonmissing_a, n_nonmissing_b, notes) {
      purrr::pmap_dfr(
        cell_grid,
        function(treatment, exposure) {
          summarise_cell(
            variable = variable,
            domain = domain,
            source_test = source_test,
            data_type = data_type,
            endpoint_type = endpoint_type,
            comparison = comparison,
            dataset_a = dataset_a,
            dataset_b = dataset_b,
            resolved_column_a = resolved_column_a,
            resolved_column_b = resolved_column_b,
            treatment_i = treatment,
            exposure_i = exposure
          )
        }
      )
    }
  )
}

rank_test_table <- if (nrow(cellwise_tests) > 0) cellwise_tests[cellwise_tests$test_method == "Mann-Whitney rank-sum", ] else data.frame()
binary_test_table <- if (nrow(cellwise_tests) > 0) cellwise_tests[cellwise_tests$test_method == "Fisher exact test", ] else data.frame()

write_table(cellwise_tests, "04_battery_cellwise_difference_tests_raw.csv")
write_table(rank_test_table, "04_battery_mann_whitney_tests_raw.csv")
write_table(binary_test_table, "04_battery_binary_fisher_tests_raw.csv")

knitr::kable(cellwise_tests)
```

---

# 04.6 Effect-size table

Effect sizes use `dataset_b - BT1` direction. Positive values indicate larger values in the alternative battery context than in BT1 within the same treatment × exposure cell.

```{r effect-size-table}
effect_size_table <- if (nrow(cellwise_tests) > 0) {
  cellwise_tests %>%
    dplyr::select(
      endpoint,
      domain,
      source_test,
      data_type,
      endpoint_type,
      comparison,
      dataset_a,
      dataset_b,
      treatment,
      exposure,
      n_a,
      n_b,
      mean_difference_b_minus_a,
      median_difference_b_minus_a,
      hedges_g,
      rank_biserial,
      cliffs_delta,
      risk_difference_b_minus_a,
      odds_ratio_b_vs_a,
      feasibility_note
    )
} else {
  data.frame()
}

write_table(effect_size_table, "04_battery_effect_sizes.csv")
knitr::kable(effect_size_table)
```

---

# 04.7 Bootstrap uncertainty

Bootstrap confidence intervals are generated within each treatment × exposure × battery-context cell.

```{r bootstrap-functions}
bootstrap_ci <- function(x, y, statistic_function, R = BOOTSTRAP_R) {
  x <- x[is.finite(x)]
  y <- y[is.finite(y)]

  if (length(x) == 0 || length(y) == 0) {
    return(c(estimate = NA_real_, ci_low = NA_real_, ci_high = NA_real_, n_boot_valid = 0))
  }

  estimate <- statistic_function(x, y)

  values <- replicate(
    R,
    {
      xb <- sample(x, size = length(x), replace = TRUE)
      yb <- sample(y, size = length(y), replace = TRUE)
      statistic_function(xb, yb)
    }
  )

  values <- values[is.finite(values)]
  if (length(values) < 20) {
    return(c(estimate = estimate, ci_low = NA_real_, ci_high = NA_real_, n_boot_valid = length(values)))
  }

  qs <- stats::quantile(values, probs = c(0.025, 0.975), na.rm = TRUE, names = FALSE)
  c(estimate = estimate, ci_low = qs[1], ci_high = qs[2], n_boot_valid = length(values))
}

bootstrap_for_cell <- function(variable, domain, source_test, data_type, endpoint_type, comparison, dataset_a, dataset_b, resolved_column_a, resolved_column_b, treatment_i, exposure_i) {
  values <- get_cell_values(variable, endpoint_type, dataset_a, dataset_b, resolved_column_a, resolved_column_b, treatment_i, exposure_i)
  x <- values$x
  y <- values$y

  if (endpoint_type == "binary") {
    statistics <- list(
      risk_difference_b_minus_a = function(a, b) mean(b > 0) - mean(a > 0)
    )
  } else {
    statistics <- list(
      mean_difference_b_minus_a = function(a, b) mean(b) - mean(a),
      median_difference_b_minus_a = function(a, b) stats::median(b) - stats::median(a),
      hedges_g = hedges_g,
      cliffs_delta = cliffs_delta
    )
  }

  purrr::imap_dfr(
    statistics,
    function(statistic_function, statistic_name) {
      ci <- bootstrap_ci(x, y, statistic_function, R = BOOTSTRAP_R)
      data.frame(
        endpoint = variable,
        domain = domain,
        source_test = source_test,
        data_type = data_type,
        endpoint_type = endpoint_type,
        comparison = comparison,
        dataset_a = dataset_a,
        dataset_b = dataset_b,
        treatment = as.character(treatment_i),
        exposure = as.character(exposure_i),
        statistic = statistic_name,
        estimate = ci["estimate"],
        ci_low = ci["ci_low"],
        ci_high = ci["ci_high"],
        bootstrap_R_requested = BOOTSTRAP_R,
        bootstrap_R_valid = ci["n_boot_valid"],
        stringsAsFactors = FALSE
      )
    }
  )
}
```

```{r run-bootstrap-cis}
set.seed(BOOTSTRAP_SEED)

if (nrow(comparable_endpoints) == 0) {
  bootstrap_ci_table <- data.frame()
} else {
  bootstrap_ci_table <- purrr::pmap_dfr(
    comparable_endpoints,
    function(variable, domain, source_test, data_type, endpoint_type, comparison, dataset_a, dataset_b, resolved_column_a, resolved_column_b, n_nonmissing_a, n_nonmissing_b, notes) {
      purrr::pmap_dfr(
        cell_grid,
        function(treatment, exposure) {
          bootstrap_for_cell(
            variable = variable,
            domain = domain,
            source_test = source_test,
            data_type = data_type,
            endpoint_type = endpoint_type,
            comparison = comparison,
            dataset_a = dataset_a,
            dataset_b = dataset_b,
            resolved_column_a = resolved_column_a,
            resolved_column_b = resolved_column_b,
            treatment_i = treatment,
            exposure_i = exposure
          )
        }
      )
    }
  )
}

write_table(bootstrap_ci_table, "04_battery_bootstrap_confidence_intervals.csv")
knitr::kable(bootstrap_ci_table)
```

---

# 04.8 Equivalence setup

Equivalence testing is run only if a predefined SESOI file is supplied. Without SESOI bounds, non-significant difference tests remain inconclusive.

Expected optional SESOI file columns:

- `endpoint`
- `sesoi_raw`
- `sesoi_unit`
- `sesoi_source`
- `equivalence_test_allowed`

```{r sesoi-setup}
sesoi_input <- read_first_existing_csv(c(
  file.path(paths$outputs_logs, "battery_effect_sesoi.csv"),
  file.path(paths$outputs_tables, "battery_effect_sesoi.csv"),
  file.path(paths$docs, "battery_effect_sesoi.csv")
))

if (nrow(comparable_endpoints) == 0) {
  sesoi_table <- data.frame()
} else if (is.null(sesoi_input)) {
  sesoi_table <- comparable_endpoints %>%
    dplyr::distinct(endpoint = variable, domain, source_test, endpoint_type) %>%
    dplyr::mutate(
      sesoi_raw = NA_real_,
      sesoi_unit = "raw endpoint units",
      sesoi_source = "not defined; confirmatory TOST not run",
      equivalence_test_allowed = FALSE
    )

  write_log(sesoi_table, "04_battery_effect_sesoi_template.csv")
} else {
  if (!all(c("endpoint", "sesoi_raw") %in% names(sesoi_input))) {
    stop("SESOI file exists but lacks required columns: endpoint, sesoi_raw")
  }

  if (!"sesoi_unit" %in% names(sesoi_input)) sesoi_input$sesoi_unit <- "raw endpoint units"
  if (!"sesoi_source" %in% names(sesoi_input)) sesoi_input$sesoi_source <- "source not specified"
  if (!"equivalence_test_allowed" %in% names(sesoi_input)) {
    sesoi_input$equivalence_test_allowed <- is.finite(sesoi_input$sesoi_raw)
  } else {
    sesoi_input$equivalence_test_allowed <- as_bool(sesoi_input$equivalence_test_allowed)
  }

  sesoi_table <- comparable_endpoints %>%
    dplyr::distinct(endpoint = variable, domain, source_test, endpoint_type) %>%
    dplyr::left_join(sesoi_input, by = "endpoint") %>%
    dplyr::mutate(
      sesoi_unit = ifelse(is.na(sesoi_unit), "raw endpoint units", sesoi_unit),
      sesoi_source = ifelse(is.na(sesoi_source), "source not specified", sesoi_source),
      equivalence_test_allowed = ifelse(is.na(equivalence_test_allowed), FALSE, equivalence_test_allowed),
      equivalence_test_allowed = equivalence_test_allowed & is.finite(sesoi_raw) & sesoi_raw > 0
    )
}

sesoi_log <- if (nrow(sesoi_table) > 0) {
  sesoi_table %>%
    dplyr::mutate(
      equivalence_decision = dplyr::case_when(
        endpoint_type == "binary" ~ "not run here: binary endpoint requires endpoint-specific equivalence definition",
        equivalence_test_allowed ~ "TOST allowed",
        TRUE ~ "TOST not run: SESOI absent or not allowed"
      )
    )
} else {
  data.frame()
}

write_log(sesoi_log, "04_equivalence_bound_log.csv")
knitr::kable(sesoi_log)
```

---

# 04.9 TOST equivalence, when allowed

```{r tost-functions}
welch_tost <- function(x, y, sesoi_raw, alpha = ALPHA) {
  x <- x[is.finite(x)]
  y <- y[is.finite(y)]
  nx <- length(x)
  ny <- length(y)

  if (nx < 2 || ny < 2 || !is.finite(sesoi_raw) || sesoi_raw <= 0) {
    return(data.frame(
      tost_p_lower = NA_real_,
      tost_p_upper = NA_real_,
      tost_p_max = NA_real_,
      tost_equivalent = FALSE,
      tost_note = "not tested: insufficient data or invalid SESOI",
      stringsAsFactors = FALSE
    ))
  }

  mean_diff <- mean(y) - mean(x)
  vx <- stats::var(x)
  vy <- stats::var(y)
  se <- sqrt(vx / nx + vy / ny)

  if (!is.finite(se) || se == 0) {
    return(data.frame(
      tost_p_lower = NA_real_,
      tost_p_upper = NA_real_,
      tost_p_max = NA_real_,
      tost_equivalent = FALSE,
      tost_note = "not tested: zero or undefined standard error",
      stringsAsFactors = FALSE
    ))
  }

  df_num <- (vx / nx + vy / ny)^2
  df_den <- ((vx / nx)^2 / (nx - 1)) + ((vy / ny)^2 / (ny - 1))
  df <- df_num / df_den

  t_lower <- (mean_diff + sesoi_raw) / se
  t_upper <- (mean_diff - sesoi_raw) / se

  p_lower <- 1 - stats::pt(t_lower, df = df)
  p_upper <- stats::pt(t_upper, df = df)
  p_max <- max(p_lower, p_upper)

  data.frame(
    tost_p_lower = p_lower,
    tost_p_upper = p_upper,
    tost_p_max = p_max,
    tost_equivalent = p_lower < alpha & p_upper < alpha,
    tost_note = "tested",
    stringsAsFactors = FALSE
  )
}

run_tost_for_cell <- function(variable, domain, source_test, data_type, endpoint_type, comparison, dataset_a, dataset_b, resolved_column_a, resolved_column_b, treatment_i, exposure_i, sesoi_raw, equivalence_test_allowed) {
  if (endpoint_type == "binary") {
    return(data.frame(
      endpoint = variable,
      domain = domain,
      source_test = source_test,
      data_type = data_type,
      endpoint_type = endpoint_type,
      comparison = comparison,
      dataset_a = dataset_a,
      dataset_b = dataset_b,
      treatment = as.character(treatment_i),
      exposure = as.character(exposure_i),
      sesoi_raw = sesoi_raw,
      tost_p_lower = NA_real_,
      tost_p_upper = NA_real_,
      tost_p_max = NA_real_,
      tost_equivalent = FALSE,
      tost_note = "not tested: binary endpoint requires endpoint-specific equivalence definition",
      stringsAsFactors = FALSE
    ))
  }

  if (!equivalence_test_allowed) {
    return(data.frame(
      endpoint = variable,
      domain = domain,
      source_test = source_test,
      data_type = data_type,
      endpoint_type = endpoint_type,
      comparison = comparison,
      dataset_a = dataset_a,
      dataset_b = dataset_b,
      treatment = as.character(treatment_i),
      exposure = as.character(exposure_i),
      sesoi_raw = sesoi_raw,
      tost_p_lower = NA_real_,
      tost_p_upper = NA_real_,
      tost_p_max = NA_real_,
      tost_equivalent = FALSE,
      tost_note = "not tested: SESOI absent or equivalence testing not allowed",
      stringsAsFactors = FALSE
    ))
  }

  values <- get_cell_values(variable, endpoint_type, dataset_a, dataset_b, resolved_column_a, resolved_column_b, treatment_i, exposure_i)
  tost <- welch_tost(values$x, values$y, sesoi_raw = sesoi_raw, alpha = ALPHA)

  cbind(
    data.frame(
      endpoint = variable,
      domain = domain,
      source_test = source_test,
      data_type = data_type,
      endpoint_type = endpoint_type,
      comparison = comparison,
      dataset_a = dataset_a,
      dataset_b = dataset_b,
      treatment = as.character(treatment_i),
      exposure = as.character(exposure_i),
      sesoi_raw = sesoi_raw,
      stringsAsFactors = FALSE
    ),
    tost
  )
}
```

```{r run-tost}
if (nrow(comparable_endpoints) == 0) {
  equivalence_tests <- data.frame()
} else {
  tost_input <- comparable_endpoints %>%
    dplyr::rename(endpoint = variable) %>%
    dplyr::left_join(
      sesoi_table %>% dplyr::select(endpoint, sesoi_raw, equivalence_test_allowed),
      by = "endpoint"
    ) %>%
    dplyr::mutate(
      sesoi_raw = ifelse(is.na(sesoi_raw), NA_real_, sesoi_raw),
      equivalence_test_allowed = ifelse(is.na(equivalence_test_allowed), FALSE, equivalence_test_allowed)
    )

  equivalence_tests <- purrr::pmap_dfr(
    tost_input,
    function(endpoint, domain, source_test, data_type, endpoint_type, comparison, dataset_a, dataset_b, resolved_column_a, resolved_column_b, n_nonmissing_a, n_nonmissing_b, notes, sesoi_raw, equivalence_test_allowed) {
      purrr::pmap_dfr(
        cell_grid,
        function(treatment, exposure) {
          run_tost_for_cell(
            variable = endpoint,
            domain = domain,
            source_test = source_test,
            data_type = data_type,
            endpoint_type = endpoint_type,
            comparison = comparison,
            dataset_a = dataset_a,
            dataset_b = dataset_b,
            resolved_column_a = resolved_column_a,
            resolved_column_b = resolved_column_b,
            treatment_i = treatment,
            exposure_i = exposure,
            sesoi_raw = sesoi_raw,
            equivalence_test_allowed = equivalence_test_allowed
          )
        }
      )
    }
  )
}

write_table(equivalence_tests, "04_battery_tost_equivalence_tests.csv")
knitr::kable(equivalence_tests)
```

---

# 04.10 BH-FDR correction and classification

Difference-test p-values are corrected across the battery-comparison testing family. Classification is rule-based:

- `Difference detected`: BH-FDR adjusted difference test is below alpha.
- `Equivalence supported`: no detected difference, but TOST supports practical equivalence using predefined SESOI.
- `Inconclusive`: neither difference nor equivalence is established.

```{r fdr-and-classification}
adjusted_difference_tests <- cellwise_tests

if (nrow(adjusted_difference_tests) > 0) {
  adjusted_difference_tests$p_value_bh <- NA_real_
  valid_p <- is.finite(adjusted_difference_tests$p_value_raw)
  adjusted_difference_tests$p_value_bh[valid_p] <- stats::p.adjust(adjusted_difference_tests$p_value_raw[valid_p], method = "BH")
  adjusted_difference_tests$difference_detected <- is.finite(adjusted_difference_tests$p_value_bh) & adjusted_difference_tests$p_value_bh < ALPHA
} else {
  adjusted_difference_tests <- data.frame()
}

if (nrow(adjusted_difference_tests) > 0) {
  battery_method_summary <- adjusted_difference_tests %>%
    dplyr::left_join(
      equivalence_tests %>%
        dplyr::select(endpoint, comparison, treatment, exposure, sesoi_raw, tost_p_max, tost_equivalent, tost_note),
      by = c("endpoint", "comparison", "treatment", "exposure")
    ) %>%
    dplyr::mutate(
      tost_equivalent = ifelse(is.na(tost_equivalent), FALSE, tost_equivalent),
      battery_method_conclusion = dplyr::case_when(
        difference_detected ~ "Difference detected",
        !difference_detected & tost_equivalent ~ "Equivalence supported",
        TRUE ~ "Inconclusive"
      ),
      interpretation_rule = dplyr::case_when(
        battery_method_conclusion == "Difference detected" ~ "Adjusted difference test supports a battery-context difference in this endpoint/cell.",
        battery_method_conclusion == "Equivalence supported" ~ "TOST supports practical similarity within predefined SESOI bounds.",
        TRUE ~ "No adjusted difference detected, but equivalence is not established. Do not write 'no difference'."
      )
    )
} else {
  battery_method_summary <- data.frame()
}

summary_counts <- if (nrow(battery_method_summary) > 0) {
  battery_method_summary %>%
    dplyr::count(comparison, source_test, domain, battery_method_conclusion, name = "n_endpoint_cells") %>%
    dplyr::arrange(comparison, source_test, domain, battery_method_conclusion)
} else {
  data.frame()
}

write_table(adjusted_difference_tests, "04_battery_adjusted_difference_tests.csv")
write_table(battery_method_summary, "04_battery_method_summary.csv")
write_table(summary_counts, "04_battery_method_summary_counts.csv")

knitr::kable(adjusted_difference_tests)
knitr::kable(battery_method_summary)
knitr::kable(summary_counts)
```

---

# 04.11 Diagnostic figure

```{r optional-effect-size-plot, fig.width=10, fig.height=7}
if (has_ggplot2 && nrow(effect_size_table) > 0 && "hedges_g" %in% names(effect_size_table)) {
  plot_data <- effect_size_table %>%
    dplyr::filter(is.finite(hedges_g)) %>%
    dplyr::mutate(
      cell = paste(treatment, exposure, sep = " / "),
      endpoint_cell = paste(endpoint, cell, sep = " | ")
    )

  if (nrow(plot_data) > 0) {
    p <- ggplot2::ggplot(plot_data, ggplot2::aes(x = hedges_g, y = endpoint_cell)) +
      ggplot2::geom_vline(xintercept = 0, linetype = "dashed") +
      ggplot2::geom_point() +
      ggplot2::facet_wrap(~ comparison, scales = "free_y") +
      ggplot2::labs(
        title = "Battery-context effect sizes by endpoint and treatment × exposure cell",
        x = "Hedges' g; positive = alternative battery context greater than BT1",
        y = NULL
      ) +
      ggplot2::theme_bw()

    print(p)

    ggplot2::ggsave(
      filename = file.path(paths$outputs_figures, "04_battery_hedges_g_forest_plot.png"),
      plot = p,
      width = 10,
      height = 7,
      dpi = 300
    )
  }
}
```

---

# 04.12 Notebook conclusion

```{r notebook-conclusion}
conclusion_text <- data.frame(
  conclusion_item = c(
    "Battery comparison structure",
    "Endpoint selection",
    "Difference testing",
    "Effect-size reporting",
    "Bootstrap uncertainty",
    "Equivalence rule",
    "Downstream use"
  ),
  conclusion = c(
    "BT1-vs-BT2 is used for LDT/NTT endpoints; BT1-vs-BT3 is used for endurance endpoints.",
    "Comparable endpoints are selected from the endpoint registry and availability checks; no primary/secondary hierarchy is used.",
    "Battery-context differences are tested cell-wise within treatment × exposure combinations.",
    "Non-binary endpoints report Hedges' g and Cliff's delta/rank-biserial; binary endpoints report risk difference and odds ratio.",
    "Bootstrap confidence intervals are generated for endpoint-appropriate statistics.",
    "Equivalence is not claimed unless predefined SESOI bounds permit confirmatory TOST.",
    "Saved outputs feed Notebook 06 concordance/robustness checks, Notebook 07 reporting, and Notebook 08 methods text."
  ),
  stringsAsFactors = FALSE
)

write_log(conclusion_text, "04_battery_effect_analysis_conclusion.csv")
knitr::kable(conclusion_text)

save_object(
  list(
    battery_structure = battery_structure,
    independence_confirmation = independence_confirmation,
    valid_battery_pairs = valid_battery_pairs,
    comparison_map = comparison_map,
    comparable_endpoints = comparable_endpoints,
    cellwise_tests = cellwise_tests,
    effect_size_table = effect_size_table,
    bootstrap_ci_table = bootstrap_ci_table,
    sesoi_log = sesoi_log,
    equivalence_tests = equivalence_tests,
    adjusted_difference_tests = adjusted_difference_tests,
    battery_method_summary = battery_method_summary,
    summary_counts = summary_counts
  ),
  "04_battery_effect_analysis_results.rds"
)
```
