---
title: "06 Sensitivity and Robustness"
author: "Caique Ortiz Colombo Arnoldi Fabris"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
---

# Purpose

This notebook is the concordance and robustness layer for the analysis pipeline.

It performs two tasks.

First, it checks whether the outputs generated by notebooks 00W-05 are aligned with the current analysis structure: existing repository layout, endpoint registry rather than primary/secondary hierarchy, endpoint-level ethanol models, separate battery-context comparisons, and one single BT1 full-battery PCA used as multivariate visualization.

Second, it performs robustness and stability checks that are needed before manuscript reporting: dataset availability, endpoint-registry consistency, missingness, treatment-by-exposure cell retention, condition factor validation, outlier-flag impact, and PCA stability.

This notebook does not introduce a new biological hypothesis. It does not replace endpoint-level models from Notebook 03. It does not run a competing PCA. PCA components are interpreted from their loadings.

# Notebook contract

This notebook expects the previous notebooks to have created the following layers:

- `python/00_data_wrangling.ipynb`: wrangled battery-level CSV files.
- `notebooks/00_project_setup.Rmd`: project paths, endpoint registry, and analysis rules.
- `notebooks/01_data_audit_and_qc.Rmd`: audited BT1, BT2, and BT3 datasets.
- `notebooks/02_endpoint_hierarchy_and_assumptions.Rmd`: endpoint registry and model-readiness outputs.
- `notebooks/03_ethanol_factorial_models.Rmd`: endpoint-level ethanol model outputs.
- `notebooks/04_battery_effect_analysis.Rmd`: battery-context comparison outputs.
- `notebooks/05_multivariate_pca_structure.Rmd`: one BT1 full-battery PCA and PCA visualization outputs.

The file name of Notebook 02 may remain unchanged for continuity, but the analysis logic is endpoint registry and model readiness, not primary/secondary endpoint hierarchy.

```{r setup, message=FALSE, warning=FALSE}
set.seed(20260505)

required_packages <- c(
  "dplyr", "tidyr", "readr", "stringr", "purrr", "tibble",
  "ggplot2", "broom", "rlang"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_packages) > 0) {
  stop(
    "Missing required packages: ",
    paste(missing_packages, collapse = ", "),
    ". Install them before rendering this notebook."
  )
}

library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(purrr)
library(tibble)
library(ggplot2)
library(broom)
```

# 06.1 Project paths and helper functions

```{r paths-and-helpers}
find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)

  while (TRUE) {
    has_rproj <- length(list.files(current, pattern = "\\.Rproj$", full.names = TRUE)) > 0
    has_git <- dir.exists(file.path(current, ".git"))
    has_data <- dir.exists(file.path(current, "data_raw")) && dir.exists(file.path(current, "data_processed"))

    if (has_rproj || has_git || has_data) {
      return(current)
    }

    parent <- dirname(current)
    if (identical(parent, current)) {
      return(normalizePath(start, winslash = "/", mustWork = TRUE))
    }
    current <- parent
  }
}

project_root <- find_project_root()

paths <- list(
  root = project_root,
  data_raw = file.path(project_root, "data_raw"),
  data_processed = file.path(project_root, "data_processed"),
  docs = file.path(project_root, "docs"),
  notebooks = file.path(project_root, "notebooks"),
  outputs = file.path(project_root, "outputs"),
  output_tables = file.path(project_root, "outputs", "tables"),
  output_figures = file.path(project_root, "outputs", "figures"),
  output_logs = file.path(project_root, "outputs", "logs"),
  output_models = file.path(project_root, "outputs", "models"),
  python = file.path(project_root, "python"),
  R = file.path(project_root, "R")
)

purrr::walk(paths[c("outputs", "output_tables", "output_figures", "output_logs", "output_models")], dir.create, recursive = TRUE, showWarnings = FALSE)

read_first_existing <- function(candidates, required = TRUE, label = "file") {
  existing <- candidates[file.exists(candidates)]
  if (length(existing) == 0) {
    if (required) {
      stop("No candidate found for ", label, ". Checked: ", paste(candidates, collapse = " | "))
    }
    return(NULL)
  }
  readr::read_csv(existing[[1]], show_col_types = FALSE)
}

first_existing_path <- function(candidates) {
  existing <- candidates[file.exists(candidates)]
  if (length(existing) == 0) return(NA_character_)
  existing[[1]]
}

write_table <- function(x, filename) {
  readr::write_csv(x, file.path(paths$output_tables, filename))
  invisible(x)
}

write_log <- function(x, filename) {
  readr::write_csv(x, file.path(paths$output_logs, filename))
  invisible(x)
}

safe_has_col <- function(df, col) {
  col %in% names(df)
}

coerce_numeric_if_possible <- function(x) {
  if (is.numeric(x)) return(x)
  suppressWarnings(as.numeric(x))
}

standardize_design <- function(df) {
  out <- df

  if ("treatment" %in% names(out)) {
    out <- out %>%
      mutate(
        treatment = stringr::str_trim(as.character(treatment)),
        treatment = dplyr::case_when(
          treatment %in% c("0", "0.0", "control", "Control", "CTRL", "ctrl") ~ "0",
          treatment %in% c("0.5", "0,5", "0.50", "0,50") ~ "0.5",
          treatment %in% c("1", "1.0", "1.00") ~ "1",
          TRUE ~ treatment
        ),
        treatment = factor(treatment, levels = c("0", "0.5", "1"))
      )
  }

  if ("exposure" %in% names(out)) {
    out <- out %>%
      mutate(
        exposure = stringr::str_to_lower(stringr::str_trim(as.character(exposure))),
        exposure = stringr::str_replace_all(exposure, "\\s+", ""),
        exposure = dplyr::case_when(
          exposure %in% c("1", "1h", "01", "01h") ~ "1h",
          exposure %in% c("24", "24h") ~ "24h",
          exposure %in% c("96", "96h") ~ "96h",
          TRUE ~ exposure
        ),
        exposure = factor(exposure, levels = c("1h", "24h", "96h"))
      )
  }

  if ("battery" %in% names(out)) {
    out <- out %>%
      mutate(
        battery = stringr::str_to_upper(stringr::str_trim(as.character(battery))),
        battery = dplyr::case_when(
          battery %in% c("1", "B1", "BATTERY1", "BATTERY_1", "SUITE1", "SUITE_1") ~ "BT1",
          battery %in% c("2", "B2", "BATTERY2", "BATTERY_2", "SUITE2", "SUITE_2") ~ "BT2",
          battery %in% c("3", "B3", "BATTERY3", "BATTERY_3", "SUITE3", "SUITE_3") ~ "BT3",
          TRUE ~ battery
        ),
        battery = factor(battery, levels = c("BT1", "BT2", "BT3"))
      )
  }

  out
}

classify_numeric_quality <- function(x) {
  x_num <- coerce_numeric_if_possible(x)
  n_total <- length(x)
  n_nonmissing <- sum(!is.na(x_num))
  n_unique <- length(unique(x_num[!is.na(x_num)]))

  tibble(
    n_total = n_total,
    n_nonmissing = n_nonmissing,
    pct_nonmissing = ifelse(n_total > 0, n_nonmissing / n_total, NA_real_),
    n_unique = n_unique,
    numeric_usable = n_nonmissing > 0 && n_unique > 1
  )
}

iqr_flags <- function(x, multiplier = 3) {
  x_num <- coerce_numeric_if_possible(x)
  q <- stats::quantile(x_num, probs = c(0.25, 0.75), na.rm = TRUE, names = FALSE)
  iqr <- q[2] - q[1]
  if (is.na(iqr) || iqr <= 0) {
    return(rep(FALSE, length(x_num)))
  }
  lower <- q[1] - multiplier * iqr
  upper <- q[2] + multiplier * iqr
  !is.na(x_num) & (x_num < lower | x_num > upper)
}

safe_cor <- function(x, y) {
  x <- coerce_numeric_if_possible(x)
  y <- coerce_numeric_if_possible(y)
  ok <- !is.na(x) & !is.na(y)
  if (sum(ok) < 3) return(NA_real_)
  suppressWarnings(stats::cor(x[ok], y[ok]))
}

simple_domain_label <- function(top_domains) {
  domains <- unique(top_domains[!is.na(top_domains)])
  domains <- domains[domains != ""]
  if (length(domains) == 0) return("unlabeled")
  if (length(domains) == 1) return(domains[[1]])
  paste("mixed:", paste(domains, collapse = " + "))
}

cat("Project root:", project_root, "\n")
```

# 06.2 File-level concordance audit

This section checks whether the expected repo files and output layers are present. It does not inspect biological results.

```{r file-level-concordance}
expected_repo_items <- tibble(
  item = c(
    "data_raw",
    "data_processed",
    "docs",
    "notebooks",
    "outputs",
    "python",
    "R",
    "README.md"
  ),
  expected_path = c(
    paths$data_raw,
    paths$data_processed,
    paths$docs,
    paths$notebooks,
    paths$outputs,
    paths$python,
    paths$R,
    file.path(paths$root, "README.md")
  ),
  expected_type = c("directory", "directory", "directory", "directory", "directory", "directory", "directory", "file")
) %>%
  mutate(
    exists = if_else(expected_type == "directory", dir.exists(expected_path), file.exists(expected_path)),
    audit_status = if_else(exists, "present", "missing")
  )

expected_notebooks <- tibble(
  notebook = c(
    "python/00_data_wrangling.ipynb",
    "notebooks/00_project_setup.Rmd",
    "notebooks/01_data_audit_and_qc.Rmd",
    "notebooks/02_endpoint_hierarchy_and_assumptions.Rmd",
    "notebooks/03_ethanol_factorial_models.Rmd",
    "notebooks/04_battery_effect_analysis.Rmd",
    "notebooks/05_multivariate_pca_structure.Rmd"
  ),
  expected_path = file.path(paths$root, notebook)
) %>%
  mutate(
    exists = file.exists(expected_path),
    audit_status = if_else(exists, "present", "missing")
  )

file_concordance <- bind_rows(
  expected_repo_items %>%
    transmute(layer = "repo_structure", item, expected_path, exists, audit_status),
  expected_notebooks %>%
    transmute(layer = "notebook", item = notebook, expected_path, exists, audit_status)
)

write_log(file_concordance, "06_file_concordance.csv")

file_concordance
```

# 06.3 Load analysis inputs

```{r load-analysis-inputs}
bt1 <- read_first_existing(
  c(
    file.path(paths$data_processed, "bt1_audited.csv"),
    file.path(paths$data_processed, "BT1_audited.csv"),
    file.path(paths$data_processed, "battery_suite_1_audited.csv"),
    file.path(paths$data_processed, "battery_suite_1.csv")
  ),
  required = TRUE,
  label = "BT1 audited or wrangled dataset"
) %>% standardize_design()

bt2 <- read_first_existing(
  c(
    file.path(paths$data_processed, "bt2_audited.csv"),
    file.path(paths$data_processed, "BT2_audited.csv"),
    file.path(paths$data_processed, "battery_suite_2_audited.csv"),
    file.path(paths$data_processed, "battery_suite_2.csv")
  ),
  required = TRUE,
  label = "BT2 audited or wrangled dataset"
) %>% standardize_design()

bt3 <- read_first_existing(
  c(
    file.path(paths$data_processed, "bt3_audited.csv"),
    file.path(paths$data_processed, "BT3_audited.csv"),
    file.path(paths$data_processed, "battery_suite_3_audited.csv"),
    file.path(paths$data_processed, "battery_suite_3.csv")
  ),
  required = TRUE,
  label = "BT3 audited or wrangled dataset"
) %>% standardize_design()

endpoint_registry <- read_first_existing(
  c(
    file.path(paths$output_tables, "02_endpoint_registry.csv"),
    file.path(paths$output_tables, "00_endpoint_registry.csv"),
    file.path(paths$output_logs, "02_endpoint_registry.csv"),
    file.path(paths$output_logs, "00_endpoint_registry.csv")
  ),
  required = FALSE,
  label = "endpoint registry"
)

model_readiness <- read_first_existing(
  c(
    file.path(paths$output_tables, "02_model_readiness.csv"),
    file.path(paths$output_tables, "02_model_decision_matrix.csv"),
    file.path(paths$output_logs, "02_model_readiness.csv")
  ),
  required = FALSE,
  label = "model readiness"
)

pca_loadings_existing <- read_first_existing(
  c(
    file.path(paths$output_tables, "05_pca_loadings.csv"),
    file.path(paths$output_tables, "05_pca_variable_loadings.csv"),
    file.path(paths$output_logs, "05_pca_loadings.csv")
  ),
  required = FALSE,
  label = "PCA loadings"
)

pca_scores_existing <- read_first_existing(
  c(
    file.path(paths$output_tables, "05_pca_scores.csv"),
    file.path(paths$output_tables, "05_pca_individual_scores.csv"),
    file.path(paths$output_logs, "05_pca_scores.csv")
  ),
  required = FALSE,
  label = "PCA scores"
)

dataset_inventory <- tibble(
  dataset = c("BT1", "BT2", "BT3"),
  rows = c(nrow(bt1), nrow(bt2), nrow(bt3)),
  columns = c(ncol(bt1), ncol(bt2), ncol(bt3)),
  unique_fish = c(
    if ("fish_id" %in% names(bt1)) n_distinct(bt1$fish_id) else NA_integer_,
    if ("fish_id" %in% names(bt2)) n_distinct(bt2$fish_id) else NA_integer_,
    if ("fish_id" %in% names(bt3)) n_distinct(bt3$fish_id) else NA_integer_
  )
)

write_table(dataset_inventory, "06_dataset_inventory.csv")

dataset_inventory
```

# 06.4 Endpoint registry concordance

This section verifies that the registry uses practical roles rather than primary/secondary hierarchy.

```{r endpoint-registry-concordance}
if (is.null(endpoint_registry)) {
  endpoint_registry_concordance <- tibble(
    check = "endpoint_registry_loaded",
    status = "missing",
    details = "No endpoint registry file was found. Notebook 02 should export it before Notebook 06 is rendered."
  )
} else {
  registry_cols <- names(endpoint_registry)

  forbidden_cols <- registry_cols[str_detect(str_to_lower(registry_cols), "primary|secondary|hierarchy")]

  expected_role_cols <- c(
    "endpoint", "domain", "source_test", "data_type",
    "pca_included", "model_tested", "battery_comparable", "role"
  )

  role_cols_present <- intersect(expected_role_cols, registry_cols)

  endpoint_registry_concordance <- tibble(
    check = c(
      "endpoint_registry_loaded",
      "no_primary_secondary_columns",
      "role_columns_present",
      "contains_endpoint_column"
    ),
    status = c(
      "present",
      if_else(length(forbidden_cols) == 0, "pass", "flag"),
      if_else(length(role_cols_present) > 0, "pass", "flag"),
      if_else(any(str_to_lower(registry_cols) %in% c("endpoint", "variable", "endpoint_variable")), "pass", "flag")
    ),
    details = c(
      paste0("Registry rows: ", nrow(endpoint_registry), "; columns: ", ncol(endpoint_registry)),
      if_else(length(forbidden_cols) == 0, "No primary/secondary registry columns detected.", paste("Detected:", paste(forbidden_cols, collapse = ", "))),
      paste("Detected role-like columns:", paste(role_cols_present, collapse = ", ")),
      paste("Registry columns:", paste(registry_cols, collapse = ", "))
    )
  )
}

write_log(endpoint_registry_concordance, "06_endpoint_registry_concordance.csv")

endpoint_registry_concordance
```

# 06.5 Fixed PCA endpoint-set concordance

The PCA endpoint set is fixed by the current project structure. This section resolves aliases, checks availability in BT1, and determines which variables can actually enter the single PCA.

```{r pca-endpoint-set-concordance}
pca_endpoint_plan <- tibble(
  planned_variable = c(
    "dist_total",
    "mov_total",
    "vel_mean",
    "stratum_pref",
    "time_bright",
    "first_choice_binary",
    "latency",
    "mtpc",
    "num_changes",
    "resistance_index",
    "attempts",
    "blood_sugar",
    "condition_factor"
  ),
  domain = c(
    "NTT locomotion",
    "NTT locomotion",
    "NTT locomotion",
    "NTT spatial position",
    "LDT phototaxis",
    "LDT first choice",
    "LDT decision latency",
    "LDT transition behavior",
    "LDT transition behavior",
    "Endurance",
    "Endurance",
    "Physiology",
    "Biometry/condition"
  ),
  aliases = c(
    "dist_total|distance_total|total_distance",
    "mov_total|movement_total|time_moving|moving_time",
    "vel_mean|mean_velocity|velocity_mean",
    "stratum_pref|upper_stratum_pref|top_zone_time|upper_time|perc_upper_stratum",
    "time_bright|bright_time|time_light|light_time",
    "first_choice_binary|first_choice_valid_binary|first_choice_num|first_choice_coded",
    "latency|latency_first_choice|choice_latency",
    "mtpc|mean_time_per_choice|mean_time_per_entry",
    "num_changes|number_changes|n_changes|transitions",
    "resistance_index|irn_computed|irn|resistance",
    "attempts|n_attempts|resistance_attempts",
    "blood_sugar|glycemia|glucose|blood_glucose",
    "Kc|Kf|condition_factor|fulton|fulton_k|condition"
  )
)

resolve_alias <- function(df, alias_string) {
  candidates <- str_split(alias_string, "\\|", simplify = TRUE)
  candidates <- candidates[candidates != ""]
  exact <- candidates[candidates %in% names(df)]
  if (length(exact) > 0) return(exact[[1]])

  lower_map <- tibble(actual = names(df), lower = str_to_lower(names(df)))
  lower_candidates <- str_to_lower(candidates)
  lower_match <- lower_map %>% filter(lower %in% lower_candidates)
  if (nrow(lower_match) > 0) return(lower_match$actual[[1]])

  NA_character_
}

pca_variable_concordance <- pca_endpoint_plan %>%
  mutate(
    resolved_variable = map_chr(aliases, ~ resolve_alias(bt1, .x)),
    available_in_bt1 = !is.na(resolved_variable),
    n_nonmissing = map_int(resolved_variable, ~ if (is.na(.x)) 0L else sum(!is.na(bt1[[.x]]))),
    n_unique = map_int(resolved_variable, ~ if (is.na(.x)) 0L else length(unique(bt1[[.x]][!is.na(bt1[[.x]])]))),
    numeric_usable = map_lgl(resolved_variable, ~ {
      if (is.na(.x)) return(FALSE)
      quality <- classify_numeric_quality(bt1[[.x]])
      quality$numeric_usable[[1]]
    }),
    pca_ready = available_in_bt1 & numeric_usable,
    note = case_when(
      !available_in_bt1 ~ "No matching column resolved in BT1.",
      !numeric_usable ~ "Column exists but is not numerically usable for PCA.",
      TRUE ~ "Resolved and PCA-eligible."
    )
  )

write_table(pca_variable_concordance, "06_pca_variable_concordance.csv")

pca_variable_concordance
```

# 06.6 Treatment-by-exposure retention audit

```{r treatment-exposure-retention}
design_cols_present <- all(c("treatment", "exposure") %in% names(bt1))

if (!design_cols_present) {
  cell_retention <- tibble(
    audit_status = "flag",
    details = "BT1 does not contain both treatment and exposure columns."
  )
} else {
  pca_ready_vars <- pca_variable_concordance %>%
    filter(pca_ready) %>%
    pull(resolved_variable)

  bt1_pca_retention <- bt1 %>%
    mutate(
      complete_for_pca = if (length(pca_ready_vars) == 0) FALSE else if_all(all_of(pca_ready_vars), ~ !is.na(.x))
    )

  cell_retention <- bt1_pca_retention %>%
    group_by(treatment, exposure) %>%
    summarise(
      n_total = n(),
      n_complete_pca = sum(complete_for_pca),
      pct_complete_pca = n_complete_pca / n_total,
      .groups = "drop"
    ) %>%
    mutate(
      audit_status = case_when(
        n_complete_pca == 0 ~ "critical_flag",
        n_complete_pca < 3 ~ "flag",
        TRUE ~ "pass"
      )
    )
}

write_table(cell_retention, "06_treatment_exposure_retention.csv")

cell_retention
```

# 06.7 Condition factor audit

Kc, Kf, and condition factor are treated as the same operational construct only if the available values are consistent with a weight-length condition-factor formula. This section checks the resolved condition-factor column against available weight and length columns.

```{r condition-factor-audit}
condition_row <- pca_variable_concordance %>%
  filter(planned_variable == "condition_factor")

condition_col <- if (nrow(condition_row) == 1) condition_row$resolved_variable[[1]] else NA_character_

weight_candidates <- c("wt", "weight", "body_weight", "mass")
length_candidates <- c("ls", "length", "standard_length", "body_length")

weight_col <- resolve_alias(bt1, paste(weight_candidates, collapse = "|"))
length_col <- resolve_alias(bt1, paste(length_candidates, collapse = "|"))

if (is.na(condition_col)) {
  condition_factor_audit <- tibble(
    condition_col = NA_character_,
    weight_col = weight_col,
    length_col = length_col,
    n_checked = 0L,
    formula_checked = FALSE,
    correlation_with_formula = NA_real_,
    median_abs_difference = NA_real_,
    audit_status = "not_available",
    details = "No condition factor column resolved in BT1."
  )
} else if (is.na(weight_col) || is.na(length_col)) {
  condition_factor_audit <- tibble(
    condition_col = condition_col,
    weight_col = weight_col,
    length_col = length_col,
    n_checked = 0L,
    formula_checked = FALSE,
    correlation_with_formula = NA_real_,
    median_abs_difference = NA_real_,
    audit_status = "partial",
    details = "Condition factor column exists, but weight or length column was not resolved."
  )
} else {
  k_observed <- coerce_numeric_if_possible(bt1[[condition_col]])
  weight <- coerce_numeric_if_possible(bt1[[weight_col]])
  length <- coerce_numeric_if_possible(bt1[[length_col]])

  k_formula_100 <- ifelse(!is.na(weight) & !is.na(length) & length != 0, (weight / (length^3)) * 100, NA_real_)
  k_formula_100000 <- ifelse(!is.na(weight) & !is.na(length) & length != 0, (weight / (length^3)) * 100000, NA_real_)

  cor_100 <- safe_cor(k_observed, k_formula_100)
  cor_100000 <- safe_cor(k_observed, k_formula_100000)

  if (!is.na(cor_100000) && (is.na(cor_100) || abs(cor_100000) > abs(cor_100))) {
    chosen_formula <- "weight / length^3 * 100000"
    k_formula <- k_formula_100000
    chosen_cor <- cor_100000
  } else {
    chosen_formula <- "weight / length^3 * 100"
    k_formula <- k_formula_100
    chosen_cor <- cor_100
  }

  ok <- !is.na(k_observed) & !is.na(k_formula)
  median_abs_diff <- if (sum(ok) > 0) median(abs(k_observed[ok] - k_formula[ok]), na.rm = TRUE) else NA_real_

  condition_factor_audit <- tibble(
    condition_col = condition_col,
    weight_col = weight_col,
    length_col = length_col,
    n_checked = sum(ok),
    formula_checked = TRUE,
    selected_formula = chosen_formula,
    correlation_with_formula = chosen_cor,
    median_abs_difference = median_abs_diff,
    observed_min = suppressWarnings(min(k_observed, na.rm = TRUE)),
    observed_max = suppressWarnings(max(k_observed, na.rm = TRUE)),
    audit_status = case_when(
      is.na(chosen_cor) ~ "flag",
      abs(chosen_cor) >= 0.99 ~ "pass",
      abs(chosen_cor) >= 0.95 ~ "partial",
      TRUE ~ "flag"
    ),
    details = "Condition factor compared against available weight-length formula."
  )
}

write_table(condition_factor_audit, "06_condition_factor_audit.csv")

condition_factor_audit
```

# 06.8 Outlier-flag impact audit

This section summarizes how many records would be affected by existing QC flags and by conservative IQR flags. It does not delete records automatically.

```{r outlier-impact-audit}
model_candidate_vars <- character(0)

if (!is.null(endpoint_registry)) {
  registry_names <- names(endpoint_registry)

  endpoint_col <- registry_names[
    stringr::str_to_lower(registry_names) %in%
      c("endpoint", "variable", "endpoint_variable")
  ]

  model_col <- registry_names[
    stringr::str_to_lower(registry_names) %in%
      c("model_tested", "model_candidate", "endpoint_model")
  ]

  if (length(endpoint_col) > 0 && length(model_col) > 0) {
    registry_model_values <- as.character(endpoint_registry[[model_col[1]]])
    registry_endpoint_values <- as.character(endpoint_registry[[endpoint_col[1]]])

    model_candidate_vars <- registry_endpoint_values[
      registry_model_values %in%
        c("TRUE", "true", "Yes", "yes", "1", "model-tested", "model_candidate")
    ]

    model_candidate_vars <- unique(model_candidate_vars[!is.na(model_candidate_vars)])
  }
}

if (length(model_candidate_vars) == 0) {
  model_candidate_vars <- c(
    "dist_total", "mov_total", "vel_mean", "stratum_pref",
    "time_bright", "first_choice_binary", "latency", "mtpc", "num_changes",
    "resistance_index", "attempts", "last_flux", "time_in_last_flux",
    "blood_sugar"
  )
}

model_candidate_vars <- unique(model_candidate_vars)

model_candidate_resolved <- tibble(variable = model_candidate_vars) %>%
  mutate(
    resolved_variable = purrr::map_chr(variable, ~ resolve_alias(bt1, .x)),
    available = !is.na(resolved_variable)
  ) %>%
  filter(available)

outlier_impact <- model_candidate_resolved %>%
  rowwise() %>%
  mutate(
    n_nonmissing = sum(!is.na(coerce_numeric_if_possible(bt1[[resolved_variable]]))),
    conservative_iqr_flags = sum(iqr_flags(bt1[[resolved_variable]], multiplier = 3), na.rm = TRUE),
    pct_conservative_iqr_flags = if_else(n_nonmissing > 0, conservative_iqr_flags / n_nonmissing, NA_real_)
  ) %>%
  ungroup()

existing_flag_cols <- names(bt1)[stringr::str_detect(names(bt1), "_flag$|flag$")]

existing_flag_summary <- if (length(existing_flag_cols) > 0) {
  tibble(flag_column = existing_flag_cols) %>%
    mutate(
      n_flagged = purrr::map_int(flag_column, ~ sum(as.logical(bt1[[.x]]), na.rm = TRUE)),
      pct_flagged = n_flagged / nrow(bt1)
    )
} else {
  tibble(flag_column = character(), n_flagged = integer(), pct_flagged = numeric())
}

write_table(outlier_impact, "06_outlier_impact_by_endpoint.csv")
write_table(existing_flag_summary, "06_existing_qc_flag_summary.csv")

outlier_impact
```

# 06.9 Single-PCA reconstruction

This section reconstructs the same single PCA using the agreed PCA endpoint set and the available PCA-ready BT1 variables. It does not run a behavioral-only or competing PCA. It is a concordance and stability check.

```{r single-pca-reconstruction}
pca_vars <- pca_variable_concordance %>%
  filter(pca_ready) %>%
  pull(resolved_variable)

if (length(pca_vars) < 2) {
  stop("Fewer than two PCA-ready variables were available. PCA stability cannot be assessed.")
}

bt1_pca <- bt1 %>%
  mutate(row_id_06 = row_number()) %>%
  select(any_of(c("row_id_06", "fish_id", "treatment", "exposure", pca_vars)))

pca_matrix <- bt1_pca %>%
  select(all_of(pca_vars)) %>%
  mutate(across(everything(), coerce_numeric_if_possible))

pca_complete_index <- stats::complete.cases(pca_matrix)
pca_matrix_complete <- pca_matrix[pca_complete_index, , drop = FALSE]

if (nrow(pca_matrix_complete) < 3) {
  stop("Fewer than three complete rows were available for the PCA endpoint set.")
}

pca_fit_06 <- stats::prcomp(pca_matrix_complete, center = TRUE, scale. = TRUE)

variance_explained <- tibble(
  component = paste0("PC", seq_along(pca_fit_06$sdev)),
  eigenvalue = pca_fit_06$sdev^2,
  variance_explained = eigenvalue / sum(eigenvalue),
  cumulative_variance = cumsum(variance_explained)
)

pca_loadings_06 <- as.data.frame(pca_fit_06$rotation) %>%
  rownames_to_column("variable") %>%
  pivot_longer(
    cols = starts_with("PC"),
    names_to = "component",
    values_to = "loading"
  ) %>%
  left_join(
    pca_variable_concordance %>% select(planned_variable, resolved_variable, domain),
    by = c("variable" = "resolved_variable")
  ) %>%
  mutate(abs_loading = abs(loading)) %>%
  arrange(component, desc(abs_loading))

pca_scores_06 <- as.data.frame(pca_fit_06$x) %>%
  mutate(row_id_06 = bt1_pca$row_id_06[pca_complete_index]) %>%
  left_join(bt1_pca %>% select(any_of(c("row_id_06", "fish_id", "treatment", "exposure"))), by = "row_id_06")

component_labels_06 <- pca_loadings_06 %>%
  group_by(component) %>%
  slice_max(abs_loading, n = 3, with_ties = FALSE) %>%
  summarise(
    top_variables = paste(variable, collapse = ", "),
    top_domains = paste(domain, collapse = " | "),
    suggested_label = simple_domain_label(domain),
    .groups = "drop"
  )

write_table(variance_explained, "06_pca_variance_explained.csv")
write_table(pca_loadings_06, "06_pca_loadings_reconstructed.csv")
write_table(pca_scores_06, "06_pca_scores_reconstructed.csv")
write_table(component_labels_06, "06_pca_component_labels_from_loadings.csv")

variance_explained %>% head(10)
```

```{r pca-component-labels}
component_labels_06
```

# 06.10 PCA concordance with Notebook 05 outputs

If Notebook 05 saved PCA loadings, this section compares them with the reconstructed PCA loadings. Sign flips are expected in PCA and are not treated as errors.

```{r pca-concordance-with-notebook05}
if (is.null(pca_loadings_existing)) {
  pca_concordance_05 <- tibble(
    check = "pca_loadings_from_notebook_05",
    status = "missing",
    details = "No Notebook 05 loading table found. The reconstructed PCA outputs from Notebook 06 are available for comparison after Notebook 05 is rerun."
  )
} else {
  existing_names <- names(pca_loadings_existing)

  existing_var_col <- existing_names[str_to_lower(existing_names) %in% c("variable", "endpoint", "pca_variable")]
  existing_component_col <- existing_names[str_to_lower(existing_names) %in% c("component", "pc")]
  existing_loading_col <- existing_names[str_to_lower(existing_names) %in% c("loading", "value", "charge", "coefficient")]

  if (length(existing_var_col) == 0 || length(existing_component_col) == 0 || length(existing_loading_col) == 0) {
    pca_concordance_05 <- tibble(
      check = "pca_loadings_from_notebook_05",
      status = "flag",
      details = paste("Could not identify variable/component/loading columns in Notebook 05 loadings. Columns:", paste(existing_names, collapse = ", "))
    )
  } else {
    existing_simple <- pca_loadings_existing %>%
      transmute(
        variable = as.character(.data[[existing_var_col[[1]]]]),
        component = as.character(.data[[existing_component_col[[1]]]]),
        loading_05 = as.numeric(.data[[existing_loading_col[[1]]]])
      )

    reconstructed_simple <- pca_loadings_06 %>%
      transmute(variable, component, loading_06 = loading)

    joined <- existing_simple %>%
      inner_join(reconstructed_simple, by = c("variable", "component"))

    component_correlations <- joined %>%
      group_by(component) %>%
      summarise(
        n_variables_compared = n(),
        loading_correlation = safe_cor(loading_05, loading_06),
        abs_loading_correlation = safe_cor(abs(loading_05), abs(loading_06)),
        .groups = "drop"
      )

    pca_concordance_05 <- component_correlations %>%
      mutate(
        check = paste0("PCA loading concordance ", component),
        status = case_when(
          is.na(abs_loading_correlation) ~ "flag",
          abs_loading_correlation >= 0.99 ~ "pass",
          abs_loading_correlation >= 0.95 ~ "partial",
          TRUE ~ "flag"
        ),
        details = paste0("Compared ", n_variables_compared, " variables; absolute-loading correlation = ", round(abs_loading_correlation, 4))
      ) %>%
      select(check, status, details)
  }
}

write_log(pca_concordance_05, "06_pca_concordance_with_notebook05.csv")

pca_concordance_05
```

# 06.11 PCA stability by bootstrap resampling

This is a stability audit of the same PCA endpoint set. It samples rows with replacement, reruns PCA on the same variables, and compares dominant loadings. This does not create a new PCA design.

```{r pca-bootstrap-stability}
n_boot <- 500
n_components_to_track <- min(5, ncol(pca_matrix_complete))

bootstrap_one <- function(i) {
  sampled_rows <- sample(seq_len(nrow(pca_matrix_complete)), size = nrow(pca_matrix_complete), replace = TRUE)
  boot_matrix <- pca_matrix_complete[sampled_rows, , drop = FALSE]

  fit <- tryCatch(
    stats::prcomp(boot_matrix, center = TRUE, scale. = TRUE),
    error = function(e) NULL
  )

  if (is.null(fit)) {
    return(tibble(iteration = i, component = NA_character_, variable = NA_character_, abs_loading = NA_real_))
  }

  as.data.frame(fit$rotation[, seq_len(n_components_to_track), drop = FALSE]) %>%
    rownames_to_column("variable") %>%
    pivot_longer(starts_with("PC"), names_to = "component", values_to = "loading") %>%
    group_by(component) %>%
    slice_max(abs(loading), n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(iteration = i, component, variable, abs_loading = abs(loading))
}

pca_boot_top <- map_dfr(seq_len(n_boot), bootstrap_one)

pca_boot_stability <- pca_boot_top %>%
  filter(!is.na(component)) %>%
  group_by(component, variable) %>%
  summarise(
    times_top_loading = n(),
    pct_top_loading = times_top_loading / n_boot,
    median_abs_loading_when_top = median(abs_loading, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(component, desc(pct_top_loading))

pca_boot_component_summary <- pca_boot_stability %>%
  group_by(component) %>%
  slice_max(pct_top_loading, n = 3, with_ties = FALSE) %>%
  summarise(
    top_loading_stability = paste0(variable, " (", round(pct_top_loading * 100, 1), "%)", collapse = "; "),
    .groups = "drop"
  )

write_table(pca_boot_stability, "06_pca_bootstrap_top_loading_stability.csv")
write_table(pca_boot_component_summary, "06_pca_bootstrap_component_summary.csv")

pca_boot_component_summary
```

# 06.12 PCA score-space annotation

This section summarizes treatment-by-exposure centroids in PCA space. These are descriptive annotations of the PCA visualization, not primary inferential tests.

```{r pca-score-space-annotation}
score_components <- names(pca_scores_06)[str_detect(names(pca_scores_06), "^PC\\d+$")]

pca_centroids <- pca_scores_06 %>%
  group_by(treatment, exposure) %>%
  summarise(
    n = n(),
    across(all_of(score_components), ~ mean(.x, na.rm = TRUE), .names = "centroid_{.col}"),
    .groups = "drop"
  )

write_table(pca_centroids, "06_pca_treatment_exposure_centroids.csv")

pca_centroids
```

```{r pca-score-plot-pc1-pc2, fig.width=8, fig.height=6}
if (all(c("PC1", "PC2", "treatment", "exposure") %in% names(pca_scores_06))) {
  ggplot(pca_scores_06, aes(x = PC1, y = PC2, color = treatment, shape = exposure)) +
    geom_point(alpha = 0.65, size = 2) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    geom_vline(xintercept = 0, linewidth = 0.3) +
    labs(
      title = "BT1 full-battery PCA score space",
      subtitle = "Descriptive visualization; component labels must be assigned from loadings",
      x = paste0("PC1 (", round(variance_explained$variance_explained[1] * 100, 1), "%)"),
      y = paste0("PC2 (", round(variance_explained$variance_explained[2] * 100, 1), "%)")
    ) +
    theme_minimal()
}
```

# 06.13 Notebook concordance decision register

```{r decision-register}
decision_register <- tibble(
  item = c(
    "Repository structure",
    "Endpoint registry",
    "Primary/secondary hierarchy",
    "Endpoint-level ethanol models",
    "Battery-context analysis",
    "Single PCA",
    "PCA axis labels",
    "Condition factor",
    "PCA score summaries",
    "Notebook 06 role"
  ),
  expected_rule = c(
    "Use existing repository structure.",
    "Use practical endpoint roles.",
    "Do not use primary/secondary endpoint hierarchy.",
    "Test treatment, exposure, and treatment x exposure for model-tested endpoints.",
    "Analyze battery context separately from ethanol pharmacology.",
    "Run one single full-battery BT1 PCA.",
    "Assign labels from loadings after PCA is run.",
    "Treat Kc/Kf/condition factor as the same construct if formula/range audit passes.",
    "Use PC scores and centroids as descriptive annotations, not primary inference.",
    "Begin with concordance audit, then robustness and stability checks."
  ),
  evidence_output = c(
    "06_file_concordance.csv",
    "06_endpoint_registry_concordance.csv",
    "06_endpoint_registry_concordance.csv",
    "03 model outputs + 06 audit",
    "04 battery outputs + 06 audit",
    "06_pca_variable_concordance.csv; 06_pca_loadings_reconstructed.csv",
    "06_pca_component_labels_from_loadings.csv",
    "06_condition_factor_audit.csv",
    "06_pca_treatment_exposure_centroids.csv",
    "This notebook"
  ),
  status = c(
    if_else(all(expected_repo_items$exists), "pass", "partial"),
    if_else(is.null(endpoint_registry), "flag", "pass"),
    if_else(is.null(endpoint_registry), "unknown", if_else(any(endpoint_registry_concordance$status == "flag" & endpoint_registry_concordance$check == "no_primary_secondary_columns"), "flag", "pass")),
    "checked_by_output_presence",
    "checked_by_output_presence",
    if_else(length(pca_vars) >= 2, "pass", "flag"),
    "pass",
    condition_factor_audit$audit_status[[1]],
    "pass",
    "pass"
  )
)

write_table(decision_register, "06_decision_register.csv")

decision_register
```

# 06.14 Final robustness summary

```{r final-summary}
critical_flags <- bind_rows(
  file_concordance %>%
    transmute(source = "file_concordance", item, status = audit_status, details = expected_path) %>%
    filter(status == "missing"),
  endpoint_registry_concordance %>%
    transmute(source = "endpoint_registry", item = check, status, details) %>%
    filter(status == "flag"),
  pca_variable_concordance %>%
    transmute(source = "pca_variable_concordance", item = planned_variable, status = if_else(pca_ready, "pass", "flag"), details = note) %>%
    filter(status == "flag"),
  condition_factor_audit %>%
    transmute(source = "condition_factor", item = "condition_factor", status = audit_status, details = details) %>%
    filter(status %in% c("flag", "partial")),
  pca_concordance_05 %>%
    transmute(source = "pca_concordance_05", item = check, status, details) %>%
    filter(status == "flag")
)

write_table(critical_flags, "06_critical_flags.csv")

summary_lines <- c(
  "# Notebook 06 summary",
  "",
  paste0("Rendered on: ", Sys.Date()),
  "",
  paste0("BT1 rows loaded: ", nrow(bt1)),
  paste0("BT2 rows loaded: ", nrow(bt2)),
  paste0("BT3 rows loaded: ", nrow(bt3)),
  paste0("PCA-ready variables resolved: ", length(pca_vars)),
  paste0("PCA complete-case rows: ", nrow(pca_matrix_complete)),
  paste0("Critical/flagged items: ", nrow(critical_flags)),
  "",
  "The PCA stability checks use the same fixed endpoint set and do not create a competing PCA.",
  "PCA score-space summaries are descriptive annotations of the visualization.",
  "Endpoint-level ethanol inference remains assigned to Notebook 03.",
  "Battery-context comparisons remain assigned to Notebook 04."
)

writeLines(summary_lines, file.path(paths$output_logs, "06_summary.md"))

critical_flags
```

# Completion statement

Notebook 06 is complete when the following outputs exist:

- `outputs/logs/06_file_concordance.csv`
- `outputs/logs/06_endpoint_registry_concordance.csv`
- `outputs/tables/06_pca_variable_concordance.csv`
- `outputs/tables/06_treatment_exposure_retention.csv`
- `outputs/tables/06_condition_factor_audit.csv`
- `outputs/tables/06_pca_loadings_reconstructed.csv`
- `outputs/tables/06_pca_component_labels_from_loadings.csv`
- `outputs/tables/06_pca_bootstrap_top_loading_stability.csv`
- `outputs/tables/06_pca_treatment_exposure_centroids.csv`
- `outputs/tables/06_decision_register.csv`
- `outputs/tables/06_critical_flags.csv`
- `outputs/logs/06_summary.md`

If critical flags remain, patch only the affected upstream notebook or output. If no critical flags remain, the pipeline can move to reporting tables and figures.
