---
title: "08 Statistical Methods Text"
author: "Caique Ortiz Colombo Arnoldi Fabris"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
---

# Purpose

This notebook drafts manuscript Methods text from completed notebooks and saved decision logs.

It does not introduce new statistical decisions. It does not rerun endpoint models, battery-context comparisons, PCA, or robustness checks. It converts the analysis pipeline into manuscript-ready methods language.

The Methods text is built from:

- Notebook 00W: data wrangling and derived variables.
- Notebook 00R: project setup, endpoint registry, and decision rules.
- Notebook 01: data audit and QC.
- Notebook 02: endpoint registry and model readiness.
- Notebook 03: endpoint-level ethanol models.
- Notebook 04: battery-context comparisons.
- Notebook 05: single BT1 full-battery PCA.
- Notebook 06: concordance and robustness checks.
- Notebook 07: reporting tables and figures.

# Notebook contract

Notebook 08 is complete when it exports:

- `outputs/reporting/text/08_statistical_methods_draft.md`
- `outputs/reporting/text/08_methods_data_preparation.md`
- `outputs/reporting/text/08_methods_endpoint_registry.md`
- `outputs/reporting/text/08_methods_endpoint_models.md`
- `outputs/reporting/text/08_methods_battery_context.md`
- `outputs/reporting/text/08_methods_pca.md`
- `outputs/reporting/text/08_methods_robustness.md`
- `outputs/reporting/logs/08_methods_text_manifest.csv`
- `outputs/reporting/logs/08_methods_source_map.csv`

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c(
  "dplyr", "tidyr", "readr", "stringr", "purrr", "tibble", "rlang"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_packages) > 0) {
  stop(
    "Missing required packages: ",
    paste(missing_packages, collapse = ", "),
    ". Install them before rendering this notebook."
  )
}

library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(purrr)
library(tibble)
library(rlang)
```

# 08.1 Project paths and helpers

```{r paths-and-helpers}
find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)

  while (TRUE) {
    has_rproj <- length(list.files(current, pattern = "\\.Rproj$", full.names = TRUE)) > 0
    has_git <- dir.exists(file.path(current, ".git"))
    has_data <- dir.exists(file.path(current, "data_processed")) || dir.exists(file.path(current, "data_raw"))

    if (has_rproj || has_git || has_data) {
      return(current)
    }

    parent <- dirname(current)
    if (identical(parent, current)) {
      return(normalizePath(start, winslash = "/", mustWork = TRUE))
    }

    current <- parent
  }
}

project_root <- find_project_root()

paths <- list(
  root = project_root,
  data_raw = file.path(project_root, "data_raw"),
  data_processed = file.path(project_root, "data_processed"),
  docs = file.path(project_root, "docs"),
  notebooks = file.path(project_root, "notebooks"),
  outputs = file.path(project_root, "outputs"),
  output_tables = file.path(project_root, "outputs", "tables"),
  output_figures = file.path(project_root, "outputs", "figures"),
  output_logs = file.path(project_root, "outputs", "logs"),
  reporting = file.path(project_root, "outputs", "reporting"),
  reporting_tables = file.path(project_root, "outputs", "reporting", "tables"),
  reporting_figures = file.path(project_root, "outputs", "reporting", "figures"),
  reporting_logs = file.path(project_root, "outputs", "reporting", "logs"),
  reporting_text = file.path(project_root, "outputs", "reporting", "text")
)

purrr::walk(
  paths[c("outputs", "reporting", "reporting_tables", "reporting_logs", "reporting_text")],
  dir.create,
  recursive = TRUE,
  showWarnings = FALSE
)

first_existing_path <- function(candidates) {
  existing <- candidates[file.exists(candidates)]
  if (length(existing) == 0) return(NA_character_)
  existing[[1]]
}

read_optional_csv <- function(candidates, label) {
  path <- first_existing_path(candidates)

  if (is.na(path)) {
    message("Missing optional input: ", label)
    return(NULL)
  }

  message("Loaded ", label, ": ", path)
  readr::read_csv(path, show_col_types = FALSE)
}

write_text_file <- function(lines, filename) {
  out <- file.path(paths$reporting_text, filename)
  writeLines(lines, out)
  out
}

write_methods_log <- function(df, filename) {
  out <- file.path(paths$reporting_logs, filename)
  readr::write_csv(df, out)
  out
}

value_or_na <- function(x) {
  if (is.null(x) || length(x) == 0) return(NA_character_)
  if (all(is.na(x))) return(NA_character_)
  as.character(x[[1]])
}

collapse_nonmissing <- function(x, sep = ", ") {
  x <- unique(as.character(x))
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0) return(NA_character_)
  paste(x, collapse = sep)
}

safe_nrow <- function(x) {
  if (is.null(x)) return(NA_integer_)
  nrow(x)
}

safe_has_col <- function(df, candidates) {
  if (is.null(df)) return(NA_character_)
  hit <- candidates[candidates %in% names(df)]
  if (length(hit) > 0) return(hit[[1]])

  lower_map <- tibble(actual = names(df), lower = stringr::str_to_lower(names(df)))
  lower_candidates <- stringr::str_to_lower(candidates)
  lower_hit <- lower_map %>% filter(lower %in% lower_candidates)

  if (nrow(lower_hit) > 0) return(lower_hit$actual[[1]])
  NA_character_
}

text_block <- function(title, paragraphs) {
  c(
    paste0("## ", title),
    "",
    paragraphs,
    ""
  )
}

make_sentence <- function(x) {
  x <- stringr::str_squish(x)
  if (x == "") return(x)
  if (!stringr::str_detect(x, "[.!?]$")) paste0(x, ".") else x
}

cat("Project root:", project_root, "\n")
```

# 08.2 Load source outputs

```{r load-source-outputs}
dataset_inventory <- read_optional_csv(
  c(
    file.path(paths$reporting_tables, "Table_01_dataset_qc_summary.csv"),
    file.path(paths$output_tables, "06_dataset_inventory.csv"),
    file.path(paths$output_tables, "01_dataset_inventory.csv"),
    file.path(paths$output_logs, "01_dataset_inventory.csv")
  ),
  "dataset inventory"
)

cell_retention <- read_optional_csv(
  c(
    file.path(paths$reporting_tables, "Table_01b_treatment_exposure_cells.csv"),
    file.path(paths$output_tables, "06_treatment_exposure_retention.csv"),
    file.path(paths$output_tables, "01_cell_size_table.csv"),
    file.path(paths$output_logs, "01_cell_size_table.csv")
  ),
  "treatment-by-exposure cells"
)

endpoint_registry <- read_optional_csv(
  c(
    file.path(paths$reporting_tables, "Table_02_endpoint_registry.csv"),
    file.path(paths$output_tables, "02_endpoint_registry.csv"),
    file.path(paths$output_tables, "00_endpoint_registry.csv"),
    file.path(paths$output_logs, "02_endpoint_registry.csv"),
    file.path(paths$output_logs, "00_endpoint_registry.csv")
  ),
  "endpoint registry"
)

model_readiness <- read_optional_csv(
  c(
    file.path(paths$output_tables, "02_model_readiness.csv"),
    file.path(paths$output_tables, "02_model_decision_matrix.csv"),
    file.path(paths$output_logs, "02_model_readiness.csv")
  ),
  "model readiness"
)

ethanol_models <- read_optional_csv(
  c(
    file.path(paths$reporting_tables, "Table_04_ethanol_endpoint_models.csv"),
    file.path(paths$output_tables, "03_ethanol_factorial_results.csv"),
    file.path(paths$output_tables, "03_factorial_results.csv"),
    file.path(paths$output_tables, "03_model_results.csv"),
    file.path(paths$output_tables, "03_endpoint_model_results.csv")
  ),
  "ethanol model outputs"
)

battery_summary <- read_optional_csv(
  c(
    file.path(paths$reporting_tables, "Table_05_battery_context_summary.csv"),
    file.path(paths$output_tables, "04_battery_context_summary.csv"),
    file.path(paths$output_tables, "04_battery_method_summary.csv"),
    file.path(paths$output_tables, "04_battery_comparison_summary.csv"),
    file.path(paths$output_tables, "04_battery_results.csv")
  ),
  "battery-context outputs"
)

pca_summary <- read_optional_csv(
  c(
    file.path(paths$reporting_tables, "Table_06_pca_summary.csv")
  ),
  "PCA reporting summary"
)

pca_variance <- read_optional_csv(
  c(
    file.path(paths$output_tables, "05_pca_variance_explained.csv"),
    file.path(paths$output_tables, "05_pca_eigenvalues.csv"),
    file.path(paths$output_tables, "06_pca_variance_explained.csv")
  ),
  "PCA variance explained"
)

pca_variable_concordance <- read_optional_csv(
  c(
    file.path(paths$output_tables, "06_pca_variable_concordance.csv"),
    file.path(paths$output_tables, "05_pca_variable_concordance.csv"),
    file.path(paths$output_tables, "05_pca_variable_availability.csv")
  ),
  "PCA variable concordance"
)

pca_labels <- read_optional_csv(
  c(
    file.path(paths$output_tables, "05_pca_component_labels_from_loadings.csv"),
    file.path(paths$output_tables, "06_pca_component_labels_from_loadings.csv")
  ),
  "PCA component labels"
)

condition_factor_audit <- read_optional_csv(
  c(
    file.path(paths$output_tables, "06_condition_factor_audit.csv"),
    file.path(paths$output_logs, "06_condition_factor_audit.csv"),
    file.path(paths$output_tables, "05_condition_factor_audit.csv")
  ),
  "condition factor audit"
)

decision_register <- read_optional_csv(
  c(
    file.path(paths$output_tables, "06_decision_register.csv"),
    file.path(paths$output_logs, "06_decision_register.csv")
  ),
  "Notebook 06 decision register"
)

critical_flags <- read_optional_csv(
  c(
    file.path(paths$output_tables, "06_critical_flags.csv"),
    file.path(paths$output_logs, "06_critical_flags.csv")
  ),
  "critical flags"
)

reporting_manifest_07 <- read_optional_csv(
  c(
    file.path(paths$reporting_logs, "07_reporting_manifest.csv")
  ),
  "Notebook 07 reporting manifest"
)
```

# 08.3 Extract method facts from outputs

```{r extract-method-facts}
dataset_summary <- if (!is.null(dataset_inventory)) {
  dataset_inventory
} else {
  tibble()
}

dataset_rows_text <- if (!is.null(dataset_inventory) && all(c("dataset", "rows") %in% names(dataset_inventory))) {
  dataset_inventory %>%
    mutate(summary = paste0(dataset, " n = ", rows)) %>%
    pull(summary) %>%
    collapse_nonmissing(sep = "; ")
} else {
  NA_character_
}

cell_count_text <- if (!is.null(cell_retention) && all(c("treatment", "exposure") %in% names(cell_retention))) {
  n_col <- safe_has_col(cell_retention, c("n", "n_total", "n_complete_pca"))
  if (!is.na(n_col)) {
    paste0(
      "Treatment-by-exposure cell sizes were tabulated for the 0, 0.5, and 1% EtOH groups across 1h, 24h, and 96h exposure levels."
    )
  } else {
    "Treatment-by-exposure cell structure was audited."
  }
} else {
  "Treatment-by-exposure cell structure was audited where the required design columns were available."
}

endpoint_col <- safe_has_col(endpoint_registry, c("endpoint", "variable", "endpoint_variable"))
domain_col <- safe_has_col(endpoint_registry, c("domain"))
pca_col <- safe_has_col(endpoint_registry, c("pca_included", "pca_candidate", "included_in_pca"))
model_col <- safe_has_col(endpoint_registry, c("model_tested", "model_candidate", "endpoint_model"))
battery_col <- safe_has_col(endpoint_registry, c("battery_comparable", "battery_candidate", "battery_comparison"))

registry_n <- safe_nrow(endpoint_registry)

registry_domains <- if (!is.null(endpoint_registry) && !is.na(domain_col)) {
  collapse_nonmissing(endpoint_registry[[domain_col]])
} else {
  NA_character_
}

pca_vars_text <- if (!is.null(pca_variable_concordance)) {
  pca_var_col <- safe_has_col(pca_variable_concordance, c("planned_variable", "endpoint", "variable"))
  pca_ready_col <- safe_has_col(pca_variable_concordance, c("pca_ready", "included", "available_in_bt1"))

  if (!is.na(pca_var_col) && !is.na(pca_ready_col)) {
    ready_values <- as.character(pca_variable_concordance[[pca_ready_col]])
    vars <- pca_variable_concordance[[pca_var_col]][ready_values %in% c("TRUE", "true", "Yes", "yes", "1", "pass")]
    collapse_nonmissing(vars)
  } else if (!is.na(pca_var_col)) {
    collapse_nonmissing(pca_variable_concordance[[pca_var_col]])
  } else {
    NA_character_
  }
} else if (!is.null(endpoint_registry) && !is.na(endpoint_col) && !is.na(pca_col)) {
  pca_values <- as.character(endpoint_registry[[pca_col]])
  vars <- endpoint_registry[[endpoint_col]][pca_values %in% c("TRUE", "true", "Yes", "yes", "1", "pca-included")]
  collapse_nonmissing(vars)
} else {
  "dist_total, mov_total, vel_mean, stratum_pref, time_bright, first_choice_binary, latency, mtpc, num_changes, resistance_index/irn_computed, attempts, blood_sugar, and condition factor (K) when valid"
}

pca_n_text <- if (!is.null(pca_variable_concordance)) {
  pca_ready_col <- safe_has_col(pca_variable_concordance, c("pca_ready", "included", "available_in_bt1"))
  if (!is.na(pca_ready_col)) {
    ready_values <- as.character(pca_variable_concordance[[pca_ready_col]])
    n_ready <- sum(ready_values %in% c("TRUE", "true", "Yes", "yes", "1", "pass"), na.rm = TRUE)
    paste0(n_ready, " PCA-ready variables")
  } else {
    paste0(nrow(pca_variable_concordance), " audited PCA candidate variables")
  }
} else {
  "the audited PCA endpoint set"
}

pca_variance_text <- if (!is.null(pca_variance)) {
  component_col <- safe_has_col(pca_variance, c("component", "PC", "pc"))
  variance_col <- safe_has_col(pca_variance, c("variance_explained", "prop_variance", "proportion", "variance"))

  if (!is.na(component_col) && !is.na(variance_col)) {
    pca_variance %>%
      mutate(
        component_value = as.character(.data[[component_col]]),
        variance_value = suppressWarnings(as.numeric(.data[[variance_col]])),
        variance_pct = ifelse(variance_value <= 1, variance_value * 100, variance_value),
        summary = paste0(component_value, " = ", round(variance_pct, 1), "%")
      ) %>%
      head(5) %>%
      pull(summary) %>%
      collapse_nonmissing(sep = "; ")
  } else {
    NA_character_
  }
} else {
  NA_character_
}

condition_status_text <- if (!is.null(condition_factor_audit)) {
  status_col <- safe_has_col(condition_factor_audit, c("audit_status", "status"))
  condition_col <- safe_has_col(condition_factor_audit, c("condition_col", "condition_factor_column"))
  formula_col <- safe_has_col(condition_factor_audit, c("selected_formula", "formula"))

  status <- if (!is.na(status_col)) value_or_na(condition_factor_audit[[status_col]]) else NA_character_
  condition_name <- if (!is.na(condition_col)) value_or_na(condition_factor_audit[[condition_col]]) else "condition factor (K)"
  formula <- if (!is.na(formula_col)) value_or_na(condition_factor_audit[[formula_col]]) else "a scaled W/L^3-type formula"

  paste0("Condition factor was audited as ", condition_name, " using ", formula, "; audit status: ", status, ".")
} else {
  "Condition factor (K) was treated as eligible only if the formula, units, range, and availability audit supported inclusion."
}

model_families_text <- if (!is.null(model_readiness)) {
  family_col <- safe_has_col(model_readiness, c("model_family", "recommended_model", "model"))
  if (!is.na(family_col)) {
    collapse_nonmissing(model_readiness[[family_col]])
  } else {
    "model families assigned by endpoint type and diagnostics"
  }
} else if (!is.null(ethanol_models)) {
  family_col <- safe_has_col(ethanol_models, c("model_family", "model", "test"))
  if (!is.na(family_col)) {
    collapse_nonmissing(ethanol_models[[family_col]])
  } else {
    "model families assigned by endpoint type and diagnostics"
  }
} else {
  "model families assigned by endpoint type and diagnostics"
}

battery_classification_text <- if (!is.null(battery_summary)) {
  class_col <- safe_has_col(battery_summary, c("classification", "decision", "interpretation", "status"))
  if (!is.na(class_col)) {
    classes <- collapse_nonmissing(battery_summary[[class_col]])
    paste0("Battery-context results were classified using the saved categories: ", classes, ".")
  } else {
    "Battery-context outputs were summarized using saved comparison tables."
  }
} else {
  "Battery-context outputs were summarized using saved comparison tables."
}

critical_flag_text <- if (!is.null(critical_flags)) {
  paste0(nrow(critical_flags), " critical or flagged robustness items were listed in Notebook 06 outputs.")
} else {
  "Robustness flags were checked using Notebook 06 outputs."
}

facts <- tibble(
  fact = c(
    "dataset_rows_text",
    "cell_count_text",
    "registry_n",
    "registry_domains",
    "model_families_text",
    "battery_classification_text",
    "pca_vars_text",
    "pca_n_text",
    "pca_variance_text",
    "condition_status_text",
    "critical_flag_text"
  ),
  value = c(
    dataset_rows_text,
    cell_count_text,
    as.character(registry_n),
    registry_domains,
    model_families_text,
    battery_classification_text,
    pca_vars_text,
    pca_n_text,
    pca_variance_text,
    condition_status_text,
    critical_flag_text
  )
)

facts
```

# 08.4 Methods text: data preparation and QC

```{r methods-data-preparation}
data_preparation_paragraphs <- c(
  make_sentence(
    "Raw source data were translated into battery-level analysis files using a dedicated Python wrangling layer. Fish identifiers, treatment labels, exposure timing, battery labels, and acquisition metadata were retained for downstream audit. Planned derived variables, including locomotor summaries, resistance metrics, first-choice coding, and condition factor where available, were computed only when the corresponding formula or coding rule was explicit"
  ),
  make_sentence(
    paste0(
      "Wrangled battery-level files were stored in `data_processed` and then audited before inferential analysis. The QC layer checked dataset dimensions, fish-ID integrity, treatment-by-exposure cell structure, missingness, impossible or out-of-range values, protocol-duration consistency, derived-variable consistency, and behavioral outlier flags. ",
      cell_count_text
    )
  ),
  make_sentence(
    paste0(
      "Audited BT1, BT2, and BT3 datasets were used for downstream analyses. Dataset sizes were summarized as: ",
      ifelse(is.na(dataset_rows_text), "see the dataset inventory output", dataset_rows_text)
    )
  )
)

data_preparation_text <- text_block("Data preparation and quality control", data_preparation_paragraphs)

path_data_prep <- write_text_file(data_preparation_text, "08_methods_data_preparation.md")
path_data_prep
```

# 08.5 Methods text: endpoint registry and model readiness

```{r methods-endpoint-registry}
endpoint_registry_paragraphs <- c(
  make_sentence(
    "Endpoints were organized using an endpoint registry rather than a primary/secondary endpoint hierarchy. Each variable was assigned practical roles, including PCA inclusion, endpoint-model eligibility, battery-comparison eligibility, covariate/QC use, or exclusion"
  ),
  make_sentence(
    paste0(
      "The registry covered the following endpoint domains: ",
      ifelse(is.na(registry_domains), "locomotion, spatial position, light-dark behavior, endurance, physiology, and biometry/condition", registry_domains)
    )
  ),
  make_sentence(
    "Model readiness was evaluated using endpoint type, distributional diagnostics, variance diagnostics, zero-inflation checks, count overdispersion screening, and binary separation checks where applicable. These diagnostics informed the model family assigned to each model-tested endpoint"
  )
)

endpoint_registry_text <- text_block("Endpoint registry and model readiness", endpoint_registry_paragraphs)

path_endpoint_registry <- write_text_file(endpoint_registry_text, "08_methods_endpoint_registry.md")
path_endpoint_registry
```

# 08.6 Methods text: endpoint-level ethanol models

```{r methods-endpoint-models}
endpoint_model_paragraphs <- c(
  make_sentence(
    "Endpoint-level ethanol effects were tested using factorial models for registry-approved model-tested endpoints. For each endpoint, the target model evaluated treatment, exposure time, and the treatment-by-exposure interaction"
  ),
  make_sentence(
    paste0(
      "Model families were selected according to endpoint type and model-readiness diagnostics. The model-family outputs included: ",
      model_families_text
    )
  ),
  make_sentence(
    "Post hoc contrasts were considered only after relevant main or interaction effects. Multiple-testing adjustment was applied within defined model-output families where the corresponding output was generated. Endpoint-level models remained the basis for ethanol-effect inference"
  )
)

endpoint_model_text <- text_block("Endpoint-level ethanol models", endpoint_model_paragraphs)

path_endpoint_models <- write_text_file(endpoint_model_text, "08_methods_endpoint_models.md")
path_endpoint_models
```

# 08.7 Methods text: battery-context comparisons

```{r methods-battery-context}
battery_context_paragraphs <- c(
  make_sentence(
    "Battery-context analyses were conducted separately from endpoint-level ethanol models. Comparable endpoints were evaluated across valid battery pairs: BT1 versus BT2 for LDT/NTT endpoints and BT1 versus BT3 for endurance endpoints where protocol comparability was supported"
  ),
  make_sentence(
    "Battery-context comparisons were performed within treatment-by-exposure cells when the required endpoint and cell data were available. Effect sizes and uncertainty intervals were reported where supported by the saved outputs"
  ),
  make_sentence(
    paste0(
      battery_classification_text,
      " Non-significant battery-context comparisons were not interpreted as absence of difference unless formal equivalence criteria were predefined and satisfied"
    )
  )
)

battery_context_text <- text_block("Battery-context analysis", battery_context_paragraphs)

path_battery_context <- write_text_file(battery_context_text, "08_methods_battery_context.md")
path_battery_context
```

# 08.8 Methods text: PCA

```{r methods-pca}
pca_paragraphs <- c(
  make_sentence(
    "A single BT1 full-battery PCA was used to visualize multivariate cross-domain endpoint structure. BT1 was used because it contains the full sequential battery measured in the same fish"
  ),
  make_sentence(
    paste0(
      "The PCA endpoint set included ",
      ifelse(is.na(pca_vars_text), "the audited full-battery endpoint set", pca_vars_text),
      ". Condition factor was reported as condition factor (K) and included only if the formula, units, range, and availability audit supported inclusion"
    )
  ),
  make_sentence(
    "Before PCA, included endpoint values were z-transformed by centering and scaling so that behavioral, endurance, physiological, and condition-factor variables were placed on a common scale. PCA was then performed on the centered and scaled matrix"
  ),
  make_sentence(
    paste0(
      "PCA variance, loadings, scores, and visualization outputs were saved for reporting. Components were interpreted from dominant loadings after the PCA was run; component signs were treated as arbitrary. ",
      ifelse(is.na(pca_variance_text), "", paste0("The saved variance summary was: ", pca_variance_text))
    )
  ),
  make_sentence(
    "PC-score plots, centroids, and biplots were treated as descriptive annotations of multivariate structure. PCA was not used as a replacement for endpoint-level ethanol models"
  ),
  make_sentence(condition_status_text)
)

pca_text <- text_block("Principal component analysis", pca_paragraphs)

path_pca <- write_text_file(pca_text, "08_methods_pca.md")
path_pca
```

# 08.9 Methods text: robustness and concordance checks

```{r methods-robustness}
robustness_paragraphs <- c(
  make_sentence(
    "A concordance and robustness notebook was used as a gatekeeper before reporting and manuscript writing. It checked whether notebooks 00W-05 matched the locked analysis structure, including repository layout, endpoint registry, absence of primary/secondary endpoint hierarchy, single-PCA logic, and loading-based PCA interpretation"
  ),
  make_sentence(
    "Robustness checks included endpoint-registry concordance, treatment-by-exposure retention, condition-factor validation, outlier-flag impact, PCA variable-set concordance, PCA loading concordance, and PCA stability checks where supported by saved outputs"
  ),
  make_sentence(
    paste0(
      critical_flag_text,
      " Issues were classified as no patch, wording patch, code patch, or roadmap revision before manuscript-facing reporting"
    )
  )
)

robustness_text <- text_block("Concordance and robustness checks", robustness_paragraphs)

path_robustness <- write_text_file(robustness_text, "08_methods_robustness.md")
path_robustness
```

# 08.10 Combined statistical methods draft

```{r combined-methods-draft}
methods_intro <- c(
  "# Statistical methods draft",
  "",
  "This draft is generated from completed analysis notebooks and saved output logs. It should be edited for journal style, but new statistical decisions should not be introduced at this stage.",
  ""
)

combined_methods <- c(
  methods_intro,
  data_preparation_text,
  endpoint_registry_text,
  endpoint_model_text,
  battery_context_text,
  pca_text,
  robustness_text
)

path_combined <- write_text_file(combined_methods, "08_statistical_methods_draft.md")

cat(paste(combined_methods, collapse = "\n"))
```

# 08.11 Methods source map

```{r methods-source-map}
methods_source_map <- tibble(
  methods_section = c(
    "Data preparation and quality control",
    "Endpoint registry and model readiness",
    "Endpoint-level ethanol models",
    "Battery-context analysis",
    "Principal component analysis",
    "Concordance and robustness checks"
  ),
  main_source_notebook = c(
    "00_data_wrangling.ipynb; 01_data_audit_and_qc.Rmd",
    "00_project_setup.Rmd; 02_endpoint_hierarchy_and_assumptions.Rmd",
    "03_ethanol_factorial_models.Rmd",
    "04_battery_effect_analysis.Rmd",
    "05_multivariate_pca_structure.Rmd; 06_sensitivity_and_robustness.Rmd",
    "06_sensitivity_and_robustness.Rmd"
  ),
  supporting_outputs = c(
    "dataset inventory, cell-size table, QC logs",
    "endpoint registry, model-readiness matrix",
    "ethanol model results, effect estimates, contrasts, FDR tables",
    "battery-comparison summary, effect-size tables, classification table",
    "PCA variable concordance, variance table, loadings, scores, condition-factor audit",
    "decision register, critical flags, PCA stability, condition-factor audit"
  ),
  manuscript_claim_boundary = c(
    "Describes preparation and QC only; no biological conclusion.",
    "Describes variable roles and model readiness; no endpoint ranking.",
    "Supports endpoint-level ethanol-effect claims.",
    "Supports battery-context claims only; inconclusive comparisons remain inconclusive.",
    "Supports multivariate visualization claims only; not proof of ethanol effects.",
    "Supports robustness classifications and patch decisions."
  )
)

path_source_map <- write_methods_log(methods_source_map, "08_methods_source_map.csv")

methods_source_map
```

# 08.12 Methods text manifest

```{r methods-text-manifest}
methods_manifest <- tibble(
  output_name = c(
    "08_statistical_methods_draft.md",
    "08_methods_data_preparation.md",
    "08_methods_endpoint_registry.md",
    "08_methods_endpoint_models.md",
    "08_methods_battery_context.md",
    "08_methods_pca.md",
    "08_methods_robustness.md",
    "08_methods_source_map.csv"
  ),
  output_type = c(
    "markdown_text",
    "markdown_text",
    "markdown_text",
    "markdown_text",
    "markdown_text",
    "markdown_text",
    "markdown_text",
    "csv_log"
  ),
  path = c(
    path_combined,
    path_data_prep,
    path_endpoint_registry,
    path_endpoint_models,
    path_battery_context,
    path_pca,
    path_robustness,
    path_source_map
  ),
  status = ifelse(file.exists(path), "generated", "missing"),
  notes = c(
    "Combined statistical methods draft.",
    "Data preparation and QC methods text.",
    "Endpoint registry and model-readiness methods text.",
    "Endpoint-level modeling methods text.",
    "Battery-context comparison methods text.",
    "PCA methods text.",
    "Robustness methods text.",
    "Traceability map from methods text to notebook outputs."
  )
)

path_manifest <- write_methods_log(methods_manifest, "08_methods_text_manifest.csv")

methods_manifest
```

# Completion statement

Notebook 08 is complete when the combined statistical methods draft and section-level Methods text files are written under `outputs/reporting/text/`.

The exported text should be reviewed for journal style and citation placement. Statistical decisions should not be changed in this notebook. If the Methods text reveals a missing or inconsistent decision, patch the relevant upstream notebook or decision log, then rerender Notebook 08.
