---
title: "09 Reproducibility Audit"
author: "Caique Ortiz Colombo Arnoldi Fabris"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
---

# Purpose

This notebook performs the final reproducibility and file-inventory audit before manuscript writing.

It does not introduce new statistical decisions. It does not rerun endpoint models, battery-context comparisons, PCA, or robustness analyses. Its role is to verify whether the project files, outputs, package environment, reporting bundle, and methods text are traceable and internally consistent.

The audit covers:

- repository and notebook inventory;
- data and output file inventory;
- expected-output presence across notebooks 00W-08;
- package and session information;
- reporting-bundle consistency;
- methods-text consistency;
- claim-source mapping;
- README-ready execution summary.

# Notebook contract

Notebook 09 is complete when it exports:

- `outputs/reproducibility/09_file_inventory.csv`
- `outputs/reproducibility/09_notebook_inventory.csv`
- `outputs/reproducibility/09_expected_output_audit.csv`
- `outputs/reproducibility/09_package_versions.csv`
- `outputs/reproducibility/09_session_info.txt`
- `outputs/reproducibility/09_reporting_bundle_audit.csv`
- `outputs/reproducibility/09_methods_text_audit.csv`
- `outputs/reproducibility/09_claim_source_map.csv`
- `outputs/reproducibility/09_readme_execution_summary.md`
- `outputs/reproducibility/09_reproducibility_manifest.csv`

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c(
  "dplyr", "tidyr", "readr", "stringr", "purrr", "tibble", "rlang"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_packages) > 0) {
  stop(
    "Missing required packages: ",
    paste(missing_packages, collapse = ", "),
    ". Install them before rendering this notebook."
  )
}

library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(purrr)
library(tibble)
library(rlang)
```

# 09.1 Project paths and helpers

```{r paths-and-helpers}
find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)

  while (TRUE) {
    has_rproj <- length(list.files(current, pattern = "\\.Rproj$", full.names = TRUE)) > 0
    has_git <- dir.exists(file.path(current, ".git"))
    has_data <- dir.exists(file.path(current, "data_processed")) || dir.exists(file.path(current, "data_raw"))

    if (has_rproj || has_git || has_data) {
      return(current)
    }

    parent <- dirname(current)
    if (identical(parent, current)) {
      return(normalizePath(start, winslash = "/", mustWork = TRUE))
    }

    current <- parent
  }
}

project_root <- find_project_root()

paths <- list(
  root = project_root,
  data_raw = file.path(project_root, "data_raw"),
  data_processed = file.path(project_root, "data_processed"),
  docs = file.path(project_root, "docs"),
  notebooks = file.path(project_root, "notebooks"),
  outputs = file.path(project_root, "outputs"),
  output_tables = file.path(project_root, "outputs", "tables"),
  output_figures = file.path(project_root, "outputs", "figures"),
  output_logs = file.path(project_root, "outputs", "logs"),
  output_models = file.path(project_root, "outputs", "models"),
  reporting = file.path(project_root, "outputs", "reporting"),
  reporting_tables = file.path(project_root, "outputs", "reporting", "tables"),
  reporting_figures = file.path(project_root, "outputs", "reporting", "figures"),
  reporting_logs = file.path(project_root, "outputs", "reporting", "logs"),
  reporting_text = file.path(project_root, "outputs", "reporting", "text"),
  reproducibility = file.path(project_root, "outputs", "reproducibility"),
  python = file.path(project_root, "python"),
  R = file.path(project_root, "R")
)

purrr::walk(
  paths[c("outputs", "reproducibility")],
  dir.create,
  recursive = TRUE,
  showWarnings = FALSE
)

first_existing_path <- function(candidates) {
  existing <- candidates[file.exists(candidates)]
  if (length(existing) == 0) return(NA_character_)
  existing[[1]]
}

safe_file_size <- function(path) {
  if (!file.exists(path)) return(NA_real_)
  file.info(path)$size
}

safe_mtime <- function(path) {
  if (!file.exists(path)) return(as.POSIXct(NA))
  file.info(path)$mtime
}

safe_md5 <- function(path) {
  if (!file.exists(path) || dir.exists(path)) return(NA_character_)
  tryCatch(
    as.character(tools::md5sum(path)),
    error = function(e) NA_character_
  )
}

write_repro_table <- function(df, filename) {
  out <- file.path(paths$reproducibility, filename)
  readr::write_csv(df, out)
  out
}

write_repro_text <- function(lines, filename) {
  out <- file.path(paths$reproducibility, filename)
  writeLines(lines, out)
  out
}

add_manifest <- function(manifest, output_name, output_type, status, path, notes = "") {
  bind_rows(
    manifest,
    tibble(
      output_name = output_name,
      output_type = output_type,
      status = status,
      path = path,
      notes = notes
    )
  )
}

read_optional_csv <- function(candidates, label) {
  path <- first_existing_path(candidates)

  if (is.na(path)) {
    message("Missing optional input: ", label)
    return(NULL)
  }

  message("Loaded ", label, ": ", path)
  readr::read_csv(path, show_col_types = FALSE)
}

read_optional_text <- function(candidates, label) {
  path <- first_existing_path(candidates)

  if (is.na(path)) {
    message("Missing optional text: ", label)
    return(NULL)
  }

  message("Loaded ", label, ": ", path)
  readLines(path, warn = FALSE)
}

audit_file_presence <- function(label, candidates, required = TRUE, source_notebook = NA_character_) {
  path <- first_existing_path(candidates)

  tibble(
    label = label,
    source_notebook = source_notebook,
    required = required,
    status = ifelse(!is.na(path), "present", ifelse(required, "missing_required", "missing_optional")),
    path = path,
    size_bytes = ifelse(!is.na(path), safe_file_size(path), NA_real_),
    modified_time = ifelse(!is.na(path), as.character(safe_mtime(path)), NA_character_),
    md5 = ifelse(!is.na(path), safe_md5(path), NA_character_),
    candidates_checked = paste(candidates, collapse = " | ")
  )
}

extract_forbidden_terms <- function(lines, patterns) {
  if (is.null(lines)) {
    return(tibble(pattern = names(patterns), count = NA_integer_, examples = NA_character_))
  }

  text <- paste(lines, collapse = "\n")

  purrr::map2_dfr(names(patterns), patterns, function(label, pattern) {
    matches <- stringr::str_extract_all(text, pattern, simplify = FALSE)[[1]]
    examples <- unique(matches)
    examples <- examples[examples != ""]
    tibble(
      pattern = label,
      count = length(matches),
      examples = ifelse(length(examples) == 0, NA_character_, paste(head(examples, 5), collapse = " | "))
    )
  })
}

cat("Project root:", project_root, "\n")
```

# 09.2 Repository file inventory

```{r repository-file-inventory}
inventory_roots <- c(
  paths$data_raw,
  paths$data_processed,
  paths$docs,
  paths$notebooks,
  paths$outputs,
  paths$python,
  paths$R
)

inventory_files <- inventory_roots[dir.exists(inventory_roots)] %>%
  purrr::map(function(root_dir) {
    list.files(root_dir, recursive = TRUE, full.names = TRUE, all.files = FALSE)
  }) %>%
  unlist(use.names = FALSE) %>%
  unique()

file_inventory <- tibble(full_path = inventory_files) %>%
  filter(file.exists(full_path), !dir.exists(full_path)) %>%
  mutate(
    relative_path = stringr::str_replace(full_path, paste0("^", stringr::fixed(paths$root), "/?"), ""),
    folder = dirname(relative_path),
    filename = basename(relative_path),
    extension = stringr::str_to_lower(tools::file_ext(filename)),
    size_bytes = file.info(full_path)$size,
    modified_time = as.character(file.info(full_path)$mtime),
    md5 = purrr::map_chr(full_path, safe_md5),
    file_class = case_when(
      extension %in% c("rmd", "ipynb", "r", "py") ~ "code_or_notebook",
      extension %in% c("csv", "tsv", "xlsx", "xls") ~ "data_or_table",
      extension %in% c("png", "jpg", "jpeg", "pdf", "svg") ~ "figure_or_document",
      extension %in% c("md", "txt") ~ "text_or_log",
      TRUE ~ "other"
    )
  ) %>%
  arrange(relative_path)

path_file_inventory <- write_repro_table(file_inventory, "09_file_inventory.csv")

file_inventory %>% head(30)
```

# 09.3 Notebook inventory

```{r notebook-inventory}
expected_notebooks <- tibble(
  order = c("00W", "00R", "01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11"),
  expected_file = c(
    "python/00_data_wrangling.ipynb",
    "notebooks/00_project_setup.Rmd",
    "notebooks/01_data_audit_and_qc.Rmd",
    "notebooks/02_endpoint_hierarchy_and_assumptions.Rmd",
    "notebooks/03_ethanol_factorial_models.Rmd",
    "notebooks/04_battery_effect_analysis.Rmd",
    "notebooks/05_multivariate_pca_structure.Rmd",
    "notebooks/06_sensitivity_and_robustness.Rmd",
    "notebooks/07_reporting_tables_and_figures.Rmd",
    "notebooks/08_statistical_methods_text.Rmd",
    "notebooks/09_reproducibility_audit.Rmd",
    "notebooks/10_manuscript_results_and_discussion.Rmd",
    "notebooks/11_submission_package_audit.Rmd"
  ),
  role = c(
    "wrangling",
    "project setup",
    "data audit and QC",
    "endpoint registry and assumptions",
    "endpoint-level ethanol models",
    "battery-context analysis",
    "single full-battery PCA",
    "concordance and robustness",
    "reporting tables and figures",
    "statistical methods text",
    "reproducibility audit",
    "manuscript results and discussion",
    "submission package audit"
  )
) %>%
  mutate(
    full_path = file.path(paths$root, expected_file),
    exists = file.exists(full_path),
    status = case_when(
      exists ~ "present",
      order %in% c("10", "11") ~ "not_yet_expected",
      TRUE ~ "missing"
    ),
    size_bytes = purrr::map_dbl(full_path, safe_file_size),
    modified_time = purrr::map_chr(full_path, ~ as.character(safe_mtime(.x))),
    md5 = purrr::map_chr(full_path, safe_md5)
  )

path_notebook_inventory <- write_repro_table(expected_notebooks, "09_notebook_inventory.csv")

expected_notebooks
```

# 09.4 Expected output audit

```{r expected-output-audit}
expected_outputs <- bind_rows(
  audit_file_presence(
    "BT1 audited dataset",
    c(
      file.path(paths$data_processed, "bt1_audited.csv"),
      file.path(paths$data_processed, "BT1_audited.csv"),
      file.path(paths$data_processed, "battery_suite_1_audited.csv")
    ),
    required = TRUE,
    source_notebook = "01_data_audit_and_qc.Rmd"
  ),
  audit_file_presence(
    "BT2 audited dataset",
    c(
      file.path(paths$data_processed, "bt2_audited.csv"),
      file.path(paths$data_processed, "BT2_audited.csv"),
      file.path(paths$data_processed, "battery_suite_2_audited.csv")
    ),
    required = TRUE,
    source_notebook = "01_data_audit_and_qc.Rmd"
  ),
  audit_file_presence(
    "BT3 audited dataset",
    c(
      file.path(paths$data_processed, "bt3_audited.csv"),
      file.path(paths$data_processed, "BT3_audited.csv"),
      file.path(paths$data_processed, "battery_suite_3_audited.csv")
    ),
    required = TRUE,
    source_notebook = "01_data_audit_and_qc.Rmd"
  ),
  audit_file_presence(
    "Endpoint registry",
    c(
      file.path(paths$output_tables, "02_endpoint_registry.csv"),
      file.path(paths$output_tables, "00_endpoint_registry.csv"),
      file.path(paths$output_logs, "02_endpoint_registry.csv"),
      file.path(paths$output_logs, "00_endpoint_registry.csv")
    ),
    required = TRUE,
    source_notebook = "02_endpoint_hierarchy_and_assumptions.Rmd"
  ),
  audit_file_presence(
    "Model readiness",
    c(
      file.path(paths$output_tables, "02_model_readiness.csv"),
      file.path(paths$output_tables, "02_model_decision_matrix.csv"),
      file.path(paths$output_logs, "02_model_readiness.csv")
    ),
    required = TRUE,
    source_notebook = "02_endpoint_hierarchy_and_assumptions.Rmd"
  ),
  audit_file_presence(
    "Endpoint-level ethanol model results",
    c(
      file.path(paths$output_tables, "03_ethanol_factorial_results.csv"),
      file.path(paths$output_tables, "03_factorial_results.csv"),
      file.path(paths$output_tables, "03_model_results.csv"),
      file.path(paths$output_tables, "03_endpoint_model_results.csv")
    ),
    required = TRUE,
    source_notebook = "03_ethanol_factorial_models.Rmd"
  ),
  audit_file_presence(
    "Battery-context summary",
    c(
      file.path(paths$output_tables, "04_battery_context_summary.csv"),
      file.path(paths$output_tables, "04_battery_method_summary.csv"),
      file.path(paths$output_tables, "04_battery_comparison_summary.csv"),
      file.path(paths$output_tables, "04_battery_results.csv")
    ),
    required = TRUE,
    source_notebook = "04_battery_effect_analysis.Rmd"
  ),
  audit_file_presence(
    "PCA variance explained",
    c(
      file.path(paths$output_tables, "05_pca_variance_explained.csv"),
      file.path(paths$output_tables, "05_pca_eigenvalues.csv"),
      file.path(paths$output_tables, "06_pca_variance_explained.csv")
    ),
    required = TRUE,
    source_notebook = "05_multivariate_pca_structure.Rmd"
  ),
  audit_file_presence(
    "PCA loadings",
    c(
      file.path(paths$output_tables, "05_pca_loadings.csv"),
      file.path(paths$output_tables, "05_pca_variable_loadings.csv"),
      file.path(paths$output_tables, "06_pca_loadings_reconstructed.csv")
    ),
    required = TRUE,
    source_notebook = "05_multivariate_pca_structure.Rmd"
  ),
  audit_file_presence(
    "PCA scores",
    c(
      file.path(paths$output_tables, "05_pca_scores.csv"),
      file.path(paths$output_tables, "05_pca_individual_scores.csv"),
      file.path(paths$output_tables, "06_pca_scores_reconstructed.csv")
    ),
    required = TRUE,
    source_notebook = "05_multivariate_pca_structure.Rmd"
  ),
  audit_file_presence(
    "Notebook 06 decision register",
    c(
      file.path(paths$output_tables, "06_decision_register.csv"),
      file.path(paths$output_logs, "06_decision_register.csv")
    ),
    required = TRUE,
    source_notebook = "06_sensitivity_and_robustness.Rmd"
  ),
  audit_file_presence(
    "Notebook 06 critical flags",
    c(
      file.path(paths$output_tables, "06_critical_flags.csv"),
      file.path(paths$output_logs, "06_critical_flags.csv")
    ),
    required = TRUE,
    source_notebook = "06_sensitivity_and_robustness.Rmd"
  ),
  audit_file_presence(
    "Notebook 07 reporting manifest",
    c(file.path(paths$reporting_logs, "07_reporting_manifest.csv")),
    required = TRUE,
    source_notebook = "07_reporting_tables_and_figures.Rmd"
  ),
  audit_file_presence(
    "Notebook 08 methods draft",
    c(file.path(paths$reporting_text, "08_statistical_methods_draft.md")),
    required = TRUE,
    source_notebook = "08_statistical_methods_text.Rmd"
  ),
  audit_file_presence(
    "Notebook 08 methods source map",
    c(file.path(paths$reporting_logs, "08_methods_source_map.csv")),
    required = TRUE,
    source_notebook = "08_statistical_methods_text.Rmd"
  )
)

expected_output_audit <- expected_outputs %>%
  mutate(
    audit_status = case_when(
      status == "present" ~ "pass",
      status == "missing_required" ~ "fail",
      TRUE ~ "partial"
    )
  )

path_expected_output_audit <- write_repro_table(expected_output_audit, "09_expected_output_audit.csv")

expected_output_audit
```

# 09.5 Package versions and session information

```{r package-versions-session-info}
installed_pkgs <- utils::installed.packages()

packages_to_report <- unique(c(
  "base", "stats", "utils",
  "dplyr", "tidyr", "readr", "stringr", "purrr", "tibble",
  "ggplot2", "broom", "rlang", "car", "ARTool", "emmeans",
  "boot", "factoextra", "Hmisc", "corrplot", "knitr", "rmarkdown"
))

package_versions <- tibble(package = packages_to_report) %>%
  mutate(
    installed = package %in% rownames(installed_pkgs) | package %in% c("base", "stats", "utils"),
    version = purrr::map_chr(package, function(pkg) {
      if (pkg %in% c("base", "stats", "utils")) {
        return(as.character(utils::packageVersion(pkg)))
      }
      if (pkg %in% rownames(installed_pkgs)) {
        return(as.character(installed_pkgs[pkg, "Version"]))
      }
      NA_character_
    })
  ) %>%
  arrange(package)

path_package_versions <- write_repro_table(package_versions, "09_package_versions.csv")

session_info_lines <- capture.output(sessionInfo())
path_session_info <- write_repro_text(session_info_lines, "09_session_info.txt")

package_versions
```

# 09.6 Reporting bundle audit

```{r reporting-bundle-audit}
reporting_manifest <- read_optional_csv(
  c(file.path(paths$reporting_logs, "07_reporting_manifest.csv")),
  "Notebook 07 reporting manifest"
)

reporting_bundle_index <- read_optional_csv(
  c(file.path(paths$reporting_logs, "07_reporting_bundle_index.csv")),
  "Notebook 07 reporting bundle index"
)

required_reporting_files <- c(
  "tables/Table_01_dataset_qc_summary.csv",
  "tables/Table_02_endpoint_registry.csv",
  "tables/Table_03_descriptive_statistics.csv",
  "tables/Table_04_ethanol_endpoint_models.csv",
  "tables/Table_05_battery_context_summary.csv",
  "tables/Table_06_pca_summary.csv",
  "tables/Table_07_robustness_summary.csv",
  "logs/07_reporting_manifest.csv"
)

reporting_bundle_audit <- tibble(
  reporting_file = required_reporting_files,
  full_path = file.path(paths$reporting, reporting_file),
  exists = file.exists(full_path),
  status = ifelse(exists, "present", "missing"),
  size_bytes = purrr::map_dbl(full_path, safe_file_size),
  modified_time = purrr::map_chr(full_path, ~ as.character(safe_mtime(.x))),
  md5 = purrr::map_chr(full_path, safe_md5)
)

if (!is.null(reporting_manifest) && "status" %in% names(reporting_manifest)) {
  reporting_manifest_summary <- reporting_manifest %>%
    count(status, name = "n_outputs")
} else {
  reporting_manifest_summary <- tibble(status = "manifest_missing", n_outputs = NA_integer_)
}

path_reporting_bundle_audit <- write_repro_table(reporting_bundle_audit, "09_reporting_bundle_audit.csv")
path_reporting_manifest_summary <- write_repro_table(reporting_manifest_summary, "09_reporting_manifest_summary.csv")

reporting_bundle_audit
```

# 09.7 Methods text audit

```{r methods-text-audit}
methods_text_path <- file.path(paths$reporting_text, "08_statistical_methods_draft.md")
methods_text <- if (file.exists(methods_text_path)) readLines(methods_text_path, warn = FALSE) else NULL

forbidden_patterns <- c(
  "thesis-style language" = "\\bthesis-style\\b|\\bthesis style\\b",
  "primary/secondary endpoint hierarchy" = "\\bprimary\\b|\\bsecondary\\b",
  "PCA as proof" = "PCA\\s+(proved|proves|demonstrated|demonstrates)\\s+ethanol|PCA\\s+(tested|tests)\\s+ethanol",
  "forced PC labels" = "PC1\\s*=|PC2\\s*=|PC3\\s*=",
  "no-difference wording" = "\\bno difference\\b|\\bno significant difference\\b"
)

methods_forbidden_terms <- extract_forbidden_terms(methods_text, forbidden_patterns) %>%
  mutate(
    audit_status = case_when(
      is.na(count) ~ "not_checked",
      count == 0 ~ "pass",
      pattern == "primary/secondary endpoint hierarchy" & count > 0 ~ "review",
      TRUE ~ "flag"
    ),
    note = case_when(
      pattern == "primary/secondary endpoint hierarchy" ~ "Review manually: 'primary' may appear in phrases such as primary inference boundary, but endpoint hierarchy wording should be avoided.",
      TRUE ~ ""
    )
  )

expected_methods_files <- tibble(
  file = c(
    "08_statistical_methods_draft.md",
    "08_methods_data_preparation.md",
    "08_methods_endpoint_registry.md",
    "08_methods_endpoint_models.md",
    "08_methods_battery_context.md",
    "08_methods_pca.md",
    "08_methods_robustness.md"
  ),
  full_path = file.path(paths$reporting_text, file),
  exists = file.exists(full_path),
  size_bytes = purrr::map_dbl(full_path, safe_file_size),
  modified_time = purrr::map_chr(full_path, ~ as.character(safe_mtime(.x))),
  status = ifelse(exists, "present", "missing")
)

methods_text_audit <- bind_rows(
  expected_methods_files %>%
    transmute(check_type = "methods_file_presence", item = file, status, details = full_path),
  methods_forbidden_terms %>%
    transmute(check_type = "methods_wording", item = pattern, status = audit_status, details = paste0("count=", count, "; examples=", examples, "; ", note))
)

path_methods_text_audit <- write_repro_table(methods_text_audit, "09_methods_text_audit.csv")

methods_text_audit
```

# 09.8 Claim-source map

```{r claim-source-map}
claim_source_map <- tibble(
  claim_type = c(
    "Data preparation and QC",
    "Endpoint registry",
    "Endpoint ethanol effect",
    "Battery-context effect",
    "PCA structure",
    "Condition factor inclusion",
    "Robustness classification",
    "Reporting table or figure",
    "Methods text"
  ),
  allowed_source = c(
    "00_data_wrangling.ipynb; 01_data_audit_and_qc.Rmd; QC logs",
    "00_project_setup.Rmd; 02_endpoint_hierarchy_and_assumptions.Rmd; endpoint registry CSV",
    "03_ethanol_factorial_models.Rmd; model result tables; effect-size/contrast tables",
    "04_battery_effect_analysis.Rmd; battery-comparison tables; Notebook 06 wording audit",
    "05_multivariate_pca_structure.Rmd; PCA variance/loadings/scores; Notebook 06 PCA audit",
    "Condition-factor audit from Notebook 05 or 06",
    "06_sensitivity_and_robustness.Rmd; decision register; critical flags; stability tables",
    "07_reporting_tables_and_figures.Rmd; reporting manifest",
    "08_statistical_methods_text.Rmd; methods source map"
  ),
  allowed_wording_boundary = c(
    "Describe transformations, exclusions, derived variables, and data usability only.",
    "Describe endpoint roles without primary/secondary hierarchy.",
    "Claim endpoint-level treatment, exposure, or treatment-by-exposure effects only when supported by endpoint models.",
    "Claim battery-context differences only where detected; otherwise use inconclusive unless formal equivalence was supported.",
    "Describe multivariate organization and loading-supported component structure; do not claim PCA proves ethanol effects.",
    "Report as condition factor (K) only if formula, units, range, and availability audit passed.",
    "Use robust, partially robust, fragile, descriptive, or unresolved according to Notebook 06 outputs.",
    "Tables and figures summarize saved outputs only.",
    "Methods text summarizes completed decisions only; no new analysis decisions."
  ),
  required_output_check = c(
    "09_expected_output_audit: BT1/BT2/BT3 audited datasets",
    "09_expected_output_audit: endpoint registry and model readiness",
    "09_expected_output_audit: ethanol model results",
    "09_expected_output_audit: battery-context summary",
    "09_expected_output_audit: PCA variance, loadings, and scores",
    "09_expected_output_audit: condition-factor audit where available",
    "09_expected_output_audit: decision register and critical flags",
    "09_reporting_bundle_audit; 07_reporting_manifest",
    "09_methods_text_audit; 08_methods_source_map"
  )
)

path_claim_source_map <- write_repro_table(claim_source_map, "09_claim_source_map.csv")

claim_source_map
```

# 09.9 README-ready execution summary

```{r readme-execution-summary}
failed_outputs <- expected_output_audit %>% filter(audit_status == "fail")
missing_notebooks <- expected_notebooks %>% filter(status == "missing")
missing_reporting <- reporting_bundle_audit %>% filter(status == "missing")
flagged_methods <- methods_text_audit %>% filter(status %in% c("flag", "review"))

readme_summary <- c(
  "# Reproducibility and execution summary",
  "",
  paste0("Generated on: ", Sys.Date()),
  "",
  "## Project execution order",
  "",
  "1. `python/00_data_wrangling.ipynb`",
  "2. `notebooks/00_project_setup.Rmd`",
  "3. `notebooks/01_data_audit_and_qc.Rmd`",
  "4. `notebooks/02_endpoint_hierarchy_and_assumptions.Rmd`",
  "5. `notebooks/03_ethanol_factorial_models.Rmd`",
  "6. `notebooks/04_battery_effect_analysis.Rmd`",
  "7. `notebooks/05_multivariate_pca_structure.Rmd`",
  "8. `notebooks/06_sensitivity_and_robustness.Rmd`",
  "9. `notebooks/07_reporting_tables_and_figures.Rmd`",
  "10. `notebooks/08_statistical_methods_text.Rmd`",
  "11. `notebooks/09_reproducibility_audit.Rmd`",
  "",
  "## Locked analysis rules",
  "",
  "- Endpoint roles are controlled by the endpoint registry, not by a primary/secondary endpoint hierarchy.",
  "- Endpoint-level ethanol effects are tested with treatment, exposure, and treatment-by-exposure models.",
  "- Battery-context comparisons are separate from endpoint-level ethanol models.",
  "- One single BT1 full-battery PCA is used for multivariate visualization.",
  "- PCA variables are centered and scaled before PCA.",
  "- PCA component labels are assigned from loadings.",
  "- PCA does not replace endpoint-level ethanol models.",
  "- Condition factor is reported as condition factor (K) only when the audit supports formula and value consistency.",
  "",
  "## Audit status",
  "",
  paste0("- Missing required notebook files: ", nrow(missing_notebooks)),
  paste0("- Missing required outputs: ", nrow(failed_outputs)),
  paste0("- Missing reporting files: ", nrow(missing_reporting)),
  paste0("- Methods wording flags/reviews: ", nrow(flagged_methods)),
  "",
  "## Reproducibility outputs",
  "",
  "- `outputs/reproducibility/09_file_inventory.csv`",
  "- `outputs/reproducibility/09_notebook_inventory.csv`",
  "- `outputs/reproducibility/09_expected_output_audit.csv`",
  "- `outputs/reproducibility/09_package_versions.csv`",
  "- `outputs/reproducibility/09_session_info.txt`",
  "- `outputs/reproducibility/09_reporting_bundle_audit.csv`",
  "- `outputs/reproducibility/09_methods_text_audit.csv`",
  "- `outputs/reproducibility/09_claim_source_map.csv`",
  "",
  "## Next step",
  "",
  "If required outputs are missing, rerender or patch the corresponding upstream notebook before manuscript writing. If no required outputs are missing, proceed to Notebook 10."
)

path_readme_summary <- write_repro_text(readme_summary, "09_readme_execution_summary.md")

cat(paste(readme_summary, collapse = "\n"))
```

# 09.10 Reproducibility manifest

```{r reproducibility-manifest}
repro_manifest <- tibble(
  output_name = c(
    "09_file_inventory.csv",
    "09_notebook_inventory.csv",
    "09_expected_output_audit.csv",
    "09_package_versions.csv",
    "09_session_info.txt",
    "09_reporting_bundle_audit.csv",
    "09_reporting_manifest_summary.csv",
    "09_methods_text_audit.csv",
    "09_claim_source_map.csv",
    "09_readme_execution_summary.md"
  ),
  output_type = c(
    "csv",
    "csv",
    "csv",
    "csv",
    "text",
    "csv",
    "csv",
    "csv",
    "csv",
    "markdown"
  ),
  path = c(
    path_file_inventory,
    path_notebook_inventory,
    path_expected_output_audit,
    path_package_versions,
    path_session_info,
    path_reporting_bundle_audit,
    path_reporting_manifest_summary,
    path_methods_text_audit,
    path_claim_source_map,
    path_readme_summary
  ),
  status = ifelse(file.exists(path), "generated", "missing"),
  notes = c(
    "Inventory of files in analysis-relevant project folders.",
    "Presence and checksum audit for expected notebooks.",
    "Presence and checksum audit for expected upstream outputs.",
    "Installed package versions for analysis-relevant packages.",
    "Full R session information.",
    "Presence audit for reporting bundle files.",
    "Summary of generated/skipped Notebook 07 outputs.",
    "Presence and wording audit for Methods text files.",
    "Claim-source map for manuscript traceability.",
    "README-ready execution and audit summary."
  )
)

path_repro_manifest <- write_repro_table(repro_manifest, "09_reproducibility_manifest.csv")

repro_manifest
```

# 09.11 Final reproducibility status

```{r final-reproducibility-status}
final_status <- tibble(
  audit_layer = c(
    "Notebook inventory",
    "Expected output audit",
    "Reporting bundle audit",
    "Methods text audit",
    "Package/session audit",
    "Claim-source map"
  ),
  status = c(
    ifelse(nrow(missing_notebooks) == 0, "pass", "flag"),
    ifelse(nrow(failed_outputs) == 0, "pass", "flag"),
    ifelse(nrow(missing_reporting) == 0, "pass", "flag"),
    ifelse(any(methods_text_audit$status == "flag"), "flag", ifelse(any(methods_text_audit$status == "review"), "review", "pass")),
    "generated",
    "generated"
  ),
  details = c(
    paste0(nrow(missing_notebooks), " missing expected notebook files."),
    paste0(nrow(failed_outputs), " missing required upstream outputs."),
    paste0(nrow(missing_reporting), " missing reporting-bundle files."),
    paste0(sum(methods_text_audit$status == "flag", na.rm = TRUE), " flags; ", sum(methods_text_audit$status == "review", na.rm = TRUE), " review items."),
    "Package versions and session info exported.",
    "Claim-source map exported."
  )
)

path_final_status <- write_repro_table(final_status, "09_final_reproducibility_status.csv")

final_status
```

# Completion statement

Notebook 09 is complete when the reproducibility folder contains the file inventory, notebook inventory, expected-output audit, package versions, session information, reporting-bundle audit, methods-text audit, claim-source map, README-ready execution summary, and reproducibility manifest.

If any required output is missing, rerender or patch the source notebook listed in `09_expected_output_audit.csv`. If only optional or future notebooks are missing, proceed according to the execution order.
