---
title: "11 Submission Package Audit"
author: "Caique Ortiz Colombo Arnoldi Fabris"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
---

# Purpose

This notebook is the final audit layer before manuscript submission or external editing.

It does not introduce new analyses. It does not change endpoint roles, model choices, PCA variables, PCA interpretation, robustness classifications, figure numbering, or manuscript claims. It checks whether the manuscript package is internally consistent, traceable, and safe from overclaiming.

The audit covers:

- manuscript-output presence;
- figure and table caption consistency;
- claim-source traceability;
- PCA wording;
- battery-comparison wording;
- endpoint-registry wording;
- condition-factor wording;
- robustness and reproducibility status;
- supplement readiness;
- final package checklist.

# Notebook contract

Notebook 11 is complete when it exports:

- `outputs/submission_package/11_submission_file_audit.csv`
- `outputs/submission_package/11_claim_traceability_audit.csv`
- `outputs/submission_package/11_pca_wording_audit.csv`
- `outputs/submission_package/11_battery_wording_audit.csv`
- `outputs/submission_package/11_endpoint_registry_wording_audit.csv`
- `outputs/submission_package/11_condition_factor_wording_audit.csv`
- `outputs/submission_package/11_figure_table_audit.csv`
- `outputs/submission_package/11_supplement_readiness_checklist.csv`
- `outputs/submission_package/11_final_submission_checklist.csv`
- `outputs/submission_package/11_submission_package_manifest.csv`
- `outputs/submission_package/11_submission_audit_summary.md`

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c(
  "dplyr", "tidyr", "readr", "stringr", "purrr", "tibble", "rlang"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_packages) > 0) {
  stop(
    "Missing required packages: ",
    paste(missing_packages, collapse = ", "),
    ". Install them before rendering this notebook."
  )
}

library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(purrr)
library(tibble)
library(rlang)
```

# 11.1 Project paths and helpers

```{r paths-and-helpers}
find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)

  while (TRUE) {
    has_rproj <- length(list.files(current, pattern = "\\.Rproj$", full.names = TRUE)) > 0
    has_git <- dir.exists(file.path(current, ".git"))
    has_data <- dir.exists(file.path(current, "data_processed")) || dir.exists(file.path(current, "data_raw"))

    if (has_rproj || has_git || has_data) {
      return(current)
    }

    parent <- dirname(current)
    if (identical(parent, current)) {
      return(normalizePath(start, winslash = "/", mustWork = TRUE))
    }

    current <- parent
  }
}

project_root <- find_project_root()

paths <- list(
  root = project_root,
  data_processed = file.path(project_root, "data_processed"),
  docs = file.path(project_root, "docs"),
  notebooks = file.path(project_root, "notebooks"),
  outputs = file.path(project_root, "outputs"),
  output_tables = file.path(project_root, "outputs", "tables"),
  output_figures = file.path(project_root, "outputs", "figures"),
  output_logs = file.path(project_root, "outputs", "logs"),
  reporting = file.path(project_root, "outputs", "reporting"),
  reporting_tables = file.path(project_root, "outputs", "reporting", "tables"),
  reporting_figures = file.path(project_root, "outputs", "reporting", "figures"),
  reporting_logs = file.path(project_root, "outputs", "reporting", "logs"),
  reporting_text = file.path(project_root, "outputs", "reporting", "text"),
  reproducibility = file.path(project_root, "outputs", "reproducibility"),
  manuscript = file.path(project_root, "outputs", "manuscript"),
  submission = file.path(project_root, "outputs", "submission_package")
)

purrr::walk(
  paths[c("outputs", "submission")],
  dir.create,
  recursive = TRUE,
  showWarnings = FALSE
)

first_existing_path <- function(candidates) {
  existing <- candidates[file.exists(candidates)]
  if (length(existing) == 0) return(NA_character_)
  existing[[1]]
}

read_optional_csv <- function(candidates, label) {
  path <- first_existing_path(candidates)

  if (is.na(path)) {
    message("Missing optional input: ", label)
    return(NULL)
  }

  message("Loaded ", label, ": ", path)
  readr::read_csv(path, show_col_types = FALSE)
}

read_optional_text <- function(candidates, label) {
  path <- first_existing_path(candidates)

  if (is.na(path)) {
    message("Missing optional text: ", label)
    return(NULL)
  }

  message("Loaded ", label, ": ", path)
  readLines(path, warn = FALSE)
}

write_submission_table <- function(df, filename) {
  out <- file.path(paths$submission, filename)
  readr::write_csv(df, out)
  out
}

write_submission_text <- function(lines, filename) {
  out <- file.path(paths$submission, filename)
  writeLines(lines, out)
  out
}

safe_file_size <- function(path) {
  if (!file.exists(path)) return(NA_real_)
  file.info(path)$size
}

safe_mtime <- function(path) {
  if (!file.exists(path)) return(as.POSIXct(NA))
  file.info(path)$mtime
}

safe_md5 <- function(path) {
  if (!file.exists(path) || dir.exists(path)) return(NA_character_)
  tryCatch(as.character(tools::md5sum(path)), error = function(e) NA_character_)
}

audit_file <- function(label, candidates, required = TRUE, file_role = NA_character_) {
  path <- first_existing_path(candidates)

  tibble(
    label = label,
    file_role = file_role,
    required = required,
    status = ifelse(!is.na(path), "present", ifelse(required, "missing_required", "missing_optional")),
    path = path,
    size_bytes = ifelse(!is.na(path), safe_file_size(path), NA_real_),
    modified_time = ifelse(!is.na(path), as.character(safe_mtime(path)), NA_character_),
    md5 = ifelse(!is.na(path), safe_md5(path), NA_character_),
    candidates_checked = paste(candidates, collapse = " | ")
  )
}

collapse_text <- function(lines) {
  if (is.null(lines)) return("")
  paste(lines, collapse = "\n")
}

count_pattern <- function(text, pattern) {
  if (is.null(text) || text == "") return(NA_integer_)
  length(stringr::str_extract_all(text, pattern, simplify = FALSE)[[1]])
}

examples_pattern <- function(text, pattern, n = 5) {
  if (is.null(text) || text == "") return(NA_character_)
  matches <- unique(stringr::str_extract_all(text, pattern, simplify = FALSE)[[1]])
  matches <- matches[matches != ""]
  if (length(matches) == 0) return(NA_character_)
  paste(head(matches, n), collapse = " | ")
}

scan_patterns <- function(text, pattern_table) {
  pattern_table %>%
    rowwise() %>%
    mutate(
      count = count_pattern(text, regex),
      examples = examples_pattern(text, regex),
      audit_status = case_when(
        is.na(count) ~ "not_checked",
        count == 0 ~ "pass",
        expected == "forbidden" ~ "flag",
        expected == "review" ~ "review",
        TRUE ~ "present"
      )
    ) %>%
    ungroup()
}

safe_has_col <- function(df, candidates) {
  if (is.null(df)) return(NA_character_)

  exact <- candidates[candidates %in% names(df)]
  if (length(exact) > 0) return(exact[[1]])

  lower_map <- tibble(actual = names(df), lower = stringr::str_to_lower(names(df)))
  lower_candidates <- stringr::str_to_lower(candidates)
  lower_hit <- lower_map %>% filter(lower %in% lower_candidates)

  if (nrow(lower_hit) > 0) return(lower_hit$actual[[1]])
  NA_character_
}

add_manifest <- function(manifest, output_name, output_type, status, path, notes = "") {
  bind_rows(
    manifest,
    tibble(
      output_name = output_name,
      output_type = output_type,
      status = status,
      path = path,
      notes = notes
    )
  )
}

cat("Project root:", project_root, "\n")
```

# 11.2 Load package inputs

```{r load-package-inputs}
full_manuscript <- read_optional_text(
  c(file.path(paths$manuscript, "10_full_manuscript_draft.md")),
  "full manuscript draft"
)

results_draft <- read_optional_text(
  c(file.path(paths$manuscript, "10_results_draft.md")),
  "results draft"
)

discussion_draft <- read_optional_text(
  c(file.path(paths$manuscript, "10_discussion_draft.md")),
  "discussion draft"
)

limitations_draft <- read_optional_text(
  c(file.path(paths$manuscript, "10_limitations_draft.md")),
  "limitations draft"
)

abstract_draft <- read_optional_text(
  c(file.path(paths$manuscript, "10_abstract_draft.md")),
  "abstract draft"
)

figure_captions <- read_optional_text(
  c(file.path(paths$manuscript, "10_figure_captions.md")),
  "figure captions"
)

table_captions <- read_optional_text(
  c(file.path(paths$manuscript, "10_table_captions.md")),
  "table captions"
)

claim_traceability <- read_optional_csv(
  c(
    file.path(paths$manuscript, "10_claim_traceability_table.csv"),
    file.path(paths$reproducibility, "09_claim_source_map.csv")
  ),
  "claim traceability table"
)

manuscript_manifest <- read_optional_csv(
  c(file.path(paths$manuscript, "10_manuscript_manifest.csv")),
  "manuscript manifest"
)

reporting_manifest <- read_optional_csv(
  c(file.path(paths$reporting_logs, "07_reporting_manifest.csv")),
  "reporting manifest"
)

repro_manifest <- read_optional_csv(
  c(file.path(paths$reproducibility, "09_reproducibility_manifest.csv")),
  "reproducibility manifest"
)

repro_status <- read_optional_csv(
  c(file.path(paths$reproducibility, "09_final_reproducibility_status.csv")),
  "final reproducibility status"
)

expected_output_audit <- read_optional_csv(
  c(file.path(paths$reproducibility, "09_expected_output_audit.csv")),
  "expected output audit"
)

methods_text_audit <- read_optional_csv(
  c(file.path(paths$reproducibility, "09_methods_text_audit.csv")),
  "methods text audit"
)

pca_summary <- read_optional_csv(
  c(file.path(paths$reporting_tables, "Table_06_pca_summary.csv")),
  "PCA summary table"
)

battery_summary <- read_optional_csv(
  c(file.path(paths$reporting_tables, "Table_05_battery_context_summary.csv")),
  "battery-context summary table"
)

endpoint_registry <- read_optional_csv(
  c(file.path(paths$reporting_tables, "Table_02_endpoint_registry.csv")),
  "endpoint registry table"
)

all_manuscript_text <- paste(
  collapse_text(full_manuscript),
  collapse_text(results_draft),
  collapse_text(discussion_draft),
  collapse_text(limitations_draft),
  collapse_text(abstract_draft),
  collapse_text(figure_captions),
  collapse_text(table_captions),
  sep = "\n"
)
```

# 11.3 Submission file audit

```{r submission-file-audit}
submission_file_audit <- bind_rows(
  audit_file(
    "Full manuscript draft",
    c(file.path(paths$manuscript, "10_full_manuscript_draft.md")),
    required = TRUE,
    file_role = "manuscript_text"
  ),
  audit_file(
    "Results draft",
    c(file.path(paths$manuscript, "10_results_draft.md")),
    required = TRUE,
    file_role = "manuscript_text"
  ),
  audit_file(
    "Discussion draft",
    c(file.path(paths$manuscript, "10_discussion_draft.md")),
    required = TRUE,
    file_role = "manuscript_text"
  ),
  audit_file(
    "Limitations draft",
    c(file.path(paths$manuscript, "10_limitations_draft.md")),
    required = TRUE,
    file_role = "manuscript_text"
  ),
  audit_file(
    "Conclusion draft",
    c(file.path(paths$manuscript, "10_conclusion_draft.md")),
    required = TRUE,
    file_role = "manuscript_text"
  ),
  audit_file(
    "Abstract draft",
    c(file.path(paths$manuscript, "10_abstract_draft.md")),
    required = TRUE,
    file_role = "manuscript_text"
  ),
  audit_file(
    "Figure captions",
    c(file.path(paths$manuscript, "10_figure_captions.md")),
    required = TRUE,
    file_role = "captions"
  ),
  audit_file(
    "Table captions",
    c(file.path(paths$manuscript, "10_table_captions.md")),
    required = TRUE,
    file_role = "captions"
  ),
  audit_file(
    "Claim traceability table",
    c(file.path(paths$manuscript, "10_claim_traceability_table.csv")),
    required = TRUE,
    file_role = "traceability"
  ),
  audit_file(
    "Manuscript manifest",
    c(file.path(paths$manuscript, "10_manuscript_manifest.csv")),
    required = TRUE,
    file_role = "manifest"
  ),
  audit_file(
    "Reporting manifest",
    c(file.path(paths$reporting_logs, "07_reporting_manifest.csv")),
    required = TRUE,
    file_role = "manifest"
  ),
  audit_file(
    "Reproducibility final status",
    c(file.path(paths$reproducibility, "09_final_reproducibility_status.csv")),
    required = TRUE,
    file_role = "reproducibility"
  )
) %>%
  mutate(
    audit_status = case_when(
      status == "present" ~ "pass",
      status == "missing_required" ~ "fail",
      TRUE ~ "review"
    )
  )

path_submission_file_audit <- write_submission_table(submission_file_audit, "11_submission_file_audit.csv")

submission_file_audit
```

# 11.4 Claim traceability audit

```{r claim-traceability-audit}
if (!is.null(claim_traceability)) {
  required_cols <- c("claim_boundary", "source_output")
  present_required_cols <- required_cols[required_cols %in% names(claim_traceability)]

  claim_traceability_audit <- tibble(
    check = c(
      "claim_traceability_table_present",
      "claim_boundary_column_present",
      "source_output_column_present",
      "traceability_rows_present"
    ),
    status = c(
      "pass",
      ifelse("claim_boundary" %in% names(claim_traceability), "pass", "flag"),
      ifelse("source_output" %in% names(claim_traceability), "pass", "flag"),
      ifelse(nrow(claim_traceability) > 0, "pass", "flag")
    ),
    details = c(
      paste0("Rows: ", nrow(claim_traceability), "; columns: ", ncol(claim_traceability)),
      paste0("Columns present: ", paste(names(claim_traceability), collapse = ", ")),
      paste0("Columns present: ", paste(names(claim_traceability), collapse = ", ")),
      paste0("Rows: ", nrow(claim_traceability))
    )
  )
} else {
  claim_traceability_audit <- tibble(
    check = "claim_traceability_table_present",
    status = "flag",
    details = "No claim traceability table found."
  )
}

# Check whether the full manuscript contains untraceable-language markers.
claim_overreach_patterns <- tibble(
  pattern = c(
    "causal_overclaim",
    "PCA_as_inference",
    "global_behavioral_score",
    "unsupported_no_difference",
    "unqualified_robustness"
  ),
  regex = c(
    "\\bcaused\\b|\\bcauses\\b|\\bcausal\\b",
    "PCA\\s+(proved|proves|demonstrated|demonstrates|tested|tests)\\s+ethanol|PCA\\s+showed\\s+ethanol",
    "\\bglobal behavioral score\\b|\\boverall behavioral score\\b",
    "\\bno difference\\b|\\bno significant difference\\b",
    "\\bfully robust\\b|\\bdefinitive\\b|\\bconclusive\\b"
  ),
  expected = "forbidden"
)

claim_overreach_audit <- scan_patterns(all_manuscript_text, claim_overreach_patterns) %>%
  transmute(
    check = paste0("wording_", pattern),
    status = audit_status,
    details = paste0("count=", count, "; examples=", examples)
  )

claim_traceability_audit <- bind_rows(claim_traceability_audit, claim_overreach_audit)

path_claim_traceability_audit <- write_submission_table(claim_traceability_audit, "11_claim_traceability_audit.csv")

claim_traceability_audit
```

# 11.5 PCA wording audit

```{r pca-wording-audit}
pca_patterns <- tibble(
  pattern = c(
    "thesis_style_language",
    "behavioral_only_pca",
    "multiple_or_competing_pca",
    "PCA_proves_ethanol_effect",
    "forced_PC_label_syntax",
    "missing_z_transformation_term",
    "loading_based_interpretation_present",
    "visualization_role_present"
  ),
  regex = c(
    "\\bthesis-style\\b|\\bthesis style\\b",
    "\\bbehavioral-only PCA\\b|\\bbehavioural-only PCA\\b",
    "\\bcompeting PCA\\b|\\bsecondary PCA\\b|\\badditional PCA\\b",
    "PCA\\s+(proved|proves|demonstrated|demonstrates|tested|tests)\\s+ethanol|PCA\\s+showed\\s+ethanol",
    "PC1\\s*=|PC2\\s*=|PC3\\s*=",
    "z-transformed|centered and scaled|centred and scaled|centering and scaling",
    "loading|loadings|charges|dominant endpoint contributions",
    "visualization|visualisation|exploratory"
  ),
  expected = c(
    "forbidden",
    "forbidden",
    "forbidden",
    "forbidden",
    "forbidden",
    "required",
    "required",
    "required"
  )
)

pca_wording_raw <- scan_patterns(all_manuscript_text, pca_patterns)

pca_wording_audit <- pca_wording_raw %>%
  mutate(
    audit_status = case_when(
      expected == "forbidden" & count == 0 ~ "pass",
      expected == "forbidden" & count > 0 ~ "flag",
      expected == "required" & count > 0 ~ "pass",
      expected == "required" & count == 0 ~ "flag",
      TRUE ~ audit_status
    ),
    details = paste0("count=", count, "; examples=", examples)
  ) %>%
  select(pattern, expected, audit_status, details)

path_pca_wording_audit <- write_submission_table(pca_wording_audit, "11_pca_wording_audit.csv")

pca_wording_audit
```

# 11.6 Battery wording audit

```{r battery-wording-audit}
battery_patterns <- tibble(
  pattern = c(
    "no_difference_wording",
    "equivalence_without_formal_support",
    "battery_context_distinction_present",
    "inconclusive_wording_present"
  ),
  regex = c(
    "\\bno difference\\b|\\bno significant difference\\b",
    "\\bequivalent\\b|\\bequivalence\\b",
    "battery-context|battery context|battery comparison|battery-effect|battery effect",
    "\\binconclusive\\b"
  ),
  expected = c(
    "forbidden",
    "review",
    "required",
    "required"
  )
)

battery_wording_raw <- scan_patterns(all_manuscript_text, battery_patterns)

battery_wording_audit <- battery_wording_raw %>%
  mutate(
    audit_status = case_when(
      pattern == "equivalence_without_formal_support" & count > 0 ~ "review",
      expected == "forbidden" & count == 0 ~ "pass",
      expected == "forbidden" & count > 0 ~ "flag",
      expected == "required" & count > 0 ~ "pass",
      expected == "required" & count == 0 ~ "review",
      TRUE ~ audit_status
    ),
    details = paste0("count=", count, "; examples=", examples)
  ) %>%
  select(pattern, expected, audit_status, details)

path_battery_wording_audit <- write_submission_table(battery_wording_audit, "11_battery_wording_audit.csv")

battery_wording_audit
```

# 11.7 Endpoint-registry wording audit

```{r endpoint-registry-wording-audit}
endpoint_patterns <- tibble(
  pattern = c(
    "primary_secondary_language",
    "endpoint_registry_present",
    "endpoint_roles_present",
    "endpoint_hierarchy_language"
  ),
  regex = c(
    "\\bprimary endpoint\\b|\\bsecondary endpoint\\b|\\bprimary/secondary\\b",
    "endpoint registry|endpoint-registry",
    "PCA inclusion|PCA-included|model-tested|battery-comparable|covariate|QC",
    "endpoint hierarchy|hierarchy"
  ),
  expected = c(
    "forbidden",
    "required",
    "required",
    "review"
  )
)

endpoint_wording_raw <- scan_patterns(all_manuscript_text, endpoint_patterns)

endpoint_registry_wording_audit <- endpoint_wording_raw %>%
  mutate(
    audit_status = case_when(
      expected == "forbidden" & count == 0 ~ "pass",
      expected == "forbidden" & count > 0 ~ "flag",
      expected == "required" & count > 0 ~ "pass",
      expected == "required" & count == 0 ~ "review",
      expected == "review" & count > 0 ~ "review",
      expected == "review" & count == 0 ~ "pass",
      TRUE ~ audit_status
    ),
    details = paste0("count=", count, "; examples=", examples)
  ) %>%
  select(pattern, expected, audit_status, details)

path_endpoint_registry_wording_audit <- write_submission_table(endpoint_registry_wording_audit, "11_endpoint_registry_wording_audit.csv")

endpoint_registry_wording_audit
```

# 11.8 Condition-factor wording audit

```{r condition-factor-wording-audit}
condition_factor_patterns <- tibble(
  pattern = c(
    "condition_factor_K_present",
    "Kc_Kf_alias_overexplained",
    "fulton_type_optional",
    "condition_factor_included_without_audit"
  ),
  regex = c(
    "condition factor \\(K\\)|condition factor|Fulton",
    "Kc.*Kf.*same|Kf.*Kc.*same|Kc/Kf",
    "Fulton-type condition factor|Fulton",
    "condition factor.*included|included.*condition factor"
  ),
  expected = c(
    "required",
    "review",
    "review",
    "review"
  )
)

condition_factor_raw <- scan_patterns(all_manuscript_text, condition_factor_patterns)

condition_factor_wording_audit <- condition_factor_raw %>%
  mutate(
    audit_status = case_when(
      pattern == "condition_factor_K_present" & count > 0 ~ "pass",
      pattern == "condition_factor_K_present" & count == 0 ~ "review",
      pattern == "condition_factor_included_without_audit" & count > 0 ~ "review",
      expected == "review" & count > 0 ~ "review",
      expected == "review" & count == 0 ~ "pass",
      TRUE ~ audit_status
    ),
    details = paste0("count=", count, "; examples=", examples)
  ) %>%
  select(pattern, expected, audit_status, details)

# Add direct audit-file check if condition factor summary exists.
condition_factor_source_check <- audit_file(
  "Condition-factor audit source",
  c(
    file.path(paths$output_tables, "06_condition_factor_audit.csv"),
    file.path(paths$output_logs, "06_condition_factor_audit.csv"),
    file.path(paths$output_tables, "05_condition_factor_audit.csv")
  ),
  required = FALSE,
  file_role = "condition_factor_audit"
) %>%
  transmute(
    pattern = "condition_factor_source_file",
    expected = "required_if_discussed",
    audit_status = case_when(
      status == "present" ~ "pass",
      count_pattern(all_manuscript_text, "condition factor|Fulton|Kc|Kf") > 0 ~ "flag",
      TRUE ~ "pass"
    ),
    details = paste0("status=", status, "; path=", path)
  )

condition_factor_wording_audit <- bind_rows(condition_factor_wording_audit, condition_factor_source_check)

path_condition_factor_wording_audit <- write_submission_table(condition_factor_wording_audit, "11_condition_factor_wording_audit.csv")

condition_factor_wording_audit
```

# 11.9 Figure and table audit

```{r figure-table-audit}
figure_caption_text <- collapse_text(figure_captions)
table_caption_text <- collapse_text(table_captions)

figure_numbers <- stringr::str_extract_all(figure_caption_text, "Figure\\s+\\d+", simplify = FALSE)[[1]]
table_numbers <- stringr::str_extract_all(table_caption_text, "Table\\s+S?\\d+|Table\\s+\\d+", simplify = FALSE)[[1]]

generated_figure_files <- if (!is.null(reporting_manifest) && all(c("output_type", "status") %in% names(reporting_manifest))) {
  reporting_manifest %>%
    filter(output_type == "figure", status == "generated")
} else {
  tibble()
}

generated_table_files <- if (!is.null(reporting_manifest) && all(c("output_type", "status") %in% names(reporting_manifest))) {
  reporting_manifest %>%
    filter(output_type == "table", status %in% c("generated", "generated_from_effect_table"))
} else {
  tibble()
}

figure_table_audit <- tibble(
  check = c(
    "figure_caption_file_present",
    "table_caption_file_present",
    "generated_figure_manifest_rows",
    "generated_table_manifest_rows",
    "figure_numbers_detected",
    "table_numbers_detected",
    "duplicate_figure_numbers",
    "duplicate_table_numbers"
  ),
  status = c(
    ifelse(is.null(figure_captions), "flag", "pass"),
    ifelse(is.null(table_captions), "flag", "pass"),
    ifelse(nrow(generated_figure_files) > 0, "pass", "review"),
    ifelse(nrow(generated_table_files) > 0, "pass", "review"),
    ifelse(length(figure_numbers) > 0, "pass", "review"),
    ifelse(length(table_numbers) > 0, "pass", "review"),
    ifelse(any(duplicated(figure_numbers)), "flag", "pass"),
    ifelse(any(duplicated(table_numbers)), "flag", "pass")
  ),
  details = c(
    ifelse(is.null(figure_captions), "Figure captions missing.", "Figure captions loaded."),
    ifelse(is.null(table_captions), "Table captions missing.", "Table captions loaded."),
    paste0(nrow(generated_figure_files), " generated figure rows in reporting manifest."),
    paste0(nrow(generated_table_files), " generated table rows in reporting manifest."),
    paste(unique(figure_numbers), collapse = ", "),
    paste(unique(table_numbers), collapse = ", "),
    paste(figure_numbers[duplicated(figure_numbers)], collapse = ", "),
    paste(table_numbers[duplicated(table_numbers)], collapse = ", ")
  )
)

# Direct file existence check for generated reporting artifacts.
if (nrow(generated_figure_files) > 0 && "path" %in% names(generated_figure_files)) {
  figure_file_checks <- generated_figure_files %>%
    transmute(
      check = paste0("figure_file_exists_", output_name),
      status = ifelse(file.exists(path), "pass", "flag"),
      details = path
    )
} else {
  figure_file_checks <- tibble(check = character(), status = character(), details = character())
}

if (nrow(generated_table_files) > 0 && "path" %in% names(generated_table_files)) {
  table_file_checks <- generated_table_files %>%
    transmute(
      check = paste0("table_file_exists_", output_name),
      status = ifelse(file.exists(path), "pass", "flag"),
      details = path
    )
} else {
  table_file_checks <- tibble(check = character(), status = character(), details = character())
}

figure_table_audit <- bind_rows(figure_table_audit, figure_file_checks, table_file_checks)

path_figure_table_audit <- write_submission_table(figure_table_audit, "11_figure_table_audit.csv")

figure_table_audit
```

# 11.10 Supplement readiness checklist

```{r supplement-readiness-checklist}
supplement_items <- tibble(
  supplement_item = c(
    "Dataset inventory",
    "Endpoint registry",
    "Descriptive statistics",
    "Endpoint model results",
    "Endpoint effect estimates or contrasts",
    "Battery-context comparison summary",
    "PCA variable concordance",
    "PCA variance/loadings/scores",
    "Condition-factor audit",
    "Robustness summary",
    "Reproducibility file inventory",
    "Package versions and session info",
    "Claim-source map"
  ),
  source_file = c(
    file.path(paths$reporting_tables, "Table_01_dataset_qc_summary.csv"),
    file.path(paths$reporting_tables, "Table_02_endpoint_registry.csv"),
    file.path(paths$reporting_tables, "Table_03_descriptive_statistics.csv"),
    file.path(paths$reporting_tables, "Table_04_ethanol_endpoint_models.csv"),
    file.path(paths$reporting_tables, "Table_04b_ethanol_effect_estimates_or_contrasts.csv"),
    file.path(paths$reporting_tables, "Table_05_battery_context_summary.csv"),
    file.path(paths$output_tables, "06_pca_variable_concordance.csv"),
    file.path(paths$reporting_tables, "Table_06_pca_summary.csv"),
    first_existing_path(c(
      file.path(paths$output_tables, "06_condition_factor_audit.csv"),
      file.path(paths$output_logs, "06_condition_factor_audit.csv"),
      file.path(paths$output_tables, "05_condition_factor_audit.csv")
    )),
    file.path(paths$reporting_tables, "Table_07_robustness_summary.csv"),
    file.path(paths$reproducibility, "09_file_inventory.csv"),
    file.path(paths$reproducibility, "09_package_versions.csv"),
    file.path(paths$reproducibility, "09_claim_source_map.csv")
  )
) %>%
  mutate(
    exists = file.exists(source_file),
    readiness_status = ifelse(exists, "ready", "missing"),
    notes = case_when(
      supplement_item == "Condition-factor audit" & !exists ~ "Required if condition factor is discussed or included in PCA.",
      supplement_item == "Endpoint effect estimates or contrasts" & !exists ~ "Optional if no post hoc/effect table was generated.",
      TRUE ~ ""
    )
  )

path_supplement_readiness <- write_submission_table(supplement_items, "11_supplement_readiness_checklist.csv")

supplement_items
```

# 11.11 Final submission checklist

```{r final-submission-checklist}
all_audits <- list(
  submission_file_audit = submission_file_audit %>% transmute(audit = "submission_file_audit", item = label, status = audit_status),
  claim_traceability_audit = claim_traceability_audit %>% transmute(audit = "claim_traceability_audit", item = check, status),
  pca_wording_audit = pca_wording_audit %>% transmute(audit = "pca_wording_audit", item = pattern, status = audit_status),
  battery_wording_audit = battery_wording_audit %>% transmute(audit = "battery_wording_audit", item = pattern, status = audit_status),
  endpoint_registry_wording_audit = endpoint_registry_wording_audit %>% transmute(audit = "endpoint_registry_wording_audit", item = pattern, status = audit_status),
  condition_factor_wording_audit = condition_factor_wording_audit %>% transmute(audit = "condition_factor_wording_audit", item = pattern, status = audit_status),
  figure_table_audit = figure_table_audit %>% transmute(audit = "figure_table_audit", item = check, status),
  supplement_readiness = supplement_items %>% transmute(audit = "supplement_readiness", item = supplement_item, status = readiness_status)
)

combined_audit_status <- bind_rows(all_audits)

blocking_statuses <- c("fail", "flag", "missing_required")
review_statuses <- c("review", "missing", "missing_optional")

final_submission_checklist <- tibble(
  checklist_item = c(
    "All required manuscript files present",
    "No claim-traceability blocking flags",
    "PCA wording compliant",
    "Battery-comparison wording compliant",
    "Endpoint-registry wording compliant",
    "Condition-factor wording/source acceptable",
    "Figure/table files and captions auditable",
    "Supplement materials ready or explicitly documented",
    "Reproducibility audit available",
    "Ready for external scientific editing"
  ),
  status = c(
    ifelse(any(submission_file_audit$audit_status %in% blocking_statuses), "block", "pass"),
    ifelse(any(claim_traceability_audit$status %in% blocking_statuses), "block", ifelse(any(claim_traceability_audit$status %in% review_statuses), "review", "pass")),
    ifelse(any(pca_wording_audit$audit_status %in% blocking_statuses), "block", ifelse(any(pca_wording_audit$audit_status %in% review_statuses), "review", "pass")),
    ifelse(any(battery_wording_audit$audit_status %in% blocking_statuses), "block", ifelse(any(battery_wording_audit$audit_status %in% review_statuses), "review", "pass")),
    ifelse(any(endpoint_registry_wording_audit$audit_status %in% blocking_statuses), "block", ifelse(any(endpoint_registry_wording_audit$audit_status %in% review_statuses), "review", "pass")),
    ifelse(any(condition_factor_wording_audit$audit_status %in% blocking_statuses), "block", ifelse(any(condition_factor_wording_audit$audit_status %in% review_statuses), "review", "pass")),
    ifelse(any(figure_table_audit$status %in% blocking_statuses), "block", ifelse(any(figure_table_audit$status %in% review_statuses), "review", "pass")),
    ifelse(any(supplement_items$readiness_status == "missing" & supplement_items$supplement_item != "Endpoint effect estimates or contrasts"), "review", "pass"),
    ifelse(!is.null(repro_status) || !is.null(repro_manifest), "pass", "review"),
    "review"
  ),
  notes = c(
    "Checked manuscript text, captions, traceability table, and manifest.",
    "Blocking wording includes unsupported causal claims, PCA-as-proof language, and untraceable claims.",
    "PCA must remain a single full-battery visualization interpreted from loadings.",
    "Battery comparisons must avoid 'no difference' unless formal equivalence supports it.",
    "Endpoint registry must replace primary/secondary hierarchy.",
    "Condition factor must be tied to audit source if discussed.",
    "Figure/table numbering, captions, and generated files checked.",
    "Supplementary files should include core tables, PCA support, robustness, and reproducibility outputs.",
    "Notebook 09 outputs checked.",
    "External editing can proceed only after block items are cleared; review items require manual decision."
  )
)

path_final_submission_checklist <- write_submission_table(final_submission_checklist, "11_final_submission_checklist.csv")
path_combined_audit_status <- write_submission_table(combined_audit_status, "11_combined_audit_status.csv")

final_submission_checklist
```

# 11.12 Submission audit summary

```{r submission-audit-summary}
n_blocks <- sum(final_submission_checklist$status == "block", na.rm = TRUE)
n_reviews <- sum(final_submission_checklist$status == "review", na.rm = TRUE)
n_pass <- sum(final_submission_checklist$status == "pass", na.rm = TRUE)

blocking_items <- final_submission_checklist %>%
  filter(status == "block") %>%
  mutate(line = paste0("- ", checklist_item, ": ", notes)) %>%
  pull(line)

review_items <- final_submission_checklist %>%
  filter(status == "review") %>%
  mutate(line = paste0("- ", checklist_item, ": ", notes)) %>%
  pull(line)

submission_audit_summary <- c(
  "# Submission package audit summary",
  "",
  paste0("Generated on: ", Sys.Date()),
  "",
  "## Final status counts",
  "",
  paste0("- Pass: ", n_pass),
  paste0("- Review: ", n_reviews),
  paste0("- Block: ", n_blocks),
  "",
  "## Blocking items",
  "",
  if (length(blocking_items) == 0) "- None detected by automated audit." else blocking_items,
  "",
  "## Review items",
  "",
  if (length(review_items) == 0) "- None detected by automated audit." else review_items,
  "",
  "## Required next action",
  "",
  if (n_blocks > 0) {
    "Clear blocking items before external submission or journal-facing circulation."
  } else if (n_reviews > 0) {
    "Review flagged items manually before external submission or journal-facing circulation."
  } else {
    "No blocking or review items were detected by the automated audit. Manual scientific and editorial review is still required."
  },
  "",
  "## Non-negotiable wording rules retained",
  "",
  "- Do not use manuscript-facing 'thesis-style' language.",
  "- Do not use primary/secondary endpoint hierarchy language.",
  "- Do not write that PCA proves ethanol effects.",
  "- Do not force PCA component labels before loadings.",
  "- Do not write 'no difference' for non-significant battery comparisons unless equivalence was formally supported.",
  "- Do not add a behavioral-only PCA.",
  "- Do not include condition factor in PCA unless the condition-factor audit supports it.",
  "- Every manuscript claim must map to a notebook output, table, figure, or decision log."
)

path_submission_audit_summary <- write_submission_text(submission_audit_summary, "11_submission_audit_summary.md")

cat(paste(submission_audit_summary, collapse = "\n"))
```

# 11.13 Submission package manifest

```{r submission-package-manifest}
submission_package_manifest <- tibble(
  output_name = c(
    "11_submission_file_audit.csv",
    "11_claim_traceability_audit.csv",
    "11_pca_wording_audit.csv",
    "11_battery_wording_audit.csv",
    "11_endpoint_registry_wording_audit.csv",
    "11_condition_factor_wording_audit.csv",
    "11_figure_table_audit.csv",
    "11_supplement_readiness_checklist.csv",
    "11_final_submission_checklist.csv",
    "11_combined_audit_status.csv",
    "11_submission_audit_summary.md"
  ),
  output_type = c(
    "csv",
    "csv",
    "csv",
    "csv",
    "csv",
    "csv",
    "csv",
    "csv",
    "csv",
    "csv",
    "markdown"
  ),
  path = c(
    path_submission_file_audit,
    path_claim_traceability_audit,
    path_pca_wording_audit,
    path_battery_wording_audit,
    path_endpoint_registry_wording_audit,
    path_condition_factor_wording_audit,
    path_figure_table_audit,
    path_supplement_readiness,
    path_final_submission_checklist,
    path_combined_audit_status,
    path_submission_audit_summary
  ),
  status = ifelse(file.exists(path), "generated", "missing"),
  notes = c(
    "Presence and checksum audit for manuscript package files.",
    "Traceability and overclaiming audit.",
    "PCA wording compliance audit.",
    "Battery-comparison wording audit.",
    "Endpoint-registry wording audit.",
    "Condition-factor wording and source audit.",
    "Figure/table caption and file audit.",
    "Supplementary material readiness checklist.",
    "Final submission gate checklist.",
    "Combined status table across all submission-package audits.",
    "Human-readable submission audit summary."
  )
)

path_submission_manifest <- write_submission_table(submission_package_manifest, "11_submission_package_manifest.csv")

submission_package_manifest
```

# 11.14 Completion statement

Notebook 11 is complete when the submission-package folder contains the final audit tables, wording audits, supplement checklist, final submission checklist, package manifest, and audit summary.

If any item is marked `block`, patch the relevant manuscript text or upstream notebook output before circulation. If an item is marked `review`, decide manually whether the wording or missing file is acceptable for the target journal. If all items pass or are manually cleared, the notebook pipeline is complete.
